### 2017-05-11 paper.tex
Initial commit

% Preamble {{{1
\documentclass[a4paper]{scrartcl}

\usepackage[utf8]{inputenc} % Assume this file is encoded in UTF-8.
\usepackage[T1]{fontenc}    % Don't fake umlauts etc.
\usepackage{lmodern}        % Use the lmodern font (http://tex.stackexchange.com/a/65103).
\usepackage{microtype}      % Better microtypography (http://www.ctan.org/pkg/microtype)
\usepackage{hyperref}       % Clickable hyperlinks, load before `glossaries`
\usepackage{cleveref}       % Use `\cref{fig:foo}` instead of `figure~\ref{fig:foo}`.
                            % Load after `hyperref`.  See
                            % <https://tex.stackexchange.com/q/36295>
\usepackage{comment}        % Comment out sections of text.
\usepackage{mathtools}      % Improved facilities for typesetting mathematical formulae
\usepackage{dirtytalk}      % ...

% Source code listings with improved syntax highlighting
\usepackage{minted}
\usemintedstyle{pastie}

\usepackage{color}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\setminted{bgcolor=bg}

\usepackage{biblatex}
\addbibresource{paper.bib}

\usepackage[xindy, toc, acronym, nomain]{glossaries} % Load after `hyperref`.

\makeglossaries

\usepackage{tikz}
\usetikzlibrary{datavisualization}
% \usetikzlibrary{datavisualization.formats.functions}

% See <https://tex.stackexchange.com/a/155317>, <https://tex.stackexchange.com/a/320521>
% and <https://tex.stackexchange.com/a/75507>.
% \usepackage{tikzscale}

\newacronym{emm}{EMM}{external memory model}
\newacronym{hdd}{HDD}{hard disk drive}
\newacronym{ram}{RAM}{random access machine}
\newacronym{smt}{SMT}{simultaneous multithreading}
\newacronym{tlb}{TLB}{translation lookaside buffer}
\newacronym{ssd}{SSD}{solid-state drive}

\newcommand{\article}{article} % report, paper?

% Top matter {{{1
\title{Algorithms for Hardware Caches}
\author{Lukas Waymann}

% Body {{{1
\begin{document}
\maketitle
\newpage

\begin{abstract}
   % What?
   Typical present-day CPUs have two or more levels of caches.  This \article{} presents
   the basic techniques used to optimize program performance based on knowledge about how
   these hardware caches function.

   The abstract \gls{emm} for memory hierarchies is explained and a small selection of
   algorithms developed for it
   % analyzed under it
   are explored mathematically and empirically.
\end{abstract}
\newpage

\tableofcontents
\newpage

\glsresetall % Reset the use status of all acronyms.

\section{Introduction}
% What is this paper about?  Why learn about them?  How much performance is at stake?

% What are hardware caches?
A hardware cache is a comparatively
% relatively
fast and small physical memory.  It stores a subset of the data present in slower, larger
memory that is expected to be used again soon.  The purpose of this additional memory is
to reduce the number of accesses to the underlying slower storage.

% Why caches?
There are fundamental reasons that having one single, uniform
% homogeneous
type of memory is not viable.  No signal can propagate faster than the speed of light.
Thus, every storage technology can only reach a finite amount of data within a desired
access latency \cite[p.~2]{afmh}.

% TODO: move this paragraph out of the introduction.
% Hardware caches are managed by hardware directly.  They are generally opaque to the
% operating system and other programs.  That is, software has no direct control over the
% contents of a hardware cache.

The most ubiquitous example for hardware caches is the hierarchy of
% are the various levels (most commonly 2 or 3) of
CPU caches that are found on almost all present-day CPUs.  They are designated L1 cache,
L2 cache, and so on, with L1 being the fastest and smallest level.  The underlying storage
for CPU caches is the main memory.

There are more storage levels that comprise
% constitute
the \emph{memory hierarchy} of a computer along with CPU caches and main memory.  For
example \glspl{hdd} and \glspl{ssd}. % Also: registers, internal buffers of HDDs and SSDs,
                                     % (tapes?), ...
% Focus on CPU caches.  Why?
However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
main memory, eliminating the need for swapping to disk under many workloads.

I will focus on how to use CPU caches effectively and the resulting
% enabled
performance gains in this \article{}.
% TODO: what about TLB?

\section{Motivation} % So what?
%    Three things Really Matter for performance. The first one is Algorithm, the second
%    one is your code being Non-Blocking, and the third one is Data Locality."
%       -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/

% We will see that
Two algorithms solving the same problem with the same asymptotic complexity (in the same
\(\Theta(g(n))\)) may differ in performance by two orders of magnitude~\cite[p.~2]{afmh}.

% TODO: how and how much can we speed up our programs by considering their interactions
% with CPU caches?

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array to
% compute the sum of its elements.  Before exiting, it prints the CPU time spend summing the
% array.

\begin{listing}
   \inputminted[firstline=3]{c}{array-sum/array-sum.c}
   \caption{This is C code}
   \label{lst:array-sum}
\end{listing}

% This program is likely to have excellent spatial locality: the array...

% \begin{figure}[htb]
\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={label={Array size (KiB)}, length=0.8\textwidth},
       y axis={label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=array-sum/size-time.csv, separator=\space];
   \end{tikzpicture}
   \caption{This figure took way too long to create}
   \label{fig:array-sum}
\end{figure}

See \cref{fig:array-sum}.  What's going on here?

% \section{Basic Principles}
% \section{Basic Concepts}
\section{Key Terms}
\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

Moving down the memory hierarchy, access latencies increase faster than the
\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.

\subsection{Temporal Locality}

\subsection{Cache Hit}

\subsection{Cache Miss}
% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?

\subsection{External Memory Model}
The \gls{emm} is a widely used
% TODO: is it?)
extension of the \gls{ram} model.

\clearpage

\printglossary[type=\acronymtype] % Print the list of acronyms.

\printbibliography[heading=bibintoc]

\end{document}

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-11 array-sum/array-sum.c

@@ -1,0 +1,31 @@
+/* vim: set tw=72 sts=-1 sw=3 et: */
+
+#include "stdint.h"
+#include "stdio.h"
+#include "stdlib.h"
+#include "time.h"
+
+#define N 5000
+
+int64_t array[SIZE];
+
+int main() {
+   srand(time(0));
+   for (size_t i = 0; i < SIZE; ++i) {
+      array[i] = rand();
+   }
+
+   clock_t t0 = clock();
+   int64_t sum = 0;
+   for (size_t n = 0; n < N; ++n) {
+      for (size_t i = 0; i < SIZE; ++i) {
+         sum += array[i];
+      }
+   }
+
+   clock_t t1 = clock();
+   printf("%d %f\n", SIZE * 8 / 1024,
+          1000. * (t1 - t0) / CLOCKS_PER_SEC);
+   // Print the sum so the loop doesn't get optimized out!
+   printf("%d\n", sum);
+}

### 2017-05-11 paper.tex
Expand content, tweak things

@@
 
+% Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
+\usepackage[style=iso]{datetime2}
+
 % Source code listings with improved syntax highlighting
@@
 
-\usepackage{biblatex}
+% Keep using ISO 8601 consistently, like an even more reasonable person.  See
+% <https://tex.stackexchange.com/q/231208>.
+\usepackage[date=edtf, urldate=edtf]{biblatex}
 \addbibresource{paper.bib}
@@
 
-% TODO: move this paragraph out of the introduction.
-% Hardware caches are managed by hardware directly.  They are generally opaque to the
-% operating system and other programs.  That is, software has no direct control over the
-% contents of a hardware cache.
-
 The most ubiquitous example for hardware caches is the hierarchy of
@@
 
+% TODO: move this paragraph?
+Hardware caches are managed by hardware directly.  They are generally opaque to the
+operating system and other programs.  That is, software has no direct control over the
+contents of a hardware cache.
+
+% So what?  Why learn about hardware caches?  How much performance do I gain/lose
+% depending on how *cache-friendly* my algorithm/code is?
+
 % We will see that
-Two algorithms solving the same problem with the same asymptotic complexity (in the same
-\(\Theta(g(n))\)) may differ in performance by two orders of magnitude~\cite[p.~2]{afmh}.
+Despite this, two algorithms solving the same problem with the same asymptotic complexity
+(in the same \(\Theta(g(n))\)) may differ in performance by two orders of magnitude
+because of different \emph{memory access patterns}~\cite{bigos}.  We will see an example
+of this in section TODO.
 
-% TODO: how and how much can we speed up our programs by considering their interactions
-% with CPU caches?
+In a nutshell, hardware caches are ubiquitous but the performance improvements they
+provide are conditional.
+\begin{comment}
+   To use them effectively,
+   % To obtain optimal performance,
+   algorithms must be designed and implemented with the architecture
+   % design, structure, manner of functioning, inner workings, properties
+   of hardware caches in mind.
+\end{comment}
+Effective use of hardware caches requires knowledge about how they work.
+% their architecture.
+Algorithms must be designed and implemented observing this knowledge.
+% their interactions with hardware caches.
+
+% TODO: and *how* do I make effective use of hardware caches?
+
+% \section{Basic Principles}
+\section{Basic Concepts}
+% \section{Key Terms}
+% That sounds boring.
+
+\subsection{Spatial Locality}
+Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
+memory into the CPU cache) in \emph{blocks}.
+
+Moving down the memory hierarchy, access latencies increase faster than the
+\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
+many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
+memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.
 
 % TODO: move this.
-% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array to
-% compute the sum of its elements.  Before exiting, it prints the CPU time spend summing the
-% array.
+% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
+% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
+% summing the array.
 
@@
 
-See \cref{fig:array-sum}.  What's going on here?
-
-% \section{Basic Principles}
-% \section{Basic Concepts}
-\section{Key Terms}
-\subsection{Spatial Locality}
-Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
-memory into the CPU cache) in \emph{blocks}.
-
-Moving down the memory hierarchy, access latencies increase faster than the
-\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
-many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
-memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.
+% See \cref{fig:array-sum}.  What's going on here?
 
@@
 % TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
+
+\subsection{Cache Line} % or Cache Block
+% On x86/x64, cache line is 64 bytes for many years now.
+%    -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
 

### 2017-05-20 paper.tex
Expand content and tweak stuff

@@
 
+% See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
+\usepackage{enumitem}
+\setlist{noitemsep}
+
+% TODO.  See <https://tex.stackexchange.com/q/119513>.
+\crefname{appsec}{Appendix}{Appendices}
+
 % Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
 \usepackage[style=iso]{datetime2}
+
+% TODO: will this just force LaTeX to make worse choices?
+% \usepackage[all]{nowidow}
 
@@
 
-\usepackage{color}
+% Define a background color for `minted` listings.  See <http://ctan.org/pkg/minted> and
+% <https://tex.stackexchange.com/q/150369>.
+\usepackage{xcolor}
 \definecolor{bg}{rgb}{0.95,0.95,0.95}
 \setminted{bgcolor=bg}
+% Don't shade the background when using `mintinline`.
+\setmintedinline{bgcolor={}}
+
+% ...
+\usepackage[titletoc]{appendix}
 
@@
 % <https://tex.stackexchange.com/q/231208>.
-\usepackage[date=edtf, urldate=edtf]{biblatex}
+\usepackage[date=edtf, urldate=edtf, seconds=true]{biblatex}
 \addbibresource{paper.bib}
 
-\usepackage[xindy, toc, acronym, nomain]{glossaries} % Load after `hyperref`.
+\usepackage[xindy, toc, acronym]{glossaries} % Load after `hyperref`.
 
@@
 
-\newcommand{\article}{article} % report, paper?
+\newglossaryentry{l1d}
+{
+   name=\texttt{L1d},
+   description={},
+}
+
+\newglossaryentry{l1i}
+{
+   name=\texttt{L1i},
+   description={},
+}
+
+\newglossaryentry{l2}
+{
+   name=\texttt{L2},
+   description={},
+}
+
+\newglossaryentry{l3}
+{
+   name=\texttt{L3},
+   description={},
+}
+
+\newcommand*{\article}{article} % report, paper?
+
+% Define a command that takes exactly 2 arguments, the first one defaulting to `1`.  The
+% second argument should be a list of items separated by `,`.  The list item at the
+% position specified by the first argument in printed.  See
+% <https://tex.stackexchange.com/a/99271>,
+% <http://mirrors.ctan.org/macros/generic/listofitems/listofitems-en.pdf>, and
+% <https://tex.stackexchange.com/q/276697>.
+\usepackage{listofitems}
+\newcommand*{\alts}[2][1]{%
+   \setsepchar{,}%
+   \readlist*\arg{#2}%
+   \arg[#1]%
+}
 
 % Top matter {{{1
-\title{Algorithms for Hardware Caches}
+\title{Basics of Hardware Cache Optimization}
+% \title{Algorithms for Hardware Caches}
+% \title{Basics of CPU Cache Optimization}
+% \title{Basics of Optimizing for Hardware Caches}
+% \title{Fundamentals of Optimizing for Hardware Caches}
+% \title{Introduction to Optimizations for Hardware Caches}
+% \title{Optimizing for Hardware Caches}
+% \subtitle{The Basics}
 \author{Lukas Waymann}
@@
 \maketitle
-\newpage
+
+% \newpage
 
@@
    The abstract \gls{emm} for memory hierarchies is explained and a small selection of
-   algorithms developed for it
-   % analyzed under it
-   are explored mathematically and empirically.
+   algorithms \alts{developed for it, analyzed under it} are explored mathematically and
+   empirically.
 \end{abstract}
@@
 
-% What are hardware caches?
-A hardware cache is a comparatively
-% relatively
-fast and small physical memory.  It stores a subset of the data present in slower, larger
-memory that is expected to be used again soon.  The purpose of this additional memory is
-to reduce the number of accesses to the underlying slower storage.
+% What are hardware caches?  Why caches?
+A hardware cache is a \alts{comparatively, relatively} fast and small physical memory.  It
+stores a subset of the data present in slower, larger memory that is expected to be used
+again soon.  The purpose of this additional memory is to reduce the number of accesses to
+the underlying slower storage.
 
-% Why caches?
-There are fundamental reasons that having one single, uniform
-% homogeneous
-type of memory is not viable.  No signal can propagate faster than the speed of light.
-Thus, every storage technology can only reach a finite amount of data within a desired
-access latency \cite[p.~2]{afmh}.
+% Hardware caches aren't going away.
+There are fundamental reasons that having one single, \alts{uniform, homogeneous} type of
+memory is not viable.  No signal can propagate faster than the speed of light.  Thus,
+every storage technology can only reach a finite amount of data within a desired access
+latency~\cite[2]{afmh}.
 
-The most ubiquitous example for hardware caches is the hierarchy of
-% are the various levels (most commonly 2 or 3) of
-CPU caches that are found on almost all present-day CPUs.  They are designated L1 cache,
-L2 cache, and so on, with L1 being the fastest and smallest level.  The underlying storage
-for CPU caches is the main memory.
+The most ubiquitous example for hardware caches \alts{is the hierarchy, are the various
+levels (most commonly 2 or 3)} of CPU caches that are found on almost all present-day
+CPUs.  They are designated L1 cache, L2 cache, and so on, with L1 being the fastest and
+smallest level.  The underlying storage for CPU caches is the main memory.
 
-There are more storage levels that comprise
-% constitute
-the \emph{memory hierarchy} of a computer along with CPU caches and main memory.  For
-example \glspl{hdd} and \glspl{ssd}. % Also: registers, internal buffers of HDDs and SSDs,
-                                     % (tapes?), ...
+There are more storage levels that \alts{comprise, constitute} the \emph{memory hierarchy}
+of a computer along with CPU caches and main memory.  For example \glspl{hdd} and
+\glspl{ssd}.
+% Also: registers, internal buffers of HDDs and SSDs, (tapes), ...
 % Focus on CPU caches.  Why?
@@
 
-I will focus on how to use CPU caches effectively and the resulting
-% enabled
+I will focus on how to use CPU caches effectively and the \alts{enabled, resulting}
 performance gains in this \article{}.
+
 % TODO: what about TLB?
@@
 \section{Motivation} % So what?
-%    Three things Really Matter for performance.  The first one is Algorithm, the second
-%    one is your code being Non-Blocking, and the third one is Data Locality."
+% > Three things Really Matter for performance.  The first one is Algorithm, the second
+% > one is your code being Non-Blocking, and the third one is Data Locality."
 %      -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/
 
-% TODO: move this paragraph?
 Hardware caches are managed by hardware directly.  They are generally opaque to the
@@
 % So what?  Why learn about hardware caches?  How much performance do I gain/lose
-% depending on how *cache-friendly* my algorithm/code is?
-
-% We will see that
-Despite this, two algorithms solving the same problem with the same asymptotic complexity
-(in the same \(\Theta(g(n))\)) may differ in performance by two orders of magnitude
-because of different \emph{memory access patterns}~\cite{bigos}.  We will see an example
-of this in section TODO.
+% depending on how cache-friendly my algorithm/code is?
+\alts{Despite this, We will see that despite this}, two algorithms solving the same
+problem with the same asymptotic complexity (in the same \(\Theta(g(n))\)) may differ in
+performance by two orders of magnitude because of different \emph{memory access
+patterns} (\cref{sec:map})~\cite{bigos}.  We will see an example of this in
+\cref{sec:vvl}.
 
@@
 \end{comment}
-Effective use of hardware caches requires knowledge about how they work.
-% their architecture.
-Algorithms must be designed and implemented observing this knowledge.
-% their interactions with hardware caches.
+Effective use of hardware caches requires knowledge about \alts{how they work, their
+architecture}.  Algorithms must be designed and implemented observing \alts{this
+knowledge, their interactions with hardware caches}.
 
-% TODO: and *how* do I make effective use of hardware caches?
+% How?
 
+\section{Types of CPU Caches}
+% > Most modern desktop and server CPUs have at least three independent caches: an
+% > instruction cache to speed up executable instruction fetch, a data cache to speed up
+% > data fetch and store, and a translation lookaside buffer (TLB) used to speed up
+% > virtual-to-physical address translation for both executable instructions and data.
+%      -- https://en.wikipedia.org/wiki/CPU_cache#Overview
+% > There are three common types of CPU caches: ...
+%      -- Scott Meyers (talk at code::dive)
+Current x86 CPUs \alts{generally, typically, commonly} have three main types of caches:
+data caches, instruction caches, and \glspl{tlb}%
+~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m07s}{11:07}]{scott-meyers-talk}.
+Some caches are used for data as well as instructions and are called \emph{unified}.%
+~\cite[20]{drepper2007}.
+\alts{{A processor may have multiple caches of each type, which}, {Multiple caches of each
+type may be present, and}} are organised into numerical \emph{levels}
+\alts{{starting at 1, the smallest and fastest level,},}
+based on their size and speed.
+% Each added level is bigger and slower than its predecessor.
+% The smallest and fastest is level 1.
+
+% TODO?  The reason to have multiple levels...
+
+% > Often there are separate L1 caches for instructions and data
+%      -- Algorithms for Memory Hierarchies, page 3
+% > Systems nowaeays have at-least two levels of cache
+%      -- Algorithms for Memory Hierarchies, page 172
+% > [T]he caches from L2 on are unified caches which contain both code and data
+%      -- Drepper, p. 31
+% > Later Intel models have shared L2 caches for dual-core processors.  For quad-core
+% > processors we have to deal with separate L2 caches for each pair of two cores.
+%      -- Drepper, p. 35
+
+% Terminology / Nomenclature.
+In practice, a \alts{currently, presently} representative%
+\footnote{%
+   % https://en.wikipedia.org/wiki/Bobcat_(microarchitecture)
+   E.g. for AMD Family 14h processors~\cite[30--32]{14h},
+   % https://en.wikipedia.org/wiki/List_of_AMD_CPU_microarchitectures
+   % https://en.wikipedia.org/wiki/Zen_(microarchitecture)
+   % 32 KiB L1d, 64 KiB L1i, 512 KiB L2, 8 to 16 MiB L3
+   AMD Zen (17h)~\cite{zen}, and
+   % https://en.wikipedia.org/wiki/Kaby_Lake
+   % https://en.wikipedia.org/wiki/Skylake_(microarchitecture)
+   % 32 KiB L1d, 32 KiB L1i, 256 KiB L2, 2 to 8 MiB L3
+   Intel Skylake desktop processors%
+   ~\cite[figure 2-1, table 2-4]{skylake}
+   % ~\cite[figure 2-1, \pno~2-2, table 2-4, \pno~2-6]{skylake}.
+   % ~\cite[{2-2}, {2-6}]{skylake}.
+   % <https://en.wikipedia.org/wiki/Bulldozer_(microarchitecture)> is too weird.
+}
+x86 cache hierarchy consists of:
+\begin{itemize}
+   % https://en.wikipedia.org/wiki/Cache_hierarchy#Shared_versus_private
+   \item Separate level 1 data and instruction caches of 32 to 64 KiB for each core
+      (denoted \gls{l1d} and \gls{l1i} by  \textcite[14--15]{drepper2007}).
+      % TODO?  Why have a separate instruction cache?
+      Machine instructions in \gls{l1i} are already decoded%
+      ~\cite[31, 56]{drepper2007}.
+      % ~\cite[14, 31, 56]{drepper2007}.
+   % \item A level 2 cache for \say{both code and data}~\cite[31]{drepper2007}.
+   \item A unified \gls{l2} cache of 256 to 512 KiB for each core.
+   \item Often a unified \gls{l3} cache of 2 to 16 MiB shared between all cores.
+   \item TODO: Some \glspl{tlb} I guess.
+\end{itemize}
+
+% \subsection{Access Times}
+% http://ithare.com/infographics-operation-costs-in-cpu-clock-cycles/
+% http://www.getitwriteonline.com/archive/040201hyphadj.htm
+\alts{Estimates, Order-of-magnitude estimates} of typical access latencies \alts[2]{are as
+follows, are given by \textcite{ithare-cycles}.}%
+\footnote{%
+   Intel~\cite[table 2-4]{skylake},
+   \textcites
+   % {ithare-paadl}{ithare-wisdoms}
+   [\href{https://youtu.be/WDIkqP4JbkE?t=17m52s}{17:52}, slide 18]{scott-meyers-talk}
+   [2--3, 171]{afmh}[16, 20--21]{drepper2007} all give comparable numbers for various
+   architectures.
+   % [\ppno~16, 20--21, fig. 3.10]{drepper2007}
+}
+
+\begin{center}
+   \begin{tabular}{ r | c c c c }
+             & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
+      Cycles & 3--4      & 10--12   & 30--70   & 100--150
+   \end{tabular}
+\end{center}
+%
+% These are taken from~\textcite{ithare-cycles} but comparable numbers are given by
+
+% > [Instruction] cache is much less problematic than the data cache.
+%      -- Drepper, p. 31
+The biggest target for optimizations is the data cache.  \say{[Instruction] cache is much
+less problematic}~\cite[31]{drepper2007} and optimizations for data and instruction cache
+tend to improve \gls{tlb} usage as well%
+~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m53s}{11:53}]{scott-meyers-talk}.
+
+\section{Basic Concepts}
 % \section{Basic Principles}
-\section{Basic Concepts}
-% \section{Key Terms}
-% That sounds boring.
+% \section{Key Terms} % That sounds boring.
+Some architectural properties of hardware caches lead to important concepts for using them
+effectively.
+
+\subsection{Cache Line} % or Cache Block
+% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
+%
+% > On x86/x64, cache line is 64 bytes for many years now.
+%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
+% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
+%      -- Drepper, p. 15
+% > It is not possible for a cache to hold partial cache lines.
+%      -- Drepper, p. 16
+\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
+memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
+x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
+{scott-meyers-talk}.%
+\footnote{%
+% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
+% > cache 256 KB in size, with 128-byte cache blocks.
+%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
+% TODO: what about the block size of main memory?  Should it be the same?
+Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
+The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
+line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
+line}~\cite[p.~11]{pentium4}.}
+\begin{comment}
+   \multiplefootnoteseparator%
+   % See <https://tex.stackexchange.com/a/71015>.
+   \footnote{This can also be checked on the command line:
+   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
+\end{comment}
+% > It means that as soon as you've accessed any single byte in a cache line, all the
+% > other 63 bytes are already in L1 cache
+%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
+\alts{%
+   This means accessing a single uncached 32-bit integer entails loading another 60
+   adjacent bytes.,
+   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
+   well.},
+   {When, for example, accessing a single byte that isn't already cached, another 63
+   adjacent bytes will be loaded.}
+}
+% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.
+
+My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
+lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
+% \begin{minted}[gobble=3]{bash}
+%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
+%    64
+%    64
+% \end{minted}
+%stopzone
+% \footnote{%
+%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
+%    %stopzone
+%    \noindent\mintinline{bash}!64!\\
+%    \noindent\mintinline{bash}!64!
+% }
+We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
+with an increment given at compile time as \texttt{STEP} and measures the processor time.
+The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
+% Starting from a step size of 16, the time roughly halves every time the step size is
+% doubled.  For the first 4 step sizes however, it is almost constant.
+As expected, the time roughly halves whenever the step size is doubled --- but only from a
+step size of 16.  For the first 4 step sizes, it is almost constant.
+
+% The reason why the loops take the same amount of time has to do with memory.  The
+% running time of these loops is dominated by the memory accesses to the array, not by the
+% integer multiplications.
+%    -- https://igoro.com/archive/gallery-of-processor-cache-effects/
+This is because the run times are \alts{primarily due to, dominated by} memory accesses.
+Up to a step size of 8, every 64-byte line has to be loaded.  At 16, the values we modify
+are 128 bytes apart,%
+\footnote{16 \texttt{int64\_t} values of 8 bytes each}
+so every other cache line is skipped.  At 32, three out of four cache lines are skipped,
+and so on~\cite[cf.][example 2]{gallery}.
+
+\begin{listing}
+   \inputminted[firstline=27]{c}{line-size/line-size.c}
+   \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
+   \label{lst:line-size}
+\end{listing}
+
+\begin{figure}
+   \centering
+   \begin{tikzpicture}
+      \datavisualization
+      [scientific axes=clean,
+       x axis={logarithmic,
+               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
+                                 8 as \textbf{8}, 16, 64, 256, 1024}},
+                      minor={at={32, 128, 512}}},
+               label={Step size}, length=0.8\textwidth},
+       y axis={logarithmic,
+               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
+               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
+               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
+               label={Processor time (ms)}, length=6cm},
+       visualize as scatter,
+       scatter={style={mark=*, mark options={scale=.65}}}]
+         % See <https://tex.stackexchange.com/q/198323>.
+         data [read from file=line-size/line-size.csv, separator=\space];
+   \end{tikzpicture}
+   \caption{Processor times for running \cref{lst:line-size}}
+   \label{fig:line-size}
+\end{figure}
+
+\subsection{Prefetching}
+% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
+% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
+% https://en.wikipedia.org/wiki/Cache_prefetching
+\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
+patterns (for instance sequential access to an array)... TODO
+
+\subsection{Associativity}
+% https://en.wikipedia.org/wiki/CPU_cache#Associativity
+
+\subsection{Cache Hits and Misses}
+% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
 
@@
 many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
-memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.
+memory is almost as fast as access to a single bit},~\cite[2]{afmh}.
 
@@
 
+\subsection{Temporal Locality}
+TODO~\cite{drepper2007}.
+
+\subsection{Memory Access Pattern}
+\label{sec:map}
+
+\section{Example: \texttt{std::vector} vs. \texttt{std::list}}
+\label{sec:vvl}
+
+% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
+% https://en.wikipedia.org/wiki/Cache_replacement_policies
+
+\section{External Memory Model}
+The \gls{emm} is a widely used
+% TODO: is it?
+extension of the \gls{ram} model.
+
 \begin{listing}
-   \inputminted[firstline=3]{c}{array-sum/array-sum.c}
+   \inputminted[firstline=8]{c}{array-sum/array-sum.c}
    \caption{This is C code}
@@
 
-\subsection{Temporal Locality}
+\section{Cache-Oblivious Algorithms} % Cache-Oblivious Model
 
-\subsection{Cache Hit}
+\section{Parallel Computing}
 
-\subsection{Cache Miss}
-% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
+\subsection{Cache Coherence} % Coherence or coherency?
+% https://en.wikipedia.org/wiki/Cache_coherence
+% Dirty.  Invalid.
 
-\subsection{Cache Line} % or Cache Block
-% On x86/x64, cache line is 64 bytes for many years now.
-%    -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
+\subsection{False Sharing}
 
-\subsection{External Memory Model}
-The \gls{emm} is a widely used
-% TODO: is it?)
-extension of the \gls{ram} model.
+\clearpage
+% \appendix
+\begin{appendices}
+% See <https://tex.stackexchange.com/q/119513>.
+\crefalias{section}{appsec}
+
+\section{Reading Information About the CPU}
+\label{app:cpuinfo}
+% https://stackoverflow.com/q/7281699
+% https://unix.stackexchange.com/q/167038
+% https://superuser.com/q/55776
+There are many ways to display information about the processor(s) the operating system is
+running on.  Among others, the \mintinline{text}{lscpu(1)} and
+\mintinline{text}{getconf(1)} programs and the \mintinline{text}{/proc/cpuinfo}
+pseudo-file on Linux.  This is \alts{how I checked, what I used to check} my CPU's cache
+line sizes, for example:
+\begin{minted}[gobble=3]{bash}
+   $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
+   64
+   64
+\end{minted}
+%stopzone
+
+\section{TODO: Measuring Stuff}
+
+\begin{comment}
+\section{Disabling the Prefetcher}
+My initial results from from running \cref{lst:array-sum} were quite different.  See TODO.
+
+TODO: I installed \texttt{msr-tools} and OProfile.
+% This seems to work.  It did affect performance.
+\begin{minted}[gobble=3]{bash}
+   $ sudo modprobe msr
+   $ sudo rdmsr -a 0xC0011022
+   600001800000000
+   600001800000000
+   $ sudo wrmsr -a 0xC0011022 0x600001800002000
+   $ sudo rdmsr -a -f '13:13' 0xC0011022
+   1
+   1
+   $ sudo rdmsr -a 0xC0011022
+   600001800002000
+   600001800002000
+\end{minted}
+%stopzone
+I think resetting the MSR to enable hardware prefetches doesn't work.  To get the same
+benchmmark results as before tweaking it, I have to reboot.
+\end{comment}
+
+\end{appendices}
 

### 2017-05-20 array-sum/array-sum.c

@@ -1,9 +1,9 @@
 /* vim: set tw=72 sts=-1 sw=3 et: */
 
-#include "stdint.h"
-#include "stdio.h"
-#include "stdlib.h"
-#include "time.h"
+#include "stdint.h"  // int64_t
+#include "stdio.h"   // printf
+#include "stdlib.h"  // calloc
+#include "time.h"    // clock
 
 #define N 5000
 

### 2017-05-20 line-size/line-size-alt.c

@@ -1,0 +1,26 @@
+/* vim: set tw=72 sts=-1 sw=3 et cms=//%s: */
+
+#include "stdint.h"  // int64_t
+#include "stdio.h"   // printf
+#include "stdlib.h"  // calloc
+#include "time.h"    // clock
+
+#define SIZE 67108864  // 64 * 1024 * 1024.  The array will be 512 MiB.
+
+int64_t array[SIZE];  // Using `int8_t` or `int32_t`: bad.
+
+int main() {
+   // int64_t* array = (int64_t*)malloc(sizeof(int64_t) * SIZE);
+   // This just forces the memory to really be allocated before we
+   // measure.
+   for (size_t i = 0; i < SIZE; ++i) array[i] = 1;
+   struct timespec t0, t1;
+   clock_gettime(CLOCK_PROCESS_CPUTIME_ID, &t0);
+   for (size_t i = 0; i < SIZE; i += STEP) {
+      array[i] *= 3;
+   }
+   clock_gettime(CLOCK_PROCESS_CPUTIME_ID, &t1);
+   printf("%d %f\n", STEP,
+          1.e3 * (t1.tv_sec - t0.tv_sec) +
+              1.e-6 * (t1.tv_nsec - t0.tv_nsec));
+}

### 2017-05-20 line-size/line-size.c

@@ -1,0 +1,37 @@
+/* vim: set tw=72 sts=-1 sw=3 et cms=//%s: */
+
+#include "stdint.h"  // int64_t
+#include "stdio.h"   // printf
+#include "stdlib.h"  // calloc
+#include "time.h"    // clock
+
+// Adapted from <igoro.com/archive/gallery-of-processor-cache-effects/>.
+//
+// Apparently malloc(3) can return non-NULL WITHOUT HAVING ACTUALLY
+// ALLOCATED ANY MEMORY on Linux!  Even when requesting MORE MEMORY THAN
+// THE SUM OF THE SYSTEM'S PHYSICAL MEMORY AND SWAP SPACE!  ...use
+// `calloc`.
+//
+// Multiplying the array entries (`array[i] *= 3`) gives better results
+// than incrementing (`++array[i]`) them.  The processor times for runs
+// that need every cache line are more even.  Also, multiplying with
+// uneven numbers gives better results than with even ones.  For a step
+// size of 1 (and only for a step size of 1), the multiplication is
+// actually faster than something like `array[i] &= 1`.  This depends on
+// having optimization enabled.
+//
+// Generally, results are more predictable when the system is idle.
+// Rebooting also helped before when I got strange results (almost the
+// same times for step sizes of 32 and 64).
+
+#define SIZE 67108864  // 64 * 1024 * 1024.  The array will be 512 MiB.
+
+int main() {
+   int64_t* array = (int64_t*)calloc(SIZE, sizeof(int64_t));
+   clock_t t0 = clock();
+   for (size_t i = 0; i < SIZE; i += STEP) {
+      array[i] &= 1;  // Do something.  Anything.
+   }
+   clock_t t1 = clock();
+   printf("%d %f\n", STEP, 1000. * (t1 - t0) / CLOCKS_PER_SEC);
+}

### 2017-05-21 appendices.tex
Split content into multiple files and tweak stuff

--- 2017-05-11 paper.tex
@@
-% Preamble {{{1
-\documentclass[a4paper]{scrartcl}
+\section{Reading Information About the CPU}
+\label{app:cpuinfo}
+% https://stackoverflow.com/q/7281699
+% https://unix.stackexchange.com/q/167038
+% https://superuser.com/q/55776
+There are many ways to display information about the processor(s) the operating system is
+running on.  Among others, the \mintinline{text}{lscpu(1)} and
+\mintinline{text}{getconf(1)} programs and the \mintinline{text}{/proc/cpuinfo}
+pseudo-file on Linux.  This is \alts{how I checked, what I used to check} my CPU's cache
+line sizes, for example:
+\begin{minted}[gobble=3]{bash}
+   $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
+   64
+   64
+\end{minted}
+%stopzone
 
-\usepackage[utf8]{inputenc} % Assume this file is encoded in UTF-8.
-\usepackage[T1]{fontenc}    % Don't fake umlauts etc.
-\usepackage{lmodern}        % Use the lmodern font (http://tex.stackexchange.com/a/65103).
-\usepackage{microtype}      % Better microtypography (http://www.ctan.org/pkg/microtype)
-\usepackage{hyperref}       % Clickable hyperlinks, load before `glossaries`
-\usepackage{cleveref}       % Use `\cref{fig:foo}` instead of `figure~\ref{fig:foo}`.
-                            % Load after `hyperref`.  See
-                            % <https://tex.stackexchange.com/q/36295>
-\usepackage{comment}        % Comment out sections of text.
-\usepackage{mathtools}      % Improved facilities for typesetting mathematical formulae
-\usepackage{dirtytalk}      % ...
+% Performance Analysis
+% Run Time Analysis
+% Timing Methodology
+% Profiling Methodology
+% Profiling Pitfalls
+% Run Time Measuring
+% Benchmarking Methodology
+% Monitoring
+\section{Profiling Methodology}
+% https://en.wikipedia.org/wiki/Profiling_(computer_programming)
+% https://en.wikipedia.org/wiki/Processor_affinity
+% On `/usr/bin/time` for benchmarks: <http://stackoverflow.com/q/9006596>.
+% Andrei Alexandrescu - Writing Fast Code: <https://youtu.be/vrfYLlR8X8k>.
+Plenty of things can go wrong when measuring \alts{execution, run} times.
+\begin{itemize}
+   \item CPU frequency scaling and boosting
+      % Not a problem when measuring cycles.
+   \item Interrupts
+      % Taking the minimum helps.
+   \item CPU jumping
+      % See <https://youtu.be/vrfYLlR8X8k?t=33m34s>.
+   \item Context switches
+   \item Warmth of cache
+      % See <https://youtu.be/vrfYLlR8X8k?t=41m00s>.
+   % \item ...
+\end{itemize}
+Some effects will \alts{add noise to, pollute} the results or render them irreproducible,
+others will invalidate them completely.
 
-% Source code listings with improved syntax highlighting
-\usepackage{minted}
-\usemintedstyle{pastie}
+\begin{comment}
+\section{Disabling the Prefetcher}
+My initial results from from running \cref{lst:array-sum} were quite different.  See TODO.
 
-\usepackage{color}
-\definecolor{bg}{rgb}{0.95,0.95,0.95}
-\setminted{bgcolor=bg}
-
-\usepackage{biblatex}
-\addbibresource{paper.bib}
-
-\usepackage[xindy, toc, acronym, nomain]{glossaries} % Load after `hyperref`.
-
-\makeglossaries
-
-\usepackage{tikz}
-\usetikzlibrary{datavisualization}
-% \usetikzlibrary{datavisualization.formats.functions}
-
-% See <https://tex.stackexchange.com/a/155317>, <https://tex.stackexchange.com/a/320521>
-% and <https://tex.stackexchange.com/a/75507>.
-% \usepackage{tikzscale}
-
-\newacronym{emm}{EMM}{external memory model}
-\newacronym{hdd}{HDD}{hard disk drive}
-\newacronym{ram}{RAM}{random access machine}
-\newacronym{smt}{SMT}{simultaneous multithreading}
-\newacronym{tlb}{TLB}{translation lookaside buffer}
-\newacronym{ssd}{SSD}{solid-state drive}
-
-\newcommand{\article}{article} % report, paper?
-
-% Top matter {{{1
-\title{Algorithms for Hardware Caches}
-\author{Lukas Waymann}
-
-% Body {{{1
-\begin{document}
-\maketitle
-\newpage
-
-\begin{abstract}
-   % What?
-   Typical present-day CPUs have two or more levels of caches.  This \article{} presents
-   the basic techniques used to optimize program performance based on knowledge about how
-   these hardware caches function.
-
-   The abstract \gls{emm} for memory hierarchies is explained and a small selection of
-   algorithms developed for it
-   % analyzed under it
-   are explored mathematically and empirically.
-\end{abstract}
-\newpage
-
-\tableofcontents
-\newpage
-
-\glsresetall % Reset the use status of all acronyms.
-
-\section{Introduction}
-% What is this paper about?  Why learn about them?  How much performance is at stake?
-
-% What are hardware caches?
-A hardware cache is a comparatively
-% relatively
-fast and small physical memory.  It stores a subset of the data present in slower, larger
-memory that is expected to be used again soon.  The purpose of this additional memory is
-to reduce the number of accesses to the underlying slower storage.
-
-% Why caches?
-There are fundamental reasons that having one single, uniform
-% homogeneous
-type of memory is not viable.  No signal can propagate faster than the speed of light.
-Thus, every storage technology can only reach a finite amount of data within a desired
-access latency \cite[p.~2]{afmh}.
-
-% TODO: move this paragraph out of the introduction.
-% Hardware caches are managed by hardware directly.  They are generally opaque to the
-% operating system and other programs.  That is, software has no direct control over the
-% contents of a hardware cache.
-
-The most ubiquitous example for hardware caches is the hierarchy of
-% are the various levels (most commonly 2 or 3) of
-CPU caches that are found on almost all present-day CPUs.  They are designated L1 cache,
-L2 cache, and so on, with L1 being the fastest and smallest level.  The underlying storage
-for CPU caches is the main memory.
-
-There are more storage levels that comprise
-% constitute
-the \emph{memory hierarchy} of a computer along with CPU caches and main memory.  For
-example \glspl{hdd} and \glspl{ssd}. % Also: registers, internal buffers of HDDs and SSDs,
-                                     % (tapes?), ...
-% Focus on CPU caches.  Why?
-However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
-as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
-main memory, eliminating the need for swapping to disk under many workloads.
-
-I will focus on how to use CPU caches effectively and the resulting
-% enabled
-performance gains in this \article{}.
-% TODO: what about TLB?
-
-\section{Motivation} % So what?
-%    Three things Really Matter for performance. The first one is Algorithm, the second
-%    one is your code being Non-Blocking, and the third one is Data Locality."
-%       -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/
-
-% We will see that
-Two algorithms solving the same problem with the same asymptotic complexity (in the same
-\(\Theta(g(n))\)) may differ in performance by two orders of magnitude~\cite[p.~2]{afmh}.
-
-% TODO: how and how much can we speed up our programs by considering their interactions
-% with CPU caches?
-
-% TODO: move this.
-% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array to
-% compute the sum of its elements.  Before exiting, it prints the CPU time spend summing the
-% array.
-
-\begin{listing}
-   \inputminted[firstline=3]{c}{array-sum/array-sum.c}
-   \caption{This is C code}
-   \label{lst:array-sum}
-\end{listing}
-
-% This program is likely to have excellent spatial locality: the array...
-
-% \begin{figure}[htb]
-\begin{figure}
-   \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={label={Array size (KiB)}, length=0.8\textwidth},
-       y axis={label={Processor time (ms)}, length=6cm},
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         % See <https://tex.stackexchange.com/q/198323>.
-         data [read from file=array-sum/size-time.csv, separator=\space];
-   \end{tikzpicture}
-   \caption{This figure took way too long to create}
-   \label{fig:array-sum}
-\end{figure}
-
-See \cref{fig:array-sum}.  What's going on here?
-
-% \section{Basic Principles}
-% \section{Basic Concepts}
-\section{Key Terms}
-\subsection{Spatial Locality}
-Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
-memory into the CPU cache) in \emph{blocks}.
-
-Moving down the memory hierarchy, access latencies increase faster than the
-\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
-many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
-memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.
-
-\subsection{Temporal Locality}
-
-\subsection{Cache Hit}
-
-\subsection{Cache Miss}
-% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
-
-\subsection{External Memory Model}
-The \gls{emm} is a widely used
-% TODO: is it?)
-extension of the \gls{ram} model.
-
-\clearpage
-
-\printglossary[type=\acronymtype] % Print the list of acronyms.
-
-\printbibliography[heading=bibintoc]
-
-\end{document}
+TODO: I installed \texttt{msr-tools} and OProfile.
+% This seems to work.  It did affect performance.
+\begin{minted}[gobble=3]{bash}
+   $ sudo modprobe msr
+   $ sudo rdmsr -a 0xC0011022
+   600001800000000
+   600001800000000
+   $ sudo wrmsr -a 0xC0011022 0x600001800002000
+   $ sudo rdmsr -a -f '13:13' 0xC0011022
+   1
+   1
+   $ sudo rdmsr -a 0xC0011022
+   600001800002000
+   600001800002000
+\end{minted}
+%stopzone
+I think resetting the MSR to enable hardware prefetches doesn't work.  To get the same
+benchmmark results as before tweaking it, I have to reboot.
+\end{comment}
 

### 2017-05-21 motivation.tex
Split content into multiple files and tweak stuff

--- 2017-05-11 paper.tex
@@
-% Preamble {{{1
-\documentclass[a4paper]{scrartcl}
-
-\usepackage[utf8]{inputenc} % Assume this file is encoded in UTF-8.
-\usepackage[T1]{fontenc}    % Don't fake umlauts etc.
-\usepackage{lmodern}        % Use the lmodern font (http://tex.stackexchange.com/a/65103).
-\usepackage{microtype}      % Better microtypography (http://www.ctan.org/pkg/microtype)
-\usepackage{hyperref}       % Clickable hyperlinks, load before `glossaries`
-\usepackage{cleveref}       % Use `\cref{fig:foo}` instead of `figure~\ref{fig:foo}`.
-                            % Load after `hyperref`.  See
-                            % <https://tex.stackexchange.com/q/36295>
-\usepackage{comment}        % Comment out sections of text.
-\usepackage{mathtools}      % Improved facilities for typesetting mathematical formulae
-\usepackage{dirtytalk}      % ...
-
-% Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
-\usepackage[style=iso]{datetime2}
-
-% Source code listings with improved syntax highlighting
-\usepackage{minted}
-\usemintedstyle{pastie}
-
-\usepackage{color}
-\definecolor{bg}{rgb}{0.95,0.95,0.95}
-\setminted{bgcolor=bg}
-
-% Keep using ISO 8601 consistently, like an even more reasonable person.  See
-% <https://tex.stackexchange.com/q/231208>.
-\usepackage[date=edtf, urldate=edtf]{biblatex}
-\addbibresource{paper.bib}
-
-\usepackage[xindy, toc, acronym, nomain]{glossaries} % Load after `hyperref`.
-
-\makeglossaries
-
-\usepackage{tikz}
-\usetikzlibrary{datavisualization}
-% \usetikzlibrary{datavisualization.formats.functions}
-
-% See <https://tex.stackexchange.com/a/155317>, <https://tex.stackexchange.com/a/320521>
-% and <https://tex.stackexchange.com/a/75507>.
-% \usepackage{tikzscale}
-
-\newacronym{emm}{EMM}{external memory model}
-\newacronym{hdd}{HDD}{hard disk drive}
-\newacronym{ram}{RAM}{random access machine}
-\newacronym{smt}{SMT}{simultaneous multithreading}
-\newacronym{tlb}{TLB}{translation lookaside buffer}
-\newacronym{ssd}{SSD}{solid-state drive}
-
-\newcommand{\article}{article} % report, paper?
-
-% Top matter {{{1
-\title{Algorithms for Hardware Caches}
-\author{Lukas Waymann}
-
-% Body {{{1
-\begin{document}
-\maketitle
-\newpage
-
-\begin{abstract}
-   % What?
-   Typical present-day CPUs have two or more levels of caches.  This \article{} presents
-   the basic techniques used to optimize program performance based on knowledge about how
-   these hardware caches function.
-
-   The abstract \gls{emm} for memory hierarchies is explained and a small selection of
-   algorithms developed for it
-   % analyzed under it
-   are explored mathematically and empirically.
-\end{abstract}
-\newpage
-
-\tableofcontents
-\newpage
-
-\glsresetall % Reset the use status of all acronyms.
-
-\section{Introduction}
-% What is this paper about?  Why learn about them?  How much performance is at stake?
-
-% What are hardware caches?
-A hardware cache is a comparatively
-% relatively
-fast and small physical memory.  It stores a subset of the data present in slower, larger
-memory that is expected to be used again soon.  The purpose of this additional memory is
-to reduce the number of accesses to the underlying slower storage.
-
-% Why caches?
-There are fundamental reasons that having one single, uniform
-% homogeneous
-type of memory is not viable.  No signal can propagate faster than the speed of light.
-Thus, every storage technology can only reach a finite amount of data within a desired
-access latency \cite[p.~2]{afmh}.
-
-The most ubiquitous example for hardware caches is the hierarchy of
-% are the various levels (most commonly 2 or 3) of
-CPU caches that are found on almost all present-day CPUs.  They are designated L1 cache,
-L2 cache, and so on, with L1 being the fastest and smallest level.  The underlying storage
-for CPU caches is the main memory.
-
-There are more storage levels that comprise
-% constitute
-the \emph{memory hierarchy} of a computer along with CPU caches and main memory.  For
-example \glspl{hdd} and \glspl{ssd}. % Also: registers, internal buffers of HDDs and SSDs,
-                                     % (tapes?), ...
-% Focus on CPU caches.  Why?
-However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
-as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
-main memory, eliminating the need for swapping to disk under many workloads.
-
-I will focus on how to use CPU caches effectively and the resulting
-% enabled
-performance gains in this \article{}.
-% TODO: what about TLB?
-
 \section{Motivation} % So what?
-%    Three things Really Matter for performance.  The first one is Algorithm, the second
-%    one is your code being Non-Blocking, and the third one is Data Locality."
+% > Three things Really Matter for performance.  The first one is Algorithm, the second
+% > one is your code being Non-Blocking, and the third one is Data Locality."
 %      -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/
 
-% TODO: move this paragraph?
 Hardware caches are managed by hardware directly.  They are generally opaque to the
@@
 \end{comment}
-Effective use of hardware caches requires knowledge about how they work.
-% their architecture.
-Algorithms must be designed and implemented observing this knowledge.
-% their interactions with hardware caches.
+Effective use of hardware caches requires knowledge about \alts{how they work, their
+architecture}.  Algorithms must be designed and implemented observing \alts{this
+knowledge, their interactions with hardware caches}.
 
-% TODO: and *how* do I make effective use of hardware caches?
-
-% \section{Basic Principles}
-\section{Basic Concepts}
-% \section{Key Terms}
-% That sounds boring.
-
-\subsection{Spatial Locality}
-Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
-memory into the CPU cache) in \emph{blocks}.
-
-Moving down the memory hierarchy, access latencies increase faster than the
-\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
-many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
-memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.
-
-% TODO: move this.
-% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
-% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
-% summing the array.
-
-\begin{listing}
-   \inputminted[firstline=3]{c}{array-sum/array-sum.c}
-   \caption{This is C code}
-   \label{lst:array-sum}
-\end{listing}
-
-% This program is likely to have excellent spatial locality: the array...
-
-% \begin{figure}[htb]
-\begin{figure}
-   \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={label={Array size (KiB)}, length=0.8\textwidth},
-       y axis={label={Processor time (ms)}, length=6cm},
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         % See <https://tex.stackexchange.com/q/198323>.
-         data [read from file=array-sum/size-time.csv, separator=\space];
-   \end{tikzpicture}
-   \caption{This figure took way too long to create}
-   \label{fig:array-sum}
-\end{figure}
-
-% See \cref{fig:array-sum}.  What's going on here?
-
-\subsection{Temporal Locality}
-
-\subsection{Cache Hit}
-
-\subsection{Cache Miss}
-% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
-
-\subsection{Cache Line} % or Cache Block
-% On x86/x64, cache line is 64 bytes for many years now.
-%    -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
-
-\subsection{External Memory Model}
-The \gls{emm} is a widely used
-% TODO: is it?)
-extension of the \gls{ram} model.
-
-\clearpage
-
-\printglossary[type=\acronymtype] % Print the list of acronyms.
-
-\printbibliography[heading=bibintoc]
-
-\end{document}
+% How?
 

### 2017-05-21 paper.tex
Split content into multiple files and tweak stuff

--- 2017-05-11 paper.tex
@@
 
+% See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
+\usepackage{enumitem}
+\setlist{noitemsep}
+
+% TODO.  See <https://tex.stackexchange.com/q/119513>.
+\crefname{appsec}{Appendix}{Appendices}
+
+% Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
+\usepackage[style=iso]{datetime2}
+
+% TODO: will this just force LaTeX to make worse choices?
+% \usepackage[all]{nowidow}
+
 % Source code listings with improved syntax highlighting
@@
 
-\usepackage{color}
+% Define a background color for `minted` listings.  See <http://ctan.org/pkg/minted> and
+% <https://tex.stackexchange.com/q/150369>.
+\usepackage{xcolor}
 \definecolor{bg}{rgb}{0.95,0.95,0.95}
 \setminted{bgcolor=bg}
+% Don't shade the background when using `mintinline`.
+\setmintedinline{bgcolor={}}
 
-\usepackage{biblatex}
+% ...
+\usepackage[titletoc]{appendix}
+
+% Keep using ISO 8601 consistently, like an even more reasonable person.  See
+% <https://tex.stackexchange.com/q/231208>.
+\usepackage[date=edtf, urldate=edtf, seconds=true]{biblatex}
 \addbibresource{paper.bib}
 
-\usepackage[xindy, toc, acronym, nomain]{glossaries} % Load after `hyperref`.
+\usepackage[xindy, toc, acronym]{glossaries} % Load after `hyperref`.
 
 \makeglossaries
+\loadglsentries{glossary}
 
@@
 
-\newacronym{emm}{EMM}{external memory model}
-\newacronym{hdd}{HDD}{hard disk drive}
-\newacronym{ram}{RAM}{random access machine}
-\newacronym{smt}{SMT}{simultaneous multithreading}
-\newacronym{tlb}{TLB}{translation lookaside buffer}
-\newacronym{ssd}{SSD}{solid-state drive}
+\newcommand*{\article}{article} % report, paper?
 
-\newcommand{\article}{article} % report, paper?
+% Define a command that takes exactly 2 arguments, the first one defaulting to `1`.  The
+% second argument should be a list of items separated by `,`.  The list item at the
+% position specified by the first argument in printed.  See
+% <https://tex.stackexchange.com/a/99271>,
+% <http://mirrors.ctan.org/macros/generic/listofitems/listofitems-en.pdf>, and
+% <https://tex.stackexchange.com/q/276697>.
+\usepackage{listofitems}
+\newcommand*{\alts}[2][1]{%
+   \setsepchar{,}%
+   \readlist*\arg{#2}%
+   \arg[#1]%
+}
 
 % Top matter {{{1
-\title{Algorithms for Hardware Caches}
+\title{Basics of Hardware Cache Optimization}
+% \title{Algorithms for Hardware Caches}
+% \title{Basics of CPU Cache Optimization}
+% \title{Basics of Optimizing for Hardware Caches}
+% \title{Fundamentals of Optimizing for Hardware Caches}
+% \title{Introduction to Optimizations for Hardware Caches}
+% \title{Optimizing for Hardware Caches}
+% \subtitle{The Basics}
 \author{Lukas Waymann}
@@
 
-\section{Introduction}
-% What is this paper about?  Why learn about them?  How much performance is at stake?
+\input{intro}
+\input{motivation}
+\input{types}
+\input{basic-concepts}
 
-% What are hardware caches?
-A hardware cache is a comparatively
-% relatively
-fast and small physical memory.  It stores a subset of the data present in slower, larger
-memory that is expected to be used again soon.  The purpose of this additional memory is
-to reduce the number of accesses to the underlying slower storage.
+\section{Example: \texttt{std::vector} vs. \texttt{std::list}}
+\label{sec:vvl}
 
-% Why caches?
-There are fundamental reasons that having one single, uniform
-% homogeneous
-type of memory is not viable.  No signal can propagate faster than the speed of light.
-Thus, every storage technology can only reach a finite amount of data within a desired
-access latency \cite[p.~2]{afmh}.
+% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
+% https://en.wikipedia.org/wiki/Cache_replacement_policies
 
-% TODO: move this paragraph out of the introduction.
-% Hardware caches are managed by hardware directly.  They are generally opaque to the
-% operating system and other programs.  That is, software has no direct control over the
-% contents of a hardware cache.
-
-The most ubiquitous example for hardware caches is the hierarchy of
-% are the various levels (most commonly 2 or 3) of
-CPU caches that are found on almost all present-day CPUs.  They are designated L1 cache,
-L2 cache, and so on, with L1 being the fastest and smallest level.  The underlying storage
-for CPU caches is the main memory.
-
-There are more storage levels that comprise
-% constitute
-the \emph{memory hierarchy} of a computer along with CPU caches and main memory.  For
-example \glspl{hdd} and \glspl{ssd}. % Also: registers, internal buffers of HDDs and SSDs,
-                                     % (tapes?), ...
-% Focus on CPU caches.  Why?
-However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
-as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
-main memory, eliminating the need for swapping to disk under many workloads.
-
-I will focus on how to use CPU caches effectively and the resulting
-% enabled
-performance gains in this \article{}.
-% TODO: what about TLB?
-
-\section{Motivation} % So what?
-%    Three things Really Matter for performance. The first one is Algorithm, the second
-%    one is your code being Non-Blocking, and the third one is Data Locality."
-%       -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/
-
-% We will see that
-Two algorithms solving the same problem with the same asymptotic complexity (in the same
-\(\Theta(g(n))\)) may differ in performance by two orders of magnitude~\cite[p.~2]{afmh}.
-
-% TODO: how and how much can we speed up our programs by considering their interactions
-% with CPU caches?
-
-% TODO: move this.
-% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array to
-% compute the sum of its elements.  Before exiting, it prints the CPU time spend summing the
-% array.
+\section{External Memory Model}
+The \gls{emm} is a widely used
+% TODO: is it?
+extension of the \gls{ram} model.
 
 \begin{listing}
-   \inputminted[firstline=3]{c}{array-sum/array-sum.c}
+   \inputminted[firstline=8]{c}{array-sum/array-sum.c}
    \caption{This is C code}
@@
 
-See \cref{fig:array-sum}.  What's going on here?
+% See \cref{fig:array-sum}.  What's going on here?
 
-% \section{Basic Principles}
-% \section{Basic Concepts}
-\section{Key Terms}
-\subsection{Spatial Locality}
-Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
-memory into the CPU cache) in \emph{blocks}.
+\section{Cache-Oblivious Algorithms} % Cache-Oblivious Model
 
-Moving down the memory hierarchy, access latencies increase faster than the
-\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
-many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
-memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.
+\section{Parallel Computing}
 
-\subsection{Temporal Locality}
+\subsection{Cache Coherence} % Coherence or coherency?
+% https://en.wikipedia.org/wiki/Cache_coherence
+% Dirty.  Invalid.
 
-\subsection{Cache Hit}
+\subsection{False Sharing}
 
-\subsection{Cache Miss}
-% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
-
-\subsection{External Memory Model}
-The \gls{emm} is a widely used
-% TODO: is it?)
-extension of the \gls{ram} model.
+\clearpage
+% \appendix
+\begin{appendices}
+   % See <https://tex.stackexchange.com/q/119513>.
+   \crefalias{section}{appsec}
+   \input{appendices}
+\end{appendices}
 

### 2017-05-21 tex/basic-concepts.tex
Move .tex files into subdirectory

\section{Basic Concepts}
% \section{Basic Principles}
% \section{Key Terms} % That sounds boring.
Some architectural properties of hardware caches lead to important concepts for using them
effectively.

\subsection{Cache Line} % or Cache Block
% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
%
% > On x86/x64, cache line is 64 bytes for many years now.
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
%      -- Drepper, p. 15
% > It is not possible for a cache to hold partial cache lines.
%      -- Drepper, p. 16
\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
{scott-meyers-talk}.%
\footnote{%
% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
% > cache 256 KB in size, with 128-byte cache blocks.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
% TODO: what about the block size of main memory?  Should it be the same?
Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
line}~\cite[p.~11]{pentium4}.}
\begin{comment}
   \multiplefootnoteseparator%
   % See <https://tex.stackexchange.com/a/71015>.
   \footnote{This can also be checked on the command line:
   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
\end{comment}
% > It means that as soon as you've accessed any single byte in a cache line, all the
% > other 63 bytes are already in L1 cache
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
\alts{%
   This means accessing a single uncached 32-bit integer entails loading another 60
   adjacent bytes.,
   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
   well.},
   {When, for example, accessing a single byte that isn't already cached, another 63
   adjacent bytes will be loaded.}
}
% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.

My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

% The reason why the loops take the same amount of time has to do with memory.  The
% running time of these loops is dominated by the memory accesses to the array, not by the
% integer multiplications.
%    -- https://igoro.com/archive/gallery-of-processor-cache-effects/
This is because the run times are \alts{primarily due to, dominated by} memory accesses.
Up to a step size of 8, every 64-byte line has to be loaded.  At 16, the values we modify
are 128 bytes apart,%
\footnote{16 \texttt{int64\_t} values of 8 bytes each}
so every other cache line is skipped.  At 32, three out of four cache lines are skipped,
and so on~\cite[cf.][example 2]{gallery}.

\begin{listing}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{listing}

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
                                 8 as \textbf{8}, 16, 64, 256, 1024}},
                      minor={at={32, 128, 512}}},
               label={Step size}, length=0.8\textwidth},
       y axis={logarithmic,
               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
               label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=line-size/line-size.csv, separator=\space];
   \end{tikzpicture}
   \caption{Processor times for running \cref{lst:line-size}}
   \label{fig:line-size}
\end{figure}

\subsection{Prefetching}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching
\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

\subsection{Associativity}
% https://en.wikipedia.org/wiki/CPU_cache#Associativity

\subsection{Cache Hits and Misses}
% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?

\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

Moving down the memory hierarchy, access latencies increase faster than the
\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
memory is almost as fast as access to a single bit},~\cite[2]{afmh}.

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
% summing the array.

\subsection{Temporal Locality}
TODO~\cite{drepper2007}.

\subsection{Memory Access Pattern}
\label{sec:map}

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-21 tex/intro.tex
Move .tex files into subdirectory

\section{Introduction}
% What is this paper about?  Why learn about them?  How much performance is at stake?

% What are hardware caches?  Why caches?
A hardware cache is a \alts{comparatively, relatively} fast and small physical memory.  It
stores a subset of the data present in slower, larger memory that is expected to be used
again soon.  The purpose of this additional memory is to reduce the number of accesses to
the underlying slower storage.

% Hardware caches aren't going away.
There are fundamental reasons that having one single, \alts{uniform, homogeneous} type of
memory is not viable.  No signal can propagate faster than the speed of light.  Thus,
every storage technology can only reach a finite amount of data within a desired access
latency~\cite[2]{afmh}.

The most ubiquitous example for hardware caches \alts{is the hierarchy, are the various
levels (most commonly 2 or 3)} of CPU caches that are found on almost all present-day
CPUs.  They are designated L1 cache, L2 cache, and so on, with L1 being the fastest and
smallest level.  The underlying storage for CPU caches is the main memory.

There are more storage levels that \alts{comprise, constitute} the \emph{memory hierarchy}
of a computer along with CPU caches and main memory.  For example \glspl{hdd} and
\glspl{ssd}.
% Also: registers, internal buffers of HDDs and SSDs, (tapes), ...
% Focus on CPU caches.  Why?
However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
main memory, eliminating the need for swapping to disk under many workloads.

I will focus on how to use CPU caches effectively and the \alts{enabled, resulting}
performance gains in this \article{}.

% TODO: what about TLB?

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-21 tex/types.tex
Move .tex files into subdirectory

\section{Types of CPU Caches}
% > Most modern desktop and server CPUs have at least three independent caches: an
% > instruction cache to speed up executable instruction fetch, a data cache to speed up
% > data fetch and store, and a translation lookaside buffer (TLB) used to speed up
% > virtual-to-physical address translation for both executable instructions and data.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Overview
% > There are three common types of CPU caches: ...
%      -- Scott Meyers (talk at code::dive)
Current x86 CPUs \alts{generally, typically, commonly} have three main types of caches:
data caches, instruction caches, and \glspl{tlb}%
~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m07s}{11:07}]{scott-meyers-talk}.
Some caches are used for data as well as instructions and are called \emph{unified}.%
~\cite[20]{drepper2007}.
\alts{{A processor may have multiple caches of each type, which}, {Multiple caches of each
type may be present, and}} are organised into numerical \emph{levels}
\alts{{starting at 1, the smallest and fastest level,},}
based on their size and speed.
% Each added level is bigger and slower than its predecessor.
% The smallest and fastest is level 1.

% TODO?  The reason to have multiple levels...

% > Often there are separate L1 caches for instructions and data
%      -- Algorithms for Memory Hierarchies, page 3
% > Systems nowaeays have at-least two levels of cache
%      -- Algorithms for Memory Hierarchies, page 172
% > [T]he caches from L2 on are unified caches which contain both code and data
%      -- Drepper, p. 31
% > Later Intel models have shared L2 caches for dual-core processors.  For quad-core
% > processors we have to deal with separate L2 caches for each pair of two cores.
%      -- Drepper, p. 35

% Terminology / Nomenclature.
In practice, a \alts{currently, presently} representative%
\footnote{%
   % https://en.wikipedia.org/wiki/Bobcat_(microarchitecture)
   E.g. for AMD Family 14h processors~\cite[30--32]{14h},
   % https://en.wikipedia.org/wiki/List_of_AMD_CPU_microarchitectures
   % https://en.wikipedia.org/wiki/Zen_(microarchitecture)
   % 32 KiB L1d, 64 KiB L1i, 512 KiB L2, 8 to 16 MiB L3
   AMD Zen (17h)~\cite{zen}, and
   % https://en.wikipedia.org/wiki/Kaby_Lake
   % https://en.wikipedia.org/wiki/Skylake_(microarchitecture)
   % 32 KiB L1d, 32 KiB L1i, 256 KiB L2, 2 to 8 MiB L3
   Intel Skylake desktop processors%
   ~\cite[figure 2-1, table 2-4]{skylake}
   % ~\cite[figure 2-1, \pno~2-2, table 2-4, \pno~2-6]{skylake}.
   % ~\cite[{2-2}, {2-6}]{skylake}.
   % <https://en.wikipedia.org/wiki/Bulldozer_(microarchitecture)> is too weird.
}
x86 cache hierarchy consists of:
\begin{itemize}
   % https://en.wikipedia.org/wiki/Cache_hierarchy#Shared_versus_private
   \item Separate level 1 data and instruction caches of 32 to 64 KiB for each core
      (denoted \gls{l1d} and \gls{l1i} by  \textcite[14--15]{drepper2007}).
      % TODO?  Why have a separate instruction cache?
      Machine instructions in \gls{l1i} are already decoded%
      ~\cite[31, 56]{drepper2007}.
      % ~\cite[14, 31, 56]{drepper2007}.
   % \item A level 2 cache for \say{both code and data}~\cite[31]{drepper2007}.
   \item A unified \gls{l2} cache of 256 to 512 KiB for each core.
   \item Often a unified \gls{l3} cache of 2 to 16 MiB shared between all cores.
   \item TODO: Some \glspl{tlb} I guess.
\end{itemize}

% \subsection{Access Times}
% http://ithare.com/infographics-operation-costs-in-cpu-clock-cycles/
% http://www.getitwriteonline.com/archive/040201hyphadj.htm
\alts{Estimates, Order-of-magnitude estimates} of typical access latencies \alts[2]{are as
follows, are given by \textcite{ithare-cycles}.}%
\footnote{%
   Intel~\cite[table 2-4]{skylake},
   \textcites
   % {ithare-paadl}{ithare-wisdoms}
   [\href{https://youtu.be/WDIkqP4JbkE?t=17m52s}{17:52}, slide 18]{scott-meyers-talk}
   [2--3, 171]{afmh}[16, 20--21]{drepper2007} all give comparable numbers for various
   architectures.
   % [\ppno~16, 20--21, fig. 3.10]{drepper2007}
}

\begin{center}
   \begin{tabular}{ r | c c c c }
             & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
      Cycles & 3--4      & 10--12   & 30--70   & 100--150
   \end{tabular}
\end{center}
%
% These are taken from~\textcite{ithare-cycles} but comparable numbers are given by

% > [Instruction] cache is much less problematic than the data cache.
%      -- Drepper, p. 31
The biggest target for optimizations is the data cache.  \say{[Instruction] cache is much
less problematic}~\cite[31]{drepper2007} and optimizations for data and instruction cache
tend to improve \gls{tlb} usage as well%
~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m53s}{11:53}]{scott-meyers-talk}.

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-24 tex/cpu-info.tex
Add awesome plot, expand content, tweak stuff

--- 2017-05-21 appendices.tex
@@
 
-% Performance Analysis
-% Run Time Analysis
-% Timing Methodology
-% Profiling Methodology
-% Profiling Pitfalls
-% Run Time Measuring
-% Benchmarking Methodology
-% Monitoring
-\section{Profiling Methodology}
-% https://en.wikipedia.org/wiki/Profiling_(computer_programming)
-% https://en.wikipedia.org/wiki/Processor_affinity
-% On `/usr/bin/time` for benchmarks: <http://stackoverflow.com/q/9006596>.
-% Andrei Alexandrescu - Writing Fast Code: <https://youtu.be/vrfYLlR8X8k>.
-Plenty of things can go wrong when measuring \alts{execution, run} times.
-\begin{itemize}
-   \item CPU frequency scaling and boosting
-      % Not a problem when measuring cycles.
-   \item Interrupts
-      % Taking the minimum helps.
-   \item CPU jumping
-      % See <https://youtu.be/vrfYLlR8X8k?t=33m34s>.
-   \item Context switches
-   \item Warmth of cache
-      % See <https://youtu.be/vrfYLlR8X8k?t=41m00s>.
-   % \item ...
-\end{itemize}
-Some effects will \alts{add noise to, pollute} the results or render them irreproducible,
-others will invalidate them completely.
-
-\begin{comment}
-\section{Disabling the Prefetcher}
-My initial results from from running \cref{lst:array-sum} were quite different.  See TODO.
-
-TODO: I installed \texttt{msr-tools} and OProfile.
-% This seems to work.  It did affect performance.
-\begin{minted}[gobble=3]{bash}
-   $ sudo modprobe msr
-   $ sudo rdmsr -a 0xC0011022
-   600001800000000
-   600001800000000
-   $ sudo wrmsr -a 0xC0011022 0x600001800002000
-   $ sudo rdmsr -a -f '13:13' 0xC0011022
-   1
-   1
-   $ sudo rdmsr -a 0xC0011022
-   600001800002000
-   600001800002000
-\end{minted}
-%stopzone
-I think resetting the MSR to enable hardware prefetches doesn't work.  To get the same
-benchmmark results as before tweaking it, I have to reboot.
-\end{comment}
-
 % vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-24 tex/methodology.tex
Add awesome plot, expand content, tweak stuff

% Programs: `time`, bad; `\time -v`, better; oprofile(1); gprof(1); perf; Valgrind's
% cachegrind?; ...

% Performance Analysis; Run Time Analysis; Timing Methodology; Profiling Pitfalls; Run
% Time Measuring; Benchmarking Methodology; Performance Monitoring
\section{[TODO] Profiling Methodology}
\label{app:meth}

Plenty of things can go wrong when
profiling.
% measuring \alts{execution, run} times.
These are the ones I was aware of and tried to account for while doing measurements for
this \article.
\begin{itemize}
   \item \emph{CPU frequency scaling and boosting}.
      CPU frequencies are usually dynamic these days and automatically adjusted based on
      at least workload and temperature.  This should be disabled when measuring execution
      times.
      \begin{minted}[autogobble]{text}
         # TODO
      \end{minted}
      % See [5].  Not a problem when measuring cycles.
   \item \emph{Interrupts and context switches.}  The process being timed has to share
      resources with other processes and the operating system.  A context switch will not
      only increase wall time, it will also increase cycles because of TLB flushes and
      cache evictions. % See <https://en.wikipedia.org/wiki/Context_switch>.

      Processor shielding, % See [5].
      taking the minimum of several measurements, and assigning a very high priority can
      \alts{help here, mitigate these problems}.
      % This is likely better than using nice(1).
      \begin{minted}[autogobble]{text}
         # chrt -f 99 \time -v ./program
      \end{minted}
   \item \emph{CPU jumping.}
      % See <https://youtu.be/vrfYLlR8X8k?t=33m34s>.
      CPU pinning (setting processor/thread affinity) may help. % See [2] and [5].
   \item \emph{Cache warmth.}
      % See <https://youtu.be/vrfYLlR8X8k?t=41m00s>.
      When comparing different solutions on the same data set, later ones may benefit from 
      data having been loaded into cache already.  Checking whether swapping the order of
      measurements changes the results is a good idea.
\end{itemize}
These effects may \alts{add noise to, pollute} the results, render them irreproducible, or
invalidate them completely.

General mitigation/alleviation strategy: take the minimum execution time of \alts{all, a
number of} runs; all noise is additive (TODO: not the one caused by the cache being hot).
% See [7].

\subsection{Measuring CPU cycles}
\label{app:cycles}

Frequency scaling isn't a problem here.  I used the \mintinline{text}{ocount(1)} event
counting tool added to the OProfile (TODO: cite) project in version 0.9.9.  [TODO] This
uses \emph{hardware performance counters}.
% `perf(1)` seems to be a newer alternative to OProfile [8].

\begin{minted}[autogobble]{text}
   ophelp
   sudo chrt -f 99 ocount -e CPU_CLK_UNHALTED program
\end{minted}

TODO~\cite[78\psqq]{drepper2007}.

% [1]:  https://en.wikipedia.org/wiki/Profiling_(computer_programming)
% [2]:  https://en.wikipedia.org/wiki/Processor_affinity
% [3]:  https://en.wikipedia.org/wiki/OProfile
% [4]:  https://en.wikipedia.org/wiki/Perf_(Linux)
% [5]:  https://github.com/JuliaCI/BenchmarkTools.jl/blob/master/doc/linuxtips.md
% [6]:  http://stackoverflow.com/q/9006596 "On `/usr/bin/time` for benchmarks and `perf`"
% [7]:  https://youtu.be/vrfYLlR8X8k "Andrei Alexandrescu - Writing Fast Code"
% [8]:  http://oprofile-list.sf.narkive.com/rVbJlHdX/oprofile-vs-perf
% [9]:  http://oprofile.sourceforge.net/about/
% [10]: https://en.wikipedia.org/wiki/Hardware_performance_counter
% [11]: http://oprofile.sourceforge.net/doc/index.html

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-24 tex/no-prefetch.tex
Add awesome plot, expand content, tweak stuff

--- 2017-05-21 appendices.tex
@@
-\section{Reading Information About the CPU}
-\label{app:cpuinfo}
-% https://stackoverflow.com/q/7281699
-% https://unix.stackexchange.com/q/167038
-% https://superuser.com/q/55776
-There are many ways to display information about the processor(s) the operating system is
-running on.  Among others, the \mintinline{text}{lscpu(1)} and
-\mintinline{text}{getconf(1)} programs and the \mintinline{text}{/proc/cpuinfo}
-pseudo-file on Linux.  This is \alts{how I checked, what I used to check} my CPU's cache
-line sizes, for example:
-\begin{minted}[gobble=3]{bash}
-   $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
-   64
-   64
-\end{minted}
-%stopzone
-
-% Performance Analysis
-% Run Time Analysis
-% Timing Methodology
-% Profiling Methodology
-% Profiling Pitfalls
-% Run Time Measuring
-% Benchmarking Methodology
-% Monitoring
-\section{Profiling Methodology}
-% https://en.wikipedia.org/wiki/Profiling_(computer_programming)
-% https://en.wikipedia.org/wiki/Processor_affinity
-% On `/usr/bin/time` for benchmarks: <http://stackoverflow.com/q/9006596>.
-% Andrei Alexandrescu - Writing Fast Code: <https://youtu.be/vrfYLlR8X8k>.
-Plenty of things can go wrong when measuring \alts{execution, run} times.
-\begin{itemize}
-   \item CPU frequency scaling and boosting
-      % Not a problem when measuring cycles.
-   \item Interrupts
-      % Taking the minimum helps.
-   \item CPU jumping
-      % See <https://youtu.be/vrfYLlR8X8k?t=33m34s>.
-   \item Context switches
-   \item Warmth of cache
-      % See <https://youtu.be/vrfYLlR8X8k?t=41m00s>.
-   % \item ...
-\end{itemize}
-Some effects will \alts{add noise to, pollute} the results or render them irreproducible,
-others will invalidate them completely.
-
-\begin{comment}
 \section{Disabling the Prefetcher}
@@
 TODO: I installed \texttt{msr-tools} and OProfile.
-% This seems to work.  It did affect performance.
+% This seems to work.  It did affect performance.  See
+% <https://community.amd.com/thread/180725>.
 \begin{minted}[gobble=3]{bash}
@@
    600001800002000
+   $ sudo ocount -e PREFETCH_INSTRUCTIONS_DISPATCHED ./a.out
 \end{minted}
@@
 benchmmark results as before tweaking it, I have to reboot.
-\end{comment}
 

### 2017-05-24 tex/types.tex
Add awesome plot, expand content, tweak stuff

@@
 
+My laptop's AMD E-450 CPU has cores with \alts{an \gls{l1d} cache of 32 KiB, 32 KiB of
+\gls{l1d} cache} and a unified \gls{l2} cache of 512 KiB each.%
+\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
+We can both verify these sizes and get \alts{a reasonably good measure, a rough measure,
+an approximation} of the access times by profiling
+\cref{lst:access-times}
+% the \alts{program, listing} below
+for different values of \mintinline{text}{SIZE}.%
+% FIXME: the label is wrong: should be "Appendix" but is "Section".
+\footnote{\Cref{app:cycles} details how.}
+% It reads random memory addresses
+% The program randomly reads
+% locations
+% The program repeatedly \alts{reads, accesses} random elements
+This program repeatedly \alts{reads, accesses} elements
+from \alts[2]{an array of the configured size, a thusly sized array}
+in random \alts{order, sequence, succession}.
+To do this
+\alts{with minimal overhead, efficiently}, the array \alts{is first set up, acts} as a
+circular, singly linked list where every element except the last \alts{points to a, has a}
+random successor.  When compiled with \mintinline{text}{-DBASELINE}, only this
+initialization is done.
+\Cref{fig:access-times} shows the deltas of CPU cycles used when and when not having
+defined \mintinline{text}{BASELINE}.
+I divided by \mintinline{text}{N} to get the cycles spend on each loop iteration.
+
+% XXX: hacks!  Use the figure environment so that LaTeX won't display this listing after
+% the plot showing the results of profiling it.  "LaTeX [only] keeps all floats of the
+% same *type* in order" [1].  Use `\captionof` to label the listing correctly as a
+% listing.
+%
+% [1]: https://tex.stackexchange.com/q/127742/#comment290982_127744
+\begin{comment}
+   \begin{figure}
+      \inputminted[firstline=9]{c}{access-times/access-times.c}
+      \captionof{listing}{TODO}
+      \label{lst:access-times}
+   \end{figure}
+\end{comment}
+
+% \newenvironment{code}{\captionsetup{type=listing}}{}
+% \begin{code}
+%    \inputminted[firstline=9]{c}{access-times/access-times.c}
+%    \caption{TODO}
+%    \label{lst:access-times}
+% \end{code}
+
+% Actually, don't use a float.  The listing should be allowed to span multiple pages which
+% floats aren't.
+%
+% [1]: https://tex.stackexchange.com/q/14522/#comment484569_75880
+% [2]: https://tex.stackexchange.com/q/175650
+%      "How to allow page break inside a float environment?"
+% [3]: https://tex.stackexchange.com/q/12428
+\inputminted[firstline=9]{c}{access-times/access-times.c}
+\captionof{listing}{}
+% \captionof{listing}{Random Reads\label{lst:access-times}}
+\label{lst:access-times}
+
+% https://tex.stackexchange.com/a/82473
+% \global\csname @topnum\endcsname 0
+
+% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
+% interrupting the source code listing.
+\begin{figure}[hbp]
+   \centering
+   \begin{tikzpicture}
+      \datavisualization
+      [scientific axes=clean,
+       x axis={logarithmic,
+               ticks={major={at={2048 as 2, 8192 as 8, 32768 as \textbf{32},
+                                 131072 as 128, 524288 as \textbf{512}, 2097152 as 2048,
+                                 8388608 as 8192, 33554432 as 32768, 134217728 as
+                                 % $128\cdot 2^{10}$}}},
+                                 % $2^{17}$}}},
+                                 131072}}},
+               grid={at={32768, 524288}},
+               label={Array Size (KiB)}, length=0.8\textwidth},
+       y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
+       % y axis={logarithmic,
+       %         ticks={major={at={3, 15, 23, 190}}},
+       %         grid={at={15, 23}},
+       %         label={Cycles / Element}, length=6cm},
+       visualize as scatter,
+       scatter={style={mark=*, mark options={scale=.65}}}]
+         data [read from file=access-times/access-times.csv, separator=\space];
+   \end{tikzpicture}
+   % \caption{Sequential Read Access}
+   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
+   \caption{Access Times for Random Reads}
+   \label{fig:access-times}
+\end{figure}
+
 % vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-24 access-times/access-times-seq.c

@@ -1,0 +1,16 @@
+// vim: set tw=72 sts=-1 sw=3 et cms=//%s
+
+#include "stdint.h"  // int64_t
+#include "stdio.h"   // printf
+#include "stdlib.h"  // calloc
+
+#define N 1000000000
+
+int main() {
+   int64_t* array = (int64_t*)calloc(SIZE, sizeof(int64_t));
+   int sum;
+   for (size_t n = 0; n < N; ++n) {
+      sum += array[(8 * n) % SIZE];
+   }
+   printf("%d\n", sum);  // Better do this...
+}

### 2017-05-24 access-times/access-times.c

@@ -1,0 +1,36 @@
+// vim: set tw=72 sts=-1 sw=3 et cms=//%s
+
+#include <stdint.h>  // int64_t, size_t
+#include <stdio.h>   // printf
+#include <stdlib.h>  // rand
+
+// Cf. Figure 3.4 on page 17 of Drepper's article.
+
+#define N 100000000  // 100 million
+
+struct elem {
+   struct elem *next;
+} array[SIZE];
+
+int main() {
+   for (size_t i = 0; i < SIZE - 1; ++i) array[i].next = &array[i + 1];
+   array[SIZE - 1].next = array;
+   // Fisher-Yates shuffle the array.
+   for (size_t i = 0; i < SIZE - 1; ++i) {
+      size_t j = i + rand() % (SIZE - i);  // j is in [i, SIZE).
+      struct elem temp = array[i];  // Swap array[i] and array[j].
+      array[i] = array[j];
+      array[j] = temp;
+   }
+#ifndef BASELINE
+   int64_t dummy = 0;
+   struct elem *i = array;
+   for (size_t n = 0; n < N; ++n) {
+      dummy += (int64_t)i;
+      i = i->next;
+   }
+   printf("%d\n", dummy);
+#else
+   asm("" ::: "memory");
+#endif
+}

### 2017-05-24 access-times/access-times.c
Remove `asm` declaration
Looks like it's not needed to prevent optimization.  See
<https://stackoverflow.com/q/14449141>.

@@ -30,7 +30,5 @@ int main() {
       i = i->next;
    }
    printf("%d\n", dummy);
-#else
-   asm("" ::: "memory");
 #endif
 }

### 2017-05-24 tex/types.tex
Expand content and tweak stuff

@@
 initialization is done.
-\Cref{fig:access-times} shows the deltas of CPU cycles used when and when not having
-defined \mintinline{text}{BASELINE}.
-I divided by \mintinline{text}{N} to get the cycles spend on each loop iteration.
 
@@
 
+% XXX: hacks!  FIXME: I want to have the same amount of space after adding a caption with
+% \captionof to a minted listing as there would be after a normal floating minted listing.
+% See <https://tex.stackexchange.com/a/162074>.
+% \captionsetup[listing]{aboveskip=5pt, belowskip=\baselineskip}
+
 % Actually, don't use a float.  The listing should be allowed to span multiple pages which
@@
 % [3]: https://tex.stackexchange.com/q/12428
-\inputminted[firstline=9]{c}{access-times/access-times.c}
+\begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
+   \inputminted[firstline=12]{c}{access-times/access-times.c}
    \captionof{listing}{}
@@
    \label{lst:access-times}
+\end{center}
 
@@
 % \global\csname @topnum\endcsname 0
+
+\Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
+when not having defined \mintinline{text}{BASELINE}.  That is, only the cycles used by the
+main loop are counted, not those for initialization.
+I divided by \mintinline{text}{N} to get the cycles spend \alts{per, on each} loop
+iteration.
+
+Up to 32 KiB, each access takes almost exactly 3 cycles.%
+\footnote{TODO: table of actual data in appendix?}
+This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
+the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
+with other processes and the operating system, so some of our data gets evicted.  The
+first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
+KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
+less \gls{l1d} hits and an \gls{l2} access time of around 25 cycles.
+
+The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
+increase when the array size matches that of the \gls{l2} is more striking than for the
+\gls{l1d} before; possibly because \gls{l2} is a unified cache that also holds
+instructions.  Eventually, more and more accesses go to main memory, \alts{causing,
+incurring} delays of up to 200 cycles.
+
+TODO~\cite[cf.][17]{drepper2007}.
 

### 2017-05-24 access-times/access-times.c

@@ -4,7 +4,10 @@
 #include <stdio.h>   // printf
 #include <stdlib.h>  // rand
 
-// Cf. Figure 3.4 on page 17 of Drepper's article.
+// Cf. Figure 3.4 on page 17 of Drepper's article.  The results of
+// profiling this program are heavily influenced by what other processes
+// are active at the same time.  It's best to kill as many as possible;
+// e.g. Firefox, which tends to always use some CPU time.
 
 #define N 100000000  // 100 million
 

### 2017-05-26 tex/types.tex
Add amazing table and tweak things

@@
    \inputminted[firstline=12]{c}{access-times/access-times.c}
-   \captionof{listing}{}
+   % \captionof{listing}{}
+   % \captionof{listing}{Randomly Access Elements of \mintinline{text}{array}}
+   \captionof{listing}{Randomly Read Array Elements}
    % \captionof{listing}{Random Reads\label{lst:access-times}}
@@
 
-\Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
-when not having defined \mintinline{text}{BASELINE}.  That is, only the cycles used by the
-main loop are counted, not those for initialization.
-I divided by \mintinline{text}{N} to get the cycles spend \alts{per, on each} loop
-iteration.
+% \Cref{fig:access-times} shows the \alts{difference, deltas} of CPU cycles used when and
+% when not having defined \mintinline{text}{BASELINE}.
+\Cref{fig:access-times} shows the extra CPU cycles used by \cref{lst:access-times}
+\alts[2]{above, in addition to, compared to} the \mintinline{text}{BASELINE} version for
+different array sizes.
+That is, only the cycles used by the main loop are \alts{given, counted}, not those for
+initialization.  I divided by \mintinline{text}{N} to get the cycles spent \alts{per, on
+each} loop iteration.
 
 Up to 32 KiB, each access takes almost exactly 3 cycles.%
-\footnote{TODO: table of actual data in appendix?}
+\footnote{The numerical results are shown in \cref{tab:access-times}.}
 This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
@@
 The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
-increase when the array size matches that of the \gls{l2} is more striking than for the
-\gls{l1d} before; possibly because \gls{l2} is a unified cache that also holds
-instructions.  Eventually, more and more accesses go to main memory, \alts{causing,
-incurring} delays of up to 200 cycles.
-
-TODO~\cite[cf.][17]{drepper2007}.
+increase when the array size matches that of the \gls{l2}, 512 KiB, is \alts{greater, more
+striking} than for the \gls{l1d} before; possibly because \gls{l2} is a unified cache that
+also holds instructions.  Eventually, more and more accesses go to main memory,
+\alts{causing, incurring} delays of up to 200 cycles~\cite[cf.][17]{drepper2007}.
 
@@
    \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={logarithmic,
-               ticks={major={at={2048 as 2, 8192 as 8, 32768 as \textbf{32},
-                                 131072 as 128, 524288 as \textbf{512}, 2097152 as 2048,
-                                 8388608 as 8192, 33554432 as 32768, 134217728 as
+      \datavisualization[%
+         scientific axes=clean,
+         x axis={%
+            logarithmic,
+            ticks={major={
+               /pgf/number format/int detect,
+               at={
+                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
+                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
+                  8388608 as \pgfmathprintnumber{8192},
+                  33554432 as \pgfmathprintnumber{32768},
+                  134217728 as \pgfmathprintnumber{131072},
                   % $128\cdot 2^{10}$}}},
                   % $2^{17}$}}},
-                                 131072}}},
+            }}},
             grid={at={32768, 524288}},
-               label={Array Size (KiB)}, length=0.8\textwidth},
+            label={Array Size (KiB)},
+            length=0.8\textwidth},
          y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
-       % y axis={logarithmic,
-       %         ticks={major={at={3, 15, 23, 190}}},
-       %         grid={at={15, 23}},
-       %         label={Cycles / Element}, length=6cm},
          visualize as scatter,
@@
    \end{tikzpicture}
-   % \caption{Sequential Read Access}
    % \caption{Access Times for Random Reads (\Cref{lst:access-times})}

### 2017-05-26 paper.tex
Move .tex files into subdirectory
Add awesome plot, expand content, tweak stuff
Add amazing table and tweak things
Use varioref in addition to hyperref and cleveref
https://ctan.org/pkg/varioref
Shorten some links

@@
 % Preamble {{{1
-\documentclass[a4paper]{scrartcl}
+% There's some issue that causes captions like "Listing 2.: Foo".  setting the `numbers`
+% option like this fixes it.  See <https://tex.stackexchange.com/q/29181>.
+\documentclass[a4paper,numbers=noenddot]{scrartcl}
 
@@
 \usepackage{lmodern}        % Use the lmodern font (http://tex.stackexchange.com/a/65103).
-\usepackage{microtype}      % Better microtypography (http://www.ctan.org/pkg/microtype)
-\usepackage{hyperref}       % Clickable hyperlinks, load before `glossaries`
-\usepackage{cleveref}       % Use `\cref{fig:foo}` instead of `figure~\ref{fig:foo}`.
-                            % Load after `hyperref`.  See
-                            % <https://tex.stackexchange.com/q/36295>
+\usepackage{microtype}      % Better microtypography (http://ctan.org/pkg/microtype)
 \usepackage{comment}        % Comment out sections of text.
@@
 \usepackage{dirtytalk}      % ...
+
+% Prevent floats from moving into another section (https://ctan.org/pkg/placeins).
+\usepackage[above, below, section]{placeins}
+
+% Stack for cross-referencing.  XXX: the packages have to be loaded in this order (see the
+% cleveref manual).
+% * The [varioref] package adds commands that automatically print the target's page number
+%   iff it is on a different page than the reference.
+% * [hyperref] turns references into clickable hyperlinks.
+% * Among other things, [cleveref] adds commands that include the type of the referenced
+%   object.  Instead of `figure~\ref{fig:foo}`, one can write `\cref{fig:foo}`.  It also
+%   enhances the varioref commands accordingly.
+% See [1], [2], and [3].
+%
+% [varioref]: https://ctan.org/pkg/varioref
+% [hyperref]: https://ctan.org/pkg/hyperref
+% [cleveref]: https://ctan.org/pkg/cleveref
+% [1]: https://en.wikibooks.org/wiki/LaTeX/Labels_and_Cross-referencing
+% [2]: https://tex.stackexchange.com/q/83037
+%      "Difference between ref, varioref and cleveref."
+% [3]: https://tex.stackexchange.com/q/36295
+%      "Cross-reference packages: which to use, which conflict?"
+\usepackage{varioref}
+\usepackage{hyperref} % XXX: load before `glossaries`
+\usepackage{cleveref} % Use `\cref{fig:foo}` instead of `figure~\ref{fig:foo}`.
+
+% Stack for generating tables from CSV files.  Loading is done by [pgfplotstable].  It can
+% also round, format and post-process data.  See [1].
+% [siunitx] adds the `S` column type for advanced aligning in tables like around the
+% decimal marker.  Getting this to work is a bit tricky; see [2].  There is some support
+% for aligning around the decimal marker in [pgfplotstable] (see [3]), but I couldn't get
+% it to work.  One can also right-align numbers relatively to each other but still center
+% the column as a whole with [siunitx] (see [4]).
+% [pgfplotstable]: https://ctan.org/pkg/pgfplotstable
+% [siunitx]: https://ctan.org/pkg/siunitx
+% [1]: https://tex.stackexchange.com/q/146716
+% [2]: https://tex.stackexchange.com/q/131081
+% [3]: https://tex.stackexchange.com/q/276485
+% [4]: https://tex.stackexchange.com/q/9203
+\usepackage{pgfplotstable, siunitx}
+% These packages are recommended by [pgfplotstable].  The booktabs package adds the
+% `\toprule`, `\midrule`, and `\bottomrule` commands for better looking tables.  See [5].
+\usepackage{booktabs, array, colortbl}
+% [booktabs]: https://ctan.org/pkg/booktabs
+% [5]: https://en.wikibooks.org/wiki/LaTeX/Tables#Professional_tables
+
+% Use a consitent group/thousands separator for siunitx and pgf.
+\pgfkeys{/pgf/number format/.cd,
+   set thousands separator=\,
+}
+% \sisetup{
+%    group-separator={,}
+% }
+
+% Provides the `\captionof` command for typesetting captions outside of floats, which is
+% an extension of th one provided by [capt-of](https://ctan.org/pkg/capt-of).  Includes
+% the [newfloat package](https://ctan.org/pkg/newfloat), which minted is set up to use.
+\usepackage{caption} % See <https://ctan.org/pkg/caption>.
+% \captionsetup{labelsep=colon}
 
@@
 % Source code listings with improved syntax highlighting
-\usepackage{minted}
+\usepackage[newfloat]{minted}
 \usemintedstyle{pastie}
@@
 \makeglossaries
-\loadglsentries{glossary}
+\loadglsentries{tex/glossary}
 
@@
 
-\input{intro}
-\input{motivation}
-\input{types}
-\input{basic-concepts}
+\input{tex/intro}
+\input{tex/motivation}
+\input{tex/types}
+\input{tex/basic-concepts}
 
@@
 \clearpage
-% \appendix
-\begin{appendices}
-   % See <https://tex.stackexchange.com/q/119513>.
-   \crefalias{section}{appsec}
-   \input{appendices}
+\begin{appendices} % <https://tex.stackexchange.com/q/49643>
+   \crefalias{section}{appsec} % See <https://tex.stackexchange.com/q/119513>.
+   \crefalias{subsection}{appsec}
+   \input{tex/cpu-info}
+   \input{tex/methodology}
+   \input{tex/data}
+   % \input{tex/no-prefetch}
 \end{appendices}

### 2017-05-26 tex/types.tex
Use varioref in addition to hyperref and cleveref
https://ctan.org/pkg/varioref
Minor update

@@
 Up to 32 KiB, each access takes almost exactly 3 cycles.%
-\footnote{The numerical results are shown in \cref{tab:access-times}.}
+\footnote{The numerical results are shown in \vref{tab:access-times}.}
 This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
@@
 \alts{causing, incurring} delays of up to 200 cycles~\cite[cf.][17]{drepper2007}.
+
+\Cref{sec:prefetch} will explain the reason for going through the trouble of conducting
+random accesses instead of just reading the array sequentially for this measurement.
+% We will see why this measurement was conducted with random instead of sequential
+% accesses in \cref{sec:prefetch}.
 

### 2017-05-28 tex/basic-concepts.tex
Add awesome plot, expand content, tweak stuff
Minor update
Revise, tweak, expand

@@
 %      -- Drepper, p. 16
+% > A cache line is the "unit" of data you transfer to a cache.
+%      -- http://www.cs.umd.edu/class/sum2003/cmsc311/Notes/Memory/introCache.html
 \emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
@@
 
-My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
-lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
+My E-450 CPU is no exception and both of its data caches have 64-byte cache lines.%
+\footnote{See \cref{app:cpuinfo}.}
 % \begin{minted}[gobble=3]{bash}
@@
 with an increment given at compile time as \texttt{STEP} and measures the processor time.
+\begin{center}
+   \inputminted[firstline=27]{c}{line-size/line-size.c}
+   \captionof{listing}{%
+      Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
+   \label{lst:line-size}
+\end{center}
 The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
@@
 
+\begin{comment}
    \begin{listing}
@@
    \end{listing}
+\end{comment}
 
@@
 
+% Talk about how the term cache line is commonly conflated to mean an appropriately sized
+% and *aligned* block of main memory that can be loaded into a cache line.  This is how it
+% works, right?  Or can we load blocks starting at arbitrary addresses into a cache line?
+% One could even imagine loading memory that isn't even contiguous.  Most sources don't
+% seem to clear any of this up.  See <https://stackoverflow.com/q/3928995> ("How do cache
+% lines work?")
+\begin{comment}
+   The term cache line is also used to refer to \alts{identically, appropriately} sized
+   blocks in main memory.
+   % that may be loaded into cache.
+   These blocks are fixed:
+   % I.e., each byte in main memory falls exactly into one cache line.
+   % The boundaries
+
+   Cache lines in main memory are fixed.
+   % Data is not loaded starting from arbitrary addresses, but only from addresses that
+   % are multiples of the cache line size.
+   Data blocks that are loaded don't start at arbitrary addresses, but at multiples of the
+   cache line size.
+\end{comment}
+% See <https://en.wikipedia.org/wiki/Partition_of_a_set>.  XXX: I feel like there's some
+% kind of conflation of ideas going on here.
+Both cache and main memory can be thought of as being partitioned%
+% \footnote{%
+%    In the set-theoretic sense
+% }
+\ (in the set-theoretic\x{al}\ sense) % https://en.wiktionary.org/wiki/set-theoretic
+into \alts[3]{cache line-sized blocks, blocks of that size, cache lines}.  \alts[2]{{That
+is, d}, D}ata is \alts{not, neither} \alts{read or written, loaded} starting from
+arbitrary main memory addresses,
+% ...nor is it loaded into arbitrarily aligned blocks in cache.
+but only from addresses that are multiples of the cache line size.
+% The starting address will by a multiple of the cache line size.
+
+% TODO: this probably has some implications about aligning data.
+
+% TODO: maybe talk about the 64-bit bus width and burst mode as well.  And maybe about how
+% a write requires a read if we only write part of a cache line and the optimization of
+% adding dummy writes (I think Andrei talked about this).
+%
+% [1]: https://stackoverflow.com/q/39182060 "Why isn't there a data bus which is as wide
+%      as the cache line size?"
+
 \subsection{Prefetching}
+\label{sec:prefetch}
 % http://ithare.com/c-for-games-performance-allocations-and-data-locality/

### 2017-05-28 tex/intro.tex
Revise, tweak, expand

@@
 A hardware cache is a \alts{comparatively, relatively} fast and small physical memory.  It
-stores a subset of the data present in slower, larger memory that is expected to be used
-again soon.  The purpose of this additional memory is to reduce the number of accesses to
-the underlying slower storage.
+stores a subset of the data present in slower, larger \alts{storage, memory} that is
+expected to be used again soon.  The purpose of this additional memory is to reduce the
+number of accesses to the underlying slower storage.
 
@@
 levels (most commonly 2 or 3)} of CPU caches that are found on almost all present-day
-CPUs.  They are designated L1 cache, L2 cache, and so on, with L1 being the fastest and
-smallest level.  The underlying storage for CPU caches is the main memory.
+CPUs.  They are designated \gls{l1} cache, \gls{l2} cache, and so on, with \gls{l1} being
+the fastest and smallest level.  The underlying storage for CPU caches is the main memory.
 

### 2017-05-28 tex/types.tex
Revise, tweak, expand

--- 2017-05-26 tex/types.tex
@@
 %      -- Drepper, p. 31
+% TODO: move this paragraph?
 The biggest target for optimizations is the data cache.  \say{[Instruction] cache is much
@@
 initialization is done.
+
+\begin{comment}
+   We will see why this measurement was conducted with random instead of sequential
+   accesses in \cref{sec:prefetch}.
+
+   \Cref{sec:prefetch} \alts{explains, will explain} the reason for \x{going through the
+   trouble of} \alts{using, conducting} random accesses instead of just reading the array
+   sequentially\x{for this measurement}.
+\end{comment}
+\alts{We use random accesses because,
+      The reason for using random accesses is that}
+the CPU will detect and optimize sequential access by a technique called
+\emph{prefetching} discussed in \cref{sec:prefetch}, which would prevent us from
+\alts{determining, assessing} access times.
 
@@
 \Cref{fig:access-times} shows the extra CPU cycles used by \cref{lst:access-times}
-\alts[2]{above, in addition to, compared to} the \mintinline{text}{BASELINE} version for
-different array sizes.
+\alts{in addition to, above, compared to, relative to} the \mintinline{text}{BASELINE}
+version for different array sizes.
 That is, only the cycles used by the main loop are \alts{given, counted}, not those for
@@
 Up to 32 KiB, each access takes almost exactly 3 cycles.%
-\footnote{The numerical results are shown in \cref{tab:access-times}.}
+\footnote{The numerical results are shown in \vref{tab:access-times}.}
 This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
 the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
-with other processes and the operating system, so some of our data gets evicted.  The
-first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
+with other processes and the operating system, so some of our data gets \emph{evicted}.
+The first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
 KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
-less \gls{l1d} hits and an \gls{l2} access time of around 25 cycles.
+less \gls{l1d} hits and an \gls{l2} access time \x{of} around 25 cycles.
 
-The values from 512 KiB to 128 MiB \alts{exhibit, follow} a similar pattern.  The relative
-increase when the array size matches that of the \gls{l2}, 512 KiB, is \alts{greater, more
-striking} than for the \gls{l1d} before; possibly because \gls{l2} is a unified cache that
-also holds instructions.  Eventually, more and more accesses go to main memory,
-\alts{causing, incurring} delays of up to 200 cycles~\cite[cf.][17]{drepper2007}.
+The values from 512 KiB (the size of the \gls{l2}) to 128 MiB \alts{exhibit, follow} a
+similar pattern.  The relative increase when the array size matches that of the \gls{l2}
+is \alts{greater, more striking} than for the \gls{l1d} before; possibly because \gls{l2}
+is a unified cache that also holds instructions.  Eventually, more and more accesses go to
+main memory, \alts{causing, incurring} delays of up to 200
+cycles~\cite[cf.][\pno~17, figure 3.4]{drepper2007}.
+
+% FOLDOC has an entry on 'working set'.
+\alts[2]{{Evidently,}, The data suggests that} keeping the \emph{working set} a process
+uses during a time interval small can yield dramatic performance improvements.
 

### 2017-05-28 tex/basic-concepts.tex
Split basic-concepts.tex subsections into files

--- 2017-05-21 tex/basic-concepts.tex
@@
 
-\subsection{Cache Line} % or Cache Block
-% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
-%
-% > On x86/x64, cache line is 64 bytes for many years now.
-%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
-% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
-%      -- Drepper, p. 15
-% > It is not possible for a cache to hold partial cache lines.
-%      -- Drepper, p. 16
-\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
-memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
-x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
-{scott-meyers-talk}.%
-\footnote{%
-% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
-% > cache 256 KB in size, with 128-byte cache blocks.
-%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
-% TODO: what about the block size of main memory?  Should it be the same?
-Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
-The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
-line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
-line}~\cite[p.~11]{pentium4}.}
-\begin{comment}
-   \multiplefootnoteseparator%
-   % See <https://tex.stackexchange.com/a/71015>.
-   \footnote{This can also be checked on the command line:
-   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
-\end{comment}
-% > It means that as soon as you've accessed any single byte in a cache line, all the
-% > other 63 bytes are already in L1 cache
-%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
-\alts{%
-   This means accessing a single uncached 32-bit integer entails loading another 60
-   adjacent bytes.,
-   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
-   well.},
-   {When, for example, accessing a single byte that isn't already cached, another 63
-   adjacent bytes will be loaded.}
-}
-% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.
-
-My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
-lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
-% \begin{minted}[gobble=3]{bash}
-%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
-%    64
-%    64
-% \end{minted}
-%stopzone
-% \footnote{%
-%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
-%    %stopzone
-%    \noindent\mintinline{bash}!64!\\
-%    \noindent\mintinline{bash}!64!
-% }
-We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
-with an increment given at compile time as \texttt{STEP} and measures the processor time.
-The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
-% Starting from a step size of 16, the time roughly halves every time the step size is
-% doubled.  For the first 4 step sizes however, it is almost constant.
-As expected, the time roughly halves whenever the step size is doubled --- but only from a
-step size of 16.  For the first 4 step sizes, it is almost constant.
-
-% The reason why the loops take the same amount of time has to do with memory.  The
-% running time of these loops is dominated by the memory accesses to the array, not by the
-% integer multiplications.
-%    -- https://igoro.com/archive/gallery-of-processor-cache-effects/
-This is because the run times are \alts{primarily due to, dominated by} memory accesses.
-Up to a step size of 8, every 64-byte line has to be loaded.  At 16, the values we modify
-are 128 bytes apart,%
-\footnote{16 \texttt{int64\_t} values of 8 bytes each}
-so every other cache line is skipped.  At 32, three out of four cache lines are skipped,
-and so on~\cite[cf.][example 2]{gallery}.
-
-\begin{listing}
-   \inputminted[firstline=27]{c}{line-size/line-size.c}
-   \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
-   \label{lst:line-size}
-\end{listing}
-
-\begin{figure}
-   \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={logarithmic,
-               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
-                                 8 as \textbf{8}, 16, 64, 256, 1024}},
-                      minor={at={32, 128, 512}}},
-               label={Step size}, length=0.8\textwidth},
-       y axis={logarithmic,
-               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
-               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
-               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
-               label={Processor time (ms)}, length=6cm},
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         % See <https://tex.stackexchange.com/q/198323>.
-         data [read from file=line-size/line-size.csv, separator=\space];
-   \end{tikzpicture}
-   \caption{Processor times for running \cref{lst:line-size}}
-   \label{fig:line-size}
-\end{figure}
-
-\subsection{Prefetching}
-% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
-% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
-% https://en.wikipedia.org/wiki/Cache_prefetching
-\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
-patterns (for instance sequential access to an array)... TODO
-
-\subsection{Associativity}
-% https://en.wikipedia.org/wiki/CPU_cache#Associativity
-
+\input{tex/basic-concepts/lines.tex}
+\input{tex/basic-concepts/prefetching.tex}
+\input{tex/basic-concepts/assoc.tex}
 \subsection{Cache Hits and Misses}
 % TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?
-
-\subsection{Spatial Locality}
-Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
-memory into the CPU cache) in \emph{blocks}.
-
-Moving down the memory hierarchy, access latencies increase faster than the
-\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
-many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
-memory is almost as fast as access to a single bit},~\cite[2]{afmh}.
-
-% TODO: move this.
-% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
-% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
-% summing the array.
-
-\subsection{Temporal Locality}
-TODO~\cite{drepper2007}.
-
+\input{tex/basic-concepts/spat_loc.tex}
+\input{tex/basic-concepts/temp_loc.tex}
 \subsection{Memory Access Pattern}

### 2017-05-28 tex/basic-concepts/lines.tex
Split basic-concepts.tex subsections into files

\subsection{Cache Line} % or Cache Block
% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
%
% > On x86/x64, cache line is 64 bytes for many years now.
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
%      -- Drepper, p. 15
% > It is not possible for a cache to hold partial cache lines.
%      -- Drepper, p. 16
% > A cache line is the "unit" of data you transfer to a cache.
%      -- http://www.cs.umd.edu/class/sum2003/cmsc311/Notes/Memory/introCache.html
\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
{scott-meyers-talk}.%
\footnote{%
% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
% > cache 256 KB in size, with 128-byte cache blocks.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
% TODO: what about the block size of main memory?  Should it be the same?
Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
line}~\cite[p.~11]{pentium4}.}
\begin{comment}
   \multiplefootnoteseparator%
   % See <https://tex.stackexchange.com/a/71015>.
   \footnote{This can also be checked on the command line:
   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
\end{comment}
% > It means that as soon as you've accessed any single byte in a cache line, all the
% > other 63 bytes are already in L1 cache
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
\alts{%
   This means accessing a single uncached 32-bit integer entails loading another 60
   adjacent bytes.,
   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
   well.},
   {When, for example, accessing a single byte that isn't already cached, another 63
   adjacent bytes will be loaded.}
}
% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.

My E-450 CPU is no exception and both of its data caches have 64-byte cache lines.%
\footnote{See \cref{app:cpuinfo}.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
\begin{center}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \captionof{listing}{%
      Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{center}
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

% The reason why the loops take the same amount of time has to do with memory.  The
% running time of these loops is dominated by the memory accesses to the array, not by the
% integer multiplications.
%    -- https://igoro.com/archive/gallery-of-processor-cache-effects/
This is because the run times are \alts{primarily due to, dominated by} memory accesses.
Up to a step size of 8, every 64-byte line has to be loaded.  At 16, the values we modify
are 128 bytes apart,%
\footnote{16 \texttt{int64\_t} values of 8 bytes each}
so every other cache line is skipped.  At 32, three out of four cache lines are skipped,
and so on~\cite[cf.][example 2]{gallery}.

\begin{comment}
   \begin{listing}
      \inputminted[firstline=27]{c}{line-size/line-size.c}
      \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
      \label{lst:line-size}
   \end{listing}
\end{comment}

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
                                 8 as \textbf{8}, 16, 64, 256, 1024}},
                      minor={at={32, 128, 512}}},
               label={Step size}, length=0.8\textwidth},
       y axis={logarithmic,
               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
               label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=line-size/line-size.csv, separator=\space];
   \end{tikzpicture}
   \caption{Processor times for running \cref{lst:line-size}}
   \label{fig:line-size}
\end{figure}

% Talk about how the term cache line is commonly conflated to mean an appropriately sized
% and *aligned* block of main memory that can be loaded into a cache line.  This is how it
% works, right?  Or can we load blocks starting at arbitrary addresses into a cache line?
% One could even imagine loading memory that isn't even contiguous.  Most sources don't
% seem to clear any of this up.  See <https://stackoverflow.com/q/3928995> ("How do cache
% lines work?")
\begin{comment}
   The term cache line is also used to refer to \alts{identically, appropriately} sized
   blocks in main memory.
   % that may be loaded into cache.
   These blocks are fixed:
   % I.e., each byte in main memory falls exactly into one cache line.
   % The boundaries

   Cache lines in main memory are fixed.
   % Data is not loaded starting from arbitrary addresses, but only from addresses that
   % are multiples of the cache line size.
   Data blocks that are loaded don't start at arbitrary addresses, but at multiples of the
   cache line size.
\end{comment}
% See <https://en.wikipedia.org/wiki/Partition_of_a_set>.  XXX: I feel like there's some
% kind of conflation of ideas going on here.
Both cache and main memory can be thought of as being partitioned%
% \footnote{%
%    In the set-theoretic sense
% }
\ (in the set-theoretic\x{al}\ sense) % https://en.wiktionary.org/wiki/set-theoretic
into \alts[3]{cache line-sized blocks, blocks of that size, cache lines}.  \alts[2]{{That
is, d}, D}ata is \alts{not, neither} \alts{read or written, loaded} starting from
arbitrary main memory addresses,
% ...nor is it loaded into arbitrarily aligned blocks in cache.
but only from addresses that are multiples of the cache line size.
% The starting address will by a multiple of the cache line size.

% TODO: this probably has some implications about aligning data.

% TODO: maybe talk about the 64-bit bus width and burst mode as well.  And maybe about how
% a write requires a read if we only write part of a cache line and the optimization of
% adding dummy writes (I think Andrei talked about this).
%
% [1]: https://stackoverflow.com/q/39182060 "Why isn't there a data bus which is as wide
%      as the cache line size?"

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-28 tex/basic-concepts/prefetching.tex
Split basic-concepts.tex subsections into files

\subsection{Prefetching}
\label{sec:prefetch}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching

\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-28 tex/basic-concepts/spat_loc.tex
Split basic-concepts.tex subsections into files

\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

Moving down the memory hierarchy, access latencies increase faster than the
\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
memory is almost as fast as access to a single bit},~\cite[2]{afmh}.

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
% summing the array.

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-30 tex/basic-concepts/lines.tex
Add new measurement; revise, expand, and tweak

@@
 \begin{center}
+% \begin{listing}
    \inputminted[firstline=27]{c}{line-size/line-size.c}
+   % \caption{%
    \captionof{listing}{%
@@
    \label{lst:line-size}
+% \end{listing}
 \end{center}
@@
 and so on~\cite[cf.][example 2]{gallery}.
-
-\begin{comment}
-   \begin{listing}
-      \inputminted[firstline=27]{c}{line-size/line-size.c}
-      \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
-      \label{lst:line-size}
-   \end{listing}
-\end{comment}
-
-\begin{figure}
-   \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={logarithmic,
-               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
-                                 8 as \textbf{8}, 16, 64, 256, 1024}},
-                      minor={at={32, 128, 512}}},
-               label={Step size}, length=0.8\textwidth},
-       y axis={logarithmic,
-               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
-               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
-               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
-               label={Processor time (ms)}, length=6cm},
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         % See <https://tex.stackexchange.com/q/198323>.
-         data [read from file=line-size/line-size.csv, separator=\space];
-   \end{tikzpicture}
-   \caption{Processor times for running \cref{lst:line-size}}
-   \label{fig:line-size}
-\end{figure}
 
@@
 
+% TODO.  Part of the memory address partially determines which cache line is used.  The
+% binary logarithm of the number of cache sets is the amount of bits required to identify
+% the cache set.  The specific line used from that set depends on the replacement policy.
+% A number of least significant (lowest-order, right-most) bits are ignored, since all
+% addresses that only differ in those go into the same cache line.  If the line size is
+% 64, log_2(64) = 6 bits are ignored.  See Drepper, page 15.
+
 % TODO: this probably has some implications about aligning data.
@@
 
+% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
+% interrupting the source code listing.
+\begin{figure}[hbp]
+   \centering
+   \begin{tikzpicture}
+      \datavisualization
+      [scientific axes=clean,
+       x axis={logarithmic,
+               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
+                                 8 as \textbf{8}, 16, 64, 256, 1024}},
+                      minor={at={32, 128, 512}}},
+               label={Step size}, length=0.8\textwidth},
+       y axis={logarithmic,
+               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
+               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
+               ticks={major={at={10, 20, 40, 80, 160, 320}}},
+               label={Processor time (ms)}, length=6cm},
+       visualize as scatter,
+       scatter={style={mark=*, mark options={scale=.65}}}]
+         % See <https://tex.stackexchange.com/q/198323>.
+         data [read from file=line-size/line-size.csv, separator=\space];
+   \end{tikzpicture}
+   \caption{Processor times for running \cref{lst:line-size}}
+   \label{fig:line-size}
+\end{figure}
+
 % vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-30 tex/basic-concepts/prefetching.tex
Add new measurement; revise, expand, and tweak

@@
 \label{sec:prefetch}
-% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
-% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
-% https://en.wikipedia.org/wiki/Cache_prefetching
 
-\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
-patterns (for instance sequential access to an array)... TODO
+% \cref{lst:TODO} is
+Consider
+a simplified version of \cref{lst:access-times} that, instead of using random accesses,
+simply walks over the array sequentially.  It still follows the pointers to do this, but
+the array is no longer shuffled.
+
+% \emph{Cache prefetching} is a technique by which CPUs recognize predictable access
+% patterns (for instance sequential access to an array)... TODO
+
+% [1]: http://ithare.com/c-for-games-performance-allocations-and-data-locality/
+% [2]: http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
+% [3]: https://en.wikipedia.org/wiki/Cache_prefetching
 

### 2017-05-30 tex/types.tex
Add new measurement; revise, expand, and tweak

@@
 
-\begin{center}
-   \begin{tabular}{ r | c c c c }
-             & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
-      Cycles & 3--4      & 10--12   & 30--70   & 100--150
-   \end{tabular}
-\end{center}
+% \begin{center}
+%    \begin{tabular}{ r | c c c c }
+%              & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
+%       Cycles & 3--4      & 10--12   & 30--70   & 100--150
+%    \end{tabular}
+% \end{center}
+\begin{center}\begin{tabular}{ r c c c c }
+   \toprule
+          & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\
+   \cmidrule[\lightrulewidth](l){2-5}
+   % \midrule
+   Cycles & 3--4      & 10--12   & 30--70   & 100--150 \\
+   \bottomrule
+\end{tabular}\end{center}
+
 %
@@
 KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
-less \gls{l1d} hits and an \gls{l2} access time \x{of} around 25 cycles.
+less \gls{l1d} \emph{hits} and an \gls{l2} access time \x{of} around 25 cycles.
 
@@
 % interrupting the source code listing.
-\begin{figure}[hbp]
+\begin{figure}%[hbp]
    \centering

### 2017-05-30 access-times/access-times.c

@@ -1,4 +1,4 @@
-// vim: set tw=72 sts=-1 sw=3 et cms=//%s
+// vim: tw=72 sts=-1 sw=3 et cms=//%s
 
 #include <stdint.h>  // int64_t, size_t
 #include <stdio.h>   // printf

### 2017-05-30 seq-access-times/access-times.c

@@ -1,0 +1,25 @@
+// vim: tw=72 sts=-1 sw=3 et cms=//%s
+
+#include <stdint.h>  // int64_t, size_t
+#include <stdio.h>   // printf
+#include <stdlib.h>  // rand
+
+#define N 100000000  // 100 million
+
+struct elem {
+   struct elem *next;
+} array[SIZE];
+
+int main() {
+   for (size_t i = 0; i < SIZE - 1; ++i) array[i].next = &array[i + 1];
+   array[SIZE - 1].next = array;
+#ifndef BASELINE
+   int64_t dummy = 0;
+   struct elem *i = array;
+   for (size_t n = 0; n < N; ++n) {
+      dummy += (int64_t)i;
+      i = i->next;
+   }
+   printf("%d\n", dummy);
+#endif
+}

### 2017-05-30 access-times/access-times-seq.c
Delete access-times/access-times-seq.c

@@ -1,16 +1,0 @@
-// vim: set tw=72 sts=-1 sw=3 et cms=//%s
-
-#include "stdint.h"  // int64_t
-#include "stdio.h"   // printf
-#include "stdlib.h"  // calloc
-
-#define N 1000000000
-
-int main() {
-   int64_t* array = (int64_t*)calloc(SIZE, sizeof(int64_t));
-   int sum;
-   for (size_t n = 0; n < N; ++n) {
-      sum += array[(8 * n) % SIZE];
-   }
-   printf("%d\n", sum);  // Better do this...
-}

### 2017-05-30 tex/basic-concepts/prefetching.tex
Add plot of access times for sequential reads

@@
 
+\begin{figure}
+   \centering
+   \begin{tikzpicture}
+      \datavisualization[%
+         scientific axes=clean,
+         x axis={%
+            logarithmic,
+            ticks={major={
+               /pgf/number format/int detect,
+               at={
+                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
+                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
+                  8388608 as \pgfmathprintnumber{8192},
+                  33554432 as \pgfmathprintnumber{32768},
+                  134217728 as \pgfmathprintnumber{131072},
+            }}},
+            grid={at={32768, 524288}},
+            label={Array Size (KiB)},
+            length=0.8\textwidth},
+         y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
+         visualize as scatter,
+         scatter={style={mark=*, mark options={scale=.65}}}]
+         data [read from file=seq-access-times/access-times.csv, separator=\space];
+   \end{tikzpicture}
+   \caption{Access Times for Sequential Reads}
+   \label{fig:seq-access-times}
+\end{figure}
+
 % [1]: http://ithare.com/c-for-games-performance-allocations-and-data-locality/

### 2017-05-30 tex/types.tex
Create style for the two access time scatter plots

@@
 
-% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
-% interrupting the source code listing.
-\begin{figure}%[hbp]
-   \centering
-   \begin{tikzpicture}
-      \datavisualization[%
+\tikzdatavisualizationset{
+   array size vs cycles plot/.style={
       scientific axes=clean,
-         x axis={%
+      x axis={
          logarithmic,
@@
       visualize as scatter,
-         scatter={style={mark=*, mark options={scale=.65}}}]
+      scatter={style={mark=*, mark options={scale=.65}}},
+   }
+}
+
+% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
+% interrupting the source code listing.
+\begin{figure}%[hbp]
+   \centering
+   \begin{tikzpicture}
+      \datavisualization[array size vs cycles plot]
          data [read from file=access-times/access-times.csv, separator=\space];

### 2017-05-30 paper.tex
Revise, tweak, expand
Add new measurement; revise, expand, and tweak
Exclude some stuff from the article

@@
 % Use a consitent group/thousands separator for siunitx and pgf.
-\pgfkeys{/pgf/number format/.cd,
-   set thousands separator=\,
+\pgfkeys{%
+   /pgf/number format/.cd,
+   1000 sep={\,},
+   min exponent for 1000 sep=4,
 }
@@
 
+% X (cross) something out.  See <https://tex.stackexchange.com/a/276698>.
+\newcommand{\x}[1]{\ignorespaces}
+
 % Top matter {{{1
@@
 \input{tex/motivation}
+% \section{Hardware Caches} % Cache, Structure of Caches
+\input{tex/operation}
 \input{tex/types}
-\input{tex/basic-concepts}
-
+\input{tex/key-concepts}
 \section{Example: \texttt{std::vector} vs. \texttt{std::list}}
@@
 extension of the \gls{ram} model.
-
-\begin{listing}
-   \inputminted[firstline=8]{c}{array-sum/array-sum.c}
-   \caption{This is C code}
-   \label{lst:array-sum}
-\end{listing}
-
-% This program is likely to have excellent spatial locality: the array...
-
-% \begin{figure}[htb]
-\begin{figure}
-   \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={label={Array size (KiB)}, length=0.8\textwidth},
-       y axis={label={Processor time (ms)}, length=6cm},
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         % See <https://tex.stackexchange.com/q/198323>.
-         data [read from file=array-sum/size-time.csv, separator=\space];
-   \end{tikzpicture}
-   \caption{This figure took way too long to create}
-   \label{fig:array-sum}
-\end{figure}
-
-% See \cref{fig:array-sum}.  What's going on here?
 

### 2017-05-30 tex/excluded.tex
Exclude some stuff from the article

\begin{listing}
   \inputminted[firstline=8]{c}{array-sum/array-sum.c}
   \caption{This is C code}
   \label{lst:array-sum}
\end{listing}

% This program is likely to have excellent spatial locality: the array...

% \begin{figure}[htb]
\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={label={Array size (KiB)}, length=0.8\textwidth},
       y axis={label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=array-sum/size-time.csv, separator=\space];
   \end{tikzpicture}
   \caption{This figure took way too long to create}
   \label{fig:array-sum}
\end{figure}

% See \cref{fig:array-sum}.  What's going on here?

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-05-31 tex/types.tex
Remove obsolete comment
Replace two tikzpicture environments with \tikz
I think these uses of \tikz are a shorter and semantically identical
alternative to using the tikzpicture environment.  See the second
paragraph of section 2.3 "Straight Path Construction" (page 31) of the
TikZ and PGF manual (version 3.0.1a).
Make KiB tick typesetting more automatic

--- 2017-05-30 tex/types.tex
@@
 
-% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
-% interrupting the source code listing.
-\begin{figure}%[hbp]
-   \centering
-   \begin{tikzpicture}
-      \datavisualization[%
+% Perform an integer division by 1024 on an argument in the format (scientific notation)
+% provided to the `tick typesetting` key.  Amazing.  This took me half a day to write.
+% See page 805 of the TikZ and PGF manual (version 3.0.1a).  We don't use
+% `\pgfmathprintnumberto` because its result is in math mode, e.g. `$42$`.
+% `pgfmathfloatparsenumber` allows arbitrary precision.
+\def\kibtypesetter#1{%
+   % \pgfmathprintnumberto[int trunc,1000 sep={}]{#1}{\a}
+   \pgfmathfloatparsenumber{#1}%
+   \pgfmathfloattoint{\pgfmathresult}%
+   \pgfmathsetmacro{\a}{\pgfmathresult}%
+   \newcount\b%
+   \b=\number\a%
+   \divide\b by 1024%
+   \pgfmathprintnumber{\the\b}%
+}
+
+\def\emphkibtypesetter#1{%
+   \ensuremath{\mathbf{\kibtypesetter{#1}}}%
+}
+
+% % This uses floating-point division.
+% \def\kibtypesetter#1{%
+%    \pgfmathparse{#1/1024}%
+%    % \pgfmathdivide{#1}{1024}%
+%    % \pgfmathdiv{#1}{1024}%
+%    \pgfmathprintnumber{\pgfmathresult}%
+% }
+
+\tikzdatavisualizationset{
+   array size vs cycles plot/.style={
       scientific axes=clean,
-         x axis={%
+      x axis={
          logarithmic,
-            ticks={major={
+         ticks={
             /pgf/number format/int detect,
+            major={
+               tick typesetter/.code=\kibtypesetter{####1},
                at={
-                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
-                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
-                  8388608 as \pgfmathprintnumber{8192},
-                  33554432 as \pgfmathprintnumber{32768},
-                  134217728 as \pgfmathprintnumber{131072},
-                  % $128\cdot 2^{10}$}}},
-                  % $2^{17}$}}},
-            }}},
+                  2048, 8192, 131072, 2097152, 8388608, 33554432, 134217728,
+                  32768 as \emphkibtypesetter{32768},
+                  524288 as \emphkibtypesetter{524288},
+                  % 32768 as [style={font=\bfseries}],
+                  % 32768 as \textbf{\kibtypesetter{32768}},
+                  % 32768 as \ensuremath{\mathbf{\kibtypesetter{32768}}},
+                  % 32768 as [tick typesetter/.code=\emphkibtypesetter],
+               },
+            },
+            % minor={
+               % style=black, tick text at low,
+               % tick typesetter/.code={\kibtypesetter{##1}},
+               % at={2048, 8192, 131072, 2097152, 8388608, 33554432, 134217728},
+            % },
+            minor at={4096, 16384, 65536, 262144, 1048576, 4194304, 16777216, 67108864},
+         },
          grid={at={32768, 524288}},
          label={Array Size (KiB)},
-            length=0.8\textwidth},
+         length=0.8\textwidth,
+      },
       y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
       visualize as scatter,
-         scatter={style={mark=*, mark options={scale=.65}}}]
+      scatter={style={mark=*, mark options={scale=.65}}},
+   },
+}
+
+\begin{figure}
+   \centering
+   \tikz \datavisualization[array size vs cycles plot]
       data [read from file=access-times/access-times.csv, separator=\space];
-   \end{tikzpicture}
    % \caption{Access Times for Random Reads (\Cref{lst:access-times})}

### 2017-05-31 tex/types.tex
Put datavisualization configuration into new file

--- 2017-05-30 tex/types.tex
@@
 
-% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
-% interrupting the source code listing.
-\begin{figure}%[hbp]
+\begin{figure}
    \centering
-   \begin{tikzpicture}
-      \datavisualization[%
-         scientific axes=clean,
-         x axis={%
-            logarithmic,
-            ticks={major={
-               /pgf/number format/int detect,
-               at={
-                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
-                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
-                  8388608 as \pgfmathprintnumber{8192},
-                  33554432 as \pgfmathprintnumber{32768},
-                  134217728 as \pgfmathprintnumber{131072},
-                  % $128\cdot 2^{10}$}}},
-                  % $2^{17}$}}},
-            }}},
-            grid={at={32768, 524288}},
-            label={Array Size (KiB)},
-            length=0.8\textwidth},
-         y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
-         visualize as scatter,
-         scatter={style={mark=*, mark options={scale=.65}}}]
+   \tikz \datavisualization[array size vs cycles plot]
       data [read from file=access-times/access-times.csv, separator=\space];
-   \end{tikzpicture}
    % \caption{Access Times for Random Reads (\Cref{lst:access-times})}

### 2017-06-02 line-size/line-size.c
Tweak line-size.c

@@ -1,6 +1,6 @@
-/* vim: set tw=72 sts=-1 sw=3 et cms=//%s: */
+// vim: tw=72 sts=-1 sw=3 et cms=//%s
 
-#include "stdint.h"  // int64_t
+#include "stdint.h"  // int64_t, size_t
 #include "stdio.h"   // printf
 #include "stdlib.h"  // calloc
 #include "time.h"    // clock

### 2017-06-02 seq-access-times/access-times.c
Add new measurement and full profiling results

@@ -11,15 +11,23 @@ struct elem {
 } array[SIZE];
 
 int main() {
+#ifndef STEP
    for (size_t i = 0; i < SIZE - 1; ++i) array[i].next = &array[i + 1];
    array[SIZE - 1].next = array;
+#else
+   for (size_t i = 0; i < SIZE - STEP; i += STEP)
+      array[i].next = array + i + STEP;
+   array[SIZE - STEP].next = array;
+#endif
 #ifndef BASELINE
    int64_t dummy = 0;
    struct elem *i = array;
    for (size_t n = 0; n < N; ++n) {
+      // dummy /= 73;
       dummy += (int64_t)i;
+      // dummy /= 8234;
       i = i->next;
    }
-   printf("%d\n", dummy);
+   printf("%ld\n", dummy);
 #endif
 }

### 2017-06-03 seq-access-times/access-times.c
Add new measurement

@@ -23,9 +23,13 @@ int main() {
    int64_t dummy = 0;
    struct elem *i = array;
    for (size_t n = 0; n < N; ++n) {
-      // dummy /= 73;
+#ifdef WORKWORK
+      dummy /= 73;
       dummy += (int64_t)i;
-      // dummy /= 8234;
+      dummy /= 8234;
+#else
+      dummy += (int64_t)i;
+#endif
       i = i->next;
    }
    printf("%ld\n", dummy);

### 2017-06-05 tex/basic-concepts/lines.tex
Swap "Prefetching" and "Cache Line" sections
Expand content, revise, tweak

@@
 Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
-The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
-line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
+The Intel Pentium 4 processor had an \gls{l1d} cache with \say{64 bytes per cache
+line}~\cite[p.~9]{pentium4} but an \gls{l2} cache with \say{128 bytes per cache
 line}~\cite[p.~11]{pentium4}.}
@@
        x axis={logarithmic,
-               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
-                                 8 as \textbf{8}, 16, 64, 256, 1024}},
-                      minor={at={32, 128, 512}}},
-               label={Step size}, length=0.8\textwidth},
+               ticks={major at={1, 2, 4, 8 as \textbf{8}, 16, 64, 256, 1024},
+                      minor at={32, 128, 512}},
+               grid={at=8},
+               label={Step Size}, length=0.8\textwidth},
        y axis={logarithmic,
@@
                % ticks={major={at={6, 12, 24, 48, 92, 184}}},
-               ticks={major={at={10, 20, 40, 80, 160, 320}}},
-               label={Processor time (ms)}, length=6cm},
+               ticks={major at={10, 20, 40, 80, 160, 320}},
+               grid=at ticks,
+               label={Processor Time (ms)}, length=6cm},
        visualize as scatter,
@@
    \end{tikzpicture}
-   \caption{Processor times for running \cref{lst:line-size}}
+   \caption{Processor Times for Running \Cref{lst:line-size}}
    \label{fig:line-size}

### 2017-06-05 tex/basic-concepts/prefetching.tex
Create style for the two access time scatter plots
Replace two tikzpicture environments with \tikz
I think these uses of \tikz are a shorter and semantically identical
alternative to using the tikzpicture environment.  See the second
paragraph of section 2.3 "Straight Path Construction" (page 31) of the
TikZ and PGF manual (version 3.0.1a).
Expand content, revise, tweak

--- 2017-05-30 tex/basic-concepts/prefetching.tex
@@
 
-% \cref{lst:TODO} is
-Consider
-a simplified version of \cref{lst:access-times} that, instead of using random accesses,
-simply walks over the array sequentially.  It still follows the pointers to do this, but
-the array is no longer shuffled.
+Consider a simplified version of \cref{lst:access-times} that, instead of using random
+accesses, simply walks over the array sequentially.  It still follows the pointers to do
+this, but the array is no longer shuffled.  The results of profiling this new program
+\alts{as, in the same way as, just as, like} \cref{lst:access-times} before are
+\alts{plotted, shown} in \cref{fig:seq-access-times}.%
+\footnote{%
+   \Vref{tab:seq-access-times} shows the numerical results.%
+   % Numerical results are shown in \vref{tab:seq-access-times}.%
+}
 
-% \emph{Cache prefetching} is a technique by which CPUs recognize predictable access
-% patterns (for instance sequential access to an array)... TODO
+% XXX: consider possible effects of [software prefetching][1].  I think GCC doesn't enable
+% this type of optimization unless `-fprefetch-loop-arrays` is explicitly specified; i.e.,
+% none of the `-O` levels enables it.  See [2].  XXX: WRONG:
+%
+%    $ gcc -O2 -Q --help=optimizers | grep prefetch
+%    -fprefetch-loop-arrays                [enabled]
+%
+% This is also interesting:
+%
+%    $ diff <(gcc -Q --help=optimizers) <(gcc -O2 -Q --help=optimizers)
+%
+% [1]: https://en.wikipedia.org/wiki/Cache_prefetching#Compiler_directed_prefetching
+% [2]: https://gcc.gnu.org/onlinedocs/gcc/Optimize-Options.html#Optimize-Options
+
+% XXX.  This section may be misleading: it could seem like it suggests all of the measured
+% speedup is a result of prefetching.  It is a huge contributor, though.  The result from
+% accessing all data in a cache line should at most be that of dividing the initial main
+% memory access (about 200 cycles) between 8 separate reads.  Since reads from L1d still
+% take about 3 cycles:
+%    (200 + 7 * 3) / 8 = 221 / 8 = 27.625
+% Since the actual numbers aren't much above 6 cycles, prefetching still has a huge effect
+% (6.3 / 27.625 ~= 22.8 %).
+\x{Compared to the nearly 200 cycles the random accesses caused, ...}
+Until the working set size \alts{matches that of, exceeds} the \gls{l1d}, the access times
+are virtually unchanged at 3 cycles, but exceeding the \gls{l1d} and hitting the \gls{l2}
+\alts{increases this by, adds} no more than \alts{a single, one} cycle.
+More \alts{strikingly, remarkably}, \alts{exceeding} the \gls{l2}
+\alts{%
+   has \alts{similarly limited, comparably little} effect,
+   is \alts{com:comparably, similarly} inconsequential,
+}.
+The access time plateaus not much above 6 cycles \x{now} --- about \alts{\SI{3}{\percent},
+3\%, 3 \%} of the maximum we saw for random reads.
+ % This is in large part thanks to \emph{prefetching}.
+Much of this can be explained by the improved use of cache lines: the penalty of loading a
+cache line is distributed among 8 accesses now.  This \alts{%
+   could at \alts{best, most} get us down to,
+   can not get us down to less than,
+}
+\SI{12.5}{\percent}.
+% \alts{In large part, To a high degree, To a great extend}, the improvements are due to
+% \emph{prefetching}.
+The missing improvements are due to \emph{prefetching}.
+
+Prefetching is a \x{heuristic} technique by which CPUs \alts{predict \x{certain},
+recognize predictable} access patterns and \alts{%
+   preemptively push cache lines up the memory hierarchy before the program needs them,
+   speculatively load data before the program needs it,
+}.
+% > This can work well only when the memory access is predictable, though.
+%      -- Drepper, p. 23
+% > Currently prefetch units do not recognize non-linear access patterns.
+%      -- Drepper, p. 60
+\alts{%
+   {This can not work unless cache line access is predictable, though, which basically
+   means \x{sequential} linear},
+   {For this to work, cache line access has to be predictable, which usually means
+   sequential},
+   % This requires,
+   % This only works if,
+}%
+~\cite[60]{drepper2007}.%
+\footnote{%
+   As an example, the most complicated \emph{stride pattern} my laptop's CPU can detect is
+   % https://en.wikipedia.org/wiki/Hyphen#Suspended_hyphens
+   one that skips over at most 3 cache lines (for- or backwards) and may alternate strides
+   (e.g.  +1, +2, +1, +2, \ldots)~\cite[278]{14h}.
+}
+
+% > The purpose of prefetching is to hide the latency of a memory access.
+% ...
+% > Prefetching has one big weakness: it cannot cross page boundaries.
+% ...
+% > [R]egardless of how good the prefetcher is at predicting the pattern, the program will
+% > experience cache misses at page boundaries
+%      -- Drepper, p. 60
+
+% > Prediction or explicit prefetching might also guess where future reads will come from
+% > and make requests ahead of time; if done correctly the latency is bypassed altogether.
+%      -- https://en.wikipedia.org/wiki/Cache_(computing)#Latency
+
+% > [P]refetching [can] remove some of the costs of accessing main memory since it happens
+% > asynchronously with respect to the execution of the program. It can [...] make the
+% > cache appear bigger than it actually is.
+%      -- Drepper, p. 14
+Prefetching happens asynchronously to normal program execution~\cite[14]{drepper2007}
+% > [T]he processor is able to hide most of the main memory and even L2 access latency by
+% > prefetching cache lines into L2 and L1d.
+%      -- Drepper, p. 23
+and can therefore\x{, in principle,}\ almost completely hide the main memory latency%
+~\cite[23]{drepper2007}.
+This is not quite what we observe in \cref{fig:seq-access-times} because the CPU
+\alts{performs, has to perform} little enough work for memory bandwidth to become the
+bottleneck.
+% XXX: the peak transfer rate of my ThinkPad's memory (DDR3-1333, I think) is much higher
+% (should be 10666.67 MB/s) than the rate observed in this test (about 2 GB/s).
+%    8 B / (6.2 / 1.65 GHz) = 8 * 1.65 GB/s / 6.2 = 105.6 ~= 2.13 GB/s
+% What limits it?  How to achieve the theoretical maximum?  My understanding is that
+% sequential read access like this should pretty much be the most efficient use of RAM
+% possible.
+%
+% I compiled and ran the [STREAM][1] benchmark ([FAQ][2]) by [Dr. John D. McCalpin][3]
+% recommended [here][4].  It gives similarly low data rates.
+%
+% [1]: https://www.cs.virginia.edu/stream/
+% [2]: https://www.cs.virginia.edu/stream/ref.html
+% [3]: http://www.cs.virginia.edu/~mccalpin/
+% [4]: http://www.admin-magazine.com/HPC/Articles/Finding-Memory-Bottlenecks-with-Stream
+% [5]: https://software.intel.com/en-us/articles/optimizing-memory-bandwidth-on-stream-triad
+% [6]: https://www.nersc.gov/users/computational-systems/cori/nersc-8-procurement/trinity-nersc-8-rfp/nersc-8-trinity-benchmarks/stream/
+% [7]: https://en.wikipedia.org/wiki/Memory_bandwidth
+%
+% TODO: I also tried [bandwidth](http://zsmith.co/bandwidth.html).
+Adding some expensive operations like integer divisions every loop iteration changes that
+and \alts{effectively, almost completely} levels the cycles spend per iteration across all
+working set sizes.%
+% I tested this.  The difference between L1d and L2 virtually disappears (~0.01 cycles)
+% and exceeding the L2 increases the time per element by a single cycle.
+\footnote{%
+   % TODO.
+   See \vref{fig:seq-access-cpu-bound}.
+}
+
+% Ubiquitous.
+
+% > Hardware based prefetching is typically accomplished by having a dedicated hardware
+% > mechanism in the processor that watches the stream of instructions or data being
+% > requested by the executing program, recognizes the next few elements that the program
+% > might need based on this stream and prefetches into the processor's cache.
+%      -- https://en.wikipedia.org/wiki/Cache_prefetching#Types_of_cache_prefetching
+\alts{What I described so far is, So far I described} \emph{hardware} prefetching.  It
+uses dedicated silicon to automatically detect access patterns.  There is also
+\emph{software} prefetching, which is triggered by special machine instructions that may
+be inserted by the compiler or manually by the programmer.  Software prefetching is
+discussed in~\cite{drepper2007}.
+
+% > The idea of [the _mm_prefetch() intrinsic] function (actually an asm instruction from
+% > x86/x64 instruction set) is to inform CPU that you're about to need certain memory
+% > location.
+%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
+
+% https://gcc.gnu.org/onlinedocs/gcc/Optimize-Options.html#index-fprefetch-loop-arrays
+
+% > Most of the time, prefetch just silently works behind the scenes, and I didn't see
+% > cases when messing with prefetch at application-level would be worth the trouble.  At
+% > least in theory, however, such cases do exist.
+%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
+
+% > The source for the prefetch operation is usually main memory.
+%      -- https://en.wikipedia.org/wiki/Cache_prefetching
+
+\begin{figure}
+   \centering
+   \tikz \datavisualization[array size vs cycles plot]
+      data [read from file=seq-access-times/access-times.csv, separator=\space];
+   \caption{Access Times for Sequential Reads}
+   \label{fig:seq-access-times}
+\end{figure}
+
+% \begin{figure}
+%    \centering
+%    \tikz \datavisualization[array size vs cycles plot]
+%       data [read from file=seq-access-times/step8/access-times.csv, separator=\space];
+%    \caption{TODO}
+%    \label{fig:seq8-access-times}
+% \end{figure}
 

### 2017-06-05 tex/methodology.tex
Use varioref in addition to hyperref and cleveref
https://ctan.org/pkg/varioref
Expand content, revise, tweak

@@
+% Performance Analysis; Run Time Analysis; Timing Methodology; Profiling Pitfalls; Run
+% Time Measuring; Benchmarking Methodology; Performance Monitoring
+\section{Profiling Methodology}
+\label{app:meth}
+
 % Programs: `time`, bad; `\time -v`, better; oprofile(1); gprof(1); perf; Valgrind's
@@
 
-% Performance Analysis; Run Time Analysis; Timing Methodology; Profiling Pitfalls; Run
-% Time Measuring; Benchmarking Methodology; Performance Monitoring
-\section{[TODO] Profiling Methodology}
-\label{app:meth}
-
-Plenty of things can go wrong when
-profiling.
-% measuring \alts{execution, run} times.
-These are the ones I was aware of and tried to account for while doing measurements for
-this \article.
+\begin{comment}
+Plenty of things can go wrong when \alts{profiling, measuring \alts{execution, run}
+times}.  These are the ones I was aware of and tried to account for while doing
+measurements for this \article.
 \begin{itemize}
@@
       at least workload and temperature.  This should be disabled when measuring execution
-      times.
+      times.  I used the cpupower(1) program.
+      % See <https://wiki.archlinux.org/index.php/CPU_frequency_scaling> and
+      % <https://github.com/torvalds/linux/tree/master/tools/power/cpupower>.
       \begin{minted}[autogobble]{text}
-         # TODO
+         # cpupower frequency-set -g performance
+         $ ./benchmark
+         # cpupower frequency-set -g ondemand
       \end{minted}
+      %stopzone
       % See [5].  Not a problem when measuring cycles.
@@
 % See [7].
+\end{comment}
+
+All programs were compiled and run on Linux using the \gls{gcc} with
+\mintinline{text}{-O2} and \mintinline{text}{-march=native}.
+% Additionally,
+I disabled CPU frequency scaling with \mintinline{text}{cpupower(1)},
+% See <https://wiki.archlinux.org/index.php/CPU_frequency_scaling>.
+reduced the number of running processes,%
+\footnote{%
+   I stopped Firefox (which added visible noise) and the X server, for example.
+}
+% (e.g. leaving Firefox open in the background added lots of noise),
+and assigned a high priority to the benchmark process with the \mintinline{text}{chrt(1)}
+command.
+
+The \emph{\href{https://github.com/meribold/cache-seminar-paper/blob/master/makefile}
+{makefile}} used to compile and run the measurements as well as the full source code of
+the programs is available \alts{here, on GitHub}:
+\url{https://github.com/meribold/cache-seminar-paper}.
 
@@
 
-Frequency scaling isn't a problem here.  I used the \mintinline{text}{ocount(1)} event
-counting tool added to the OProfile (TODO: cite) project in version 0.9.9.  [TODO] This
-uses \emph{hardware performance counters}.
-% `perf(1)` seems to be a newer alternative to OProfile [8].
+% Frequency scaling is still a problem here.  For example, when the program is memory
+% bound, it may take fewer cycles at a lower frequency while completing in the same wall
+% time.  I.e., not everything is proportional to CPU frequency.
+
+% > OProfile leverages the hardware performance counters of the CPU.
+%      -- http://oprofile.sourceforge.net/about/
+
+% > `ocount` is an OProfile tool that can be used to count native hardware events
+% > occurring in [...] a specific application.
+%      -- http://oprofile.sourceforge.net/doc/getting-started-with-ocount.html
+
+% A list of AMD family 14h CPU performance counters is available at [1].  `ophelp(1)` can
+% also be used.
+% [1]: http://oprofile.sourceforge.net/docs/amd-family14h-events.php
+
+I used the \mintinline{text}{ocount(1)} event counting tool added to the
+\citetitle{oprofile}~\cite{oprofile} project in version 0.9.9.
+% `perf(1)` seems to be a newer alternative to OProfile [8].  TODO: use it?
+%
+% > Newer CPU models support a hardware performance counter mode which uses hardware logic
+% > to record events without any active code needed
+%      -- https://en.wikipedia.org/wiki/OProfile
+It uses \emph{hardware performance counters}%
+\footnote{%
+   \url{https://en.wikipedia.org/wiki/Hardware_performance_counter}
+}
+to provide low-overhead performance information without source code modifications.
+\alts{%
+   {Essentially, this is the procedure:},
+   The \x{exact} process is more or less
+}
+
+% diff <(gcc -march=native -E -v - </dev/null 2>&1) <(gcc -E -v - </dev/null 2>&1)
+% gcc -O2 -fno-prefetch-loop-arrays -Q --help=optimizers
 
 \begin{minted}[autogobble]{text}
-   ophelp
-   sudo chrt -f 99 ocount -e CPU_CLK_UNHALTED program
+   $ gcc -march=native -O2 program.c
+   # cpupower frequency-set -g ondemand
+   # chrt -f 99 ocount -e CPU_CLK_UNHALTED ./a.out
 \end{minted}
+%stopzone
 
-TODO~\cite[78\psqq]{drepper2007}.
+% ~\cite[78\psqq]{drepper2007}.
 
@@
 % [8]: http://oprofile-list.sf.narkive.com/rVbJlHdX/oprofile-vs-perf
-% [9]:  http://oprofile.sourceforge.net/about/
-% [10]: https://en.wikipedia.org/wiki/Hardware_performance_counter
-% [11]: http://oprofile.sourceforge.net/doc/index.html
+% [9]: https://serverfault.com/q/161008
+%      "What is the difference between renice and chrt commands in Linux?"
 

### 2017-06-05 tex/basic-concepts/lines.tex
Rerun measurements that are included in the paper
`pstree.out` contains the output of `pstree` just before running the
measurements.  I.e., it shows how many and which other processes were
running during the measurements that could potentially add noise.

@@
                % ticks={major={at={6, 12, 24, 48, 92, 184}}},
-               ticks={major at={10, 20, 40, 80, 160, 320}},
+               ticks={major at={5, 10, 20, 40, 80, 160, 320}},
                grid=at ticks,

### 2017-06-06 tex/methodology.tex
Draft section on locality, tweak stuff

@@
          $ ./benchmark
-         # cpupower frequency-set -g ondemand
+         # cpupower frequency-set -g schedutil
       \end{minted}
@@
    $ gcc -march=native -O2 program.c
-   # cpupower frequency-set -g ondemand
+   # cpupower frequency-set -g schedutil
    # chrt -f 99 ocount -e CPU_CLK_UNHALTED ./a.out

### 2017-06-06 tex/methodology.tex
Tweak, fix typo

@@
    $ gcc -march=native -O2 program.c
-   # cpupower frequency-set -g schedutil
+   # cpupower frequency-set -g performance
    # chrt -f 99 ocount -e CPU_CLK_UNHALTED ./a.out

### 2017-06-06 tex/motivation.tex
Revise, tweak, expand
Tweak, fix typo

--- 2017-05-21 motivation.tex
@@
 performance by two orders of magnitude because of different \emph{memory access
-patterns} (\cref{sec:map})~\cite{bigos}.  We will see an example of this in
+patterns}\x{ (\cref{sec:map})}~\cite{bigos}.  We will see an example of this in
 \cref{sec:vvl}.
 
-In a nutshell, hardware caches are ubiquitous but the performance improvements they
-provide are conditional.
+In a nutshell, hardware caches are ubiquitous but the performance \alts{gains,
+improvements} they provide are conditional.
 \begin{comment}

### 2017-06-07 tex/list-vs-vector.tex
Add program comparing std::list and std::vector

\section{Example: \texttt{std::vector} vs. \texttt{std::list}}

\alts{%
   The C++ program shown in \cref{lst:ithare},
   The following C++ program,
}
\x{, adapted from \textcite{bigos},} initializes a number of \acrshort{stl} containers
with random numbers and measures the processor time needed to \alts{sum all of them,
compute the sum}.
%
I first ran it with \mintinline{cpp}{Container} being a type alias for
\mintinline{cpp}{std::list}, then for \mintinline{cpp}{std::vector}.%
\footnote{%
   Each compiled with \x{the \gls{gcc}} and \mintinline{text}{-O3} and
   \mintinline{text}{-march=native}
}

\begin{center}
   \inputminted[firstline=20]{cpp}{ithare/list-vs-vector.cpp}
   \captionof{listing}{Compute the Sum of a Container, adapted from \textcite{bigos}}
   \label{lst:ithare}
\end{center}

\newread\file
\openin\file=ithare/speedup.txt
\read\file to \speedup
\closein\file

% This doesn't work: `\num{\input{ithare/speedup.txt}}`.

When using \mintinline{cpp}{std::vector}, computing the sum is
\num[round-mode=places, round-precision=0]{\speedup} times faster.

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-07 ithare/list-vs-vector.cpp

@@ -1,0 +1,42 @@
+// vim: tw=72 sts=-1 sw=3 et
+
+#include <cassert>
+#include <iostream>
+
+#ifdef VECTOR
+#include <vector>
+#else
+#include <list>
+#endif
+
+using namespace std;
+
+#ifdef VECTOR
+using Container = vector<int>;
+#else
+using Container = list<int>;
+#endif
+
+constexpr int N = 5000;
+
+int main() {
+   Container containers[N];
+   srand(time(0));
+   // Append an average of 5000 random values to each container.
+   for (int i = 0; i < N * 5000; ++i) {
+      containers[rand() % N].push_back(rand());
+   }
+
+   int sum = 0;
+   clock_t t0 = clock();
+   for (int m = 0; m < N; ++m) {
+      for (int num : containers[m]) {
+         sum += num;
+      }
+   }
+   clock_t t1 = clock();
+
+   // Print the sum so the loop doesn't get optimized out.
+   cout << sum << '\n';
+   cout << (t1 - t0) << '\n';
+}

### 2017-06-08 tex/list-vs-vector.tex
Expand, revise, tweak

@@
    The following C++ program,
-}
-\x{, adapted from \textcite{bigos},} initializes a number of \acrshort{stl} containers
+}%
+, adapted from \textcite{bigos}, initializes a number of \acrshort{stl} containers
 with random numbers and measures the processor time needed to \alts{sum all of them,
@@
 }
+Either way, the asymptotic complexity is \(\Theta(N)\).
 
 \begin{center}
-   \inputminted[firstline=20]{cpp}{ithare/list-vs-vector.cpp}
+   \inputminted[firstline=29]{cpp}{ithare/list-vs-vector.cpp}
    \captionof{listing}{Compute the Sum of a Container, adapted from \textcite{bigos}}
@@
 
-When using \mintinline{cpp}{std::vector}, computing the sum is
-\num[round-mode=places, round-precision=0]{\speedup} times faster.
+% > In trying to explain this difference, we'll need to get into educated-guesswork area.
+% > For MSVC and gcc, the performance difference between vector<> and list<> is pretty
+% > much in line with the difference between typical cached access times (single-digit
+% > clocks) and typical uncached access times (100-150 clocks).  As access patterns for
+% > vector<> are expected to use CPU prefetch fully and list<> under the patterns in
+% > Listing 1 is pretty much about random access to memory, which cannot be cached due to
+% > the size, this 100-150× difference in access times can be expected to translate into
+% > 100-150× difference in performance.
+%      -- https://accu.org/index.php/journals/2268
+
+% > [T]he performance difference between vector<> and list<> is pretty
+% > much in line with the difference between typical cached [...] and typical uncached
+% > access times.
+%      -- https://accu.org/index.php/journals/2268
+
+% \noindent
+% \alts{{The result is that, when}, When}  computing the
+% sum \alts{runs, is} \num[round-mode=places, round-precision=0]{\speedup} times faster.
+\alts{My, The} result is that computing the sum \alts{runs, is} \num[round-mode=places,
+round-precision=0]{\speedup} times faster when using \mintinline{cpp}{std::vector}.  Some
+of this difference can certainly be attributed to space overhead of the linked list and
+the added indirection, but the more \emph{cache-friendly} memory access pattern of
+\mintinline{cpp}{std::vector} is \alts{decisive, key, crucial, vital}~\cite[5\psq]{bigos}.
+Using \mintinline{cpp}{std::list} incurs \say{random access to memory} in this
+example~\cite[6]{bigos}.
 

### 2017-06-08 tex/methodology.tex
Expand, revise, tweak

@@
 All programs were compiled and run on Linux using the \gls{gcc} with
-\mintinline{text}{-O2} and \mintinline{text}{-march=native}.
+\mintinline{text}{-O2} (unless noted otherwise) and \mintinline{text}{-march=native}.
 % Additionally,
@@
 {makefile}} used to compile and run the measurements as well as the full source code of
-the programs is available \alts{here, on GitHub}:
+the programs is available \alts{at, here:, on GitHub:}
 \url{https://github.com/meribold/cache-seminar-paper}.

### 2017-06-08 ithare/list-vs-vector.cpp

@@ -1,42 +1,50 @@
 // vim: tw=72 sts=-1 sw=3 et
 
-#include <cassert>
+#include <cstdlib>  // std::rand, std::srand
+#include <ctime>    // std::clock, std::time, std::clock_t
 #include <iostream>
 
 #ifdef VECTOR
 #include <vector>
 #else
-#include <list>
+#include <list>  // Or <forward_list>?
 #endif
-
-using namespace std;
 
 #ifdef VECTOR
-using Container = vector<int>;
+using Container = std::vector<int>;
 #else
-using Container = list<int>;
+using Container = std::list<int>;
 #endif
+
+// When I try to lay the list out sequentially in memory, the difference
+// changes to a factor of about 11.
+//
+//    for (int n = 0; n < N; ++n) {
+//       containers[n].resize(5000, std::rand());
+//       for (int i = 0; i < 5000; ++i) {
+//          containers[n].push_back(std::rand());
+//       }
+//    }
 
 constexpr int N = 5000;
 
 int main() {
    Container containers[N];
-   srand(time(0));
+   std::srand(std::time(nullptr));
    // Append an average of 5000 random values to each container.
    for (int i = 0; i < N * 5000; ++i) {
-      containers[rand() % N].push_back(rand());
+      containers[std::rand() % N].push_back(std::rand());
    }
 
    int sum = 0;
-   clock_t t0 = clock();
+   std::clock_t t0 = std::clock();
    for (int m = 0; m < N; ++m) {
       for (int num : containers[m]) {
          sum += num;
       }
    }
-   clock_t t1 = clock();
+   std::clock_t t1 = std::clock();
 
-   // Print the sum so the loop doesn't get optimized out.
-   cout << sum << '\n';
-   cout << (t1 - t0) << '\n';
+   // Also print the sum so the loop doesn't get optimized out.
+   std::cout << sum << '\n' << (t1 - t0) << '\n';
 }

### 2017-06-13 paper.tex
Put datavisualization configuration into new file
Expand content, revise, tweak
Draft section on locality, tweak stuff
Tweak, fix typo
Add program comparing std::list and std::vector
Write a bit about the EMM; revise and tweak stuff

@@
 \usepackage{dirtytalk}      % ...
+
+% `$\nicefrac{a}{b}$` looks better than `$a/b$` in my opinion.  I also think it looks
+% better than `$\sfrac{a}{b}` from the more recent [xfrac][5] package.  See [3], [4], and
+% [5].
+%
+% [1]: https://www.ctan.org/pkg/nicefrac
+% [2]: https://www.ctan.org/pkg/xfrac
+% [3]: https://tex.stackexchange.com/q/3372
+%      "How do I typeset arbitrary fractions like the standard symbol for .5 = ½?"
+% [4]: https://tex.stackexchange.com/q/128496 "Elegant fractions in one line"
+% [5]: https://tex.stackexchange.com/q/43119 "Improved kerning in fractions?"
+\usepackage{nicefrac}
 
@@
 % [4]: https://tex.stackexchange.com/q/9203
-\usepackage{pgfplotstable, siunitx}
+\usepackage{pgfplotstable}
+\usepackage[binary-units]{siunitx} % Load `\kibi` and `\mebi` etc. (`binary-units`).
 % These packages are recommended by [pgfplotstable].  The booktabs package adds the
@@
 % [5]: https://en.wikibooks.org/wiki/LaTeX/Tables#Professional_tables
+
+% This gets loaded by the `pgfplotstable` sub-package as far as I know.
+%
+% \usepackage{pgfplots}
+%
+% It then suggests:
+%
+%    Package pgfplots Warning: running in backwards compatibility mode (unsuitable t
+%    ick labels; missing features). Consider writing \pgfplotsset{compat=1.14} into
+%    your preamble
+%
+% So I guess I'll go that.
+\pgfplotsset{compat=1.14}
 
@@
 % See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
-\usepackage{enumitem}
+\usepackage[inline]{enumitem}
 \setlist{noitemsep}
+
+% \usepackage{tasks}
 
@@
 \usepackage{tikz}
-\usetikzlibrary{datavisualization}
-% \usetikzlibrary{datavisualization.formats.functions}
+\input{tex/datavis.tex}
 
@@
    % What?
-   Typical present-day CPUs have two or more levels of caches.  This \article{} presents
-   the basic techniques used to optimize program performance based on knowledge about how
-   these hardware caches function.
+   Typical present-day CPUs have two or more levels of caches.
+   \alts[3]{
+      {This \article{} presents the basic techniques used to optimize program performance
+      based on knowledge about how these hardware caches function.},
+      {This \article{} explains their key architectural properties with the intend of
+      \alts{providing, giving} basic insight on possible program optimizations.},
+      {This \article{} provides basic insight into their operation and presents key
+      architectural properties which suggest possible program optimizations.},
+   }
 
@@
 \input{tex/key-concepts}
-\section{Example: \texttt{std::vector} vs. \texttt{std::list}}
-\label{sec:vvl}
-
+\input{tex/list-vs-vector.tex}
+\label{sec:lvv}
 % \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
 % https://en.wikipedia.org/wiki/Cache_replacement_policies
-
-\section{External Memory Model}
-The \gls{emm} is a widely used
-% TODO: is it?
-extension of the \gls{ram} model.
-
-\section{Cache-Oblivious Algorithms} % Cache-Oblivious Model
-
+\input{tex/emm.tex}
 \section{Parallel Computing}
-
 \subsection{Cache Coherence} % Coherence or coherency?
@@
 % Dirty.  Invalid.
-
 \subsection{False Sharing}

### 2017-06-13 tex/com.tex
Write a bit about the EMM; revise and tweak stuff

\subsection{Cache-Oblivious Model} % Cache-Oblivious Algorithms

% The \gls{com} is like the \gls{emm}, but we pretend not to know the values of B and M.
% This seems a bit silly.

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-13 tex/emm.tex
Write a bit about the EMM; revise and tweak stuff

\section{Abstract?}
% \section{Analysis of Algorithms}

% > However the efficiencies of any two "reasonable" implementations of a given algorithm
% > are related by a constant multiplicative factor called a *hidden constant*.
%      -- https://en.wikipedia.org/wiki/Analysis_of_algorithms

% Mathematical analysis of algorithms to obtain asymptotic upper bounds is nice.  Ignoring
% the non-uniform nature of memory hierarchies can yield pretty bad results, though.  What
% can be done about this?

We have seen that
\alts{
  {the hidden constant \alts{separating, distinguishing} the time complexities of two
  reasonable algorithms under asymptotic analysis can get quite big in the presence of a
  memory hierarchy.},
  {analyzing \alts{an algorithm's,} \alts{time, asymptotic, algorithmic} complexity,
  % This is not true: it doesn't matter for asymptotic complexity if there are different
  % types of memory.
  \x{under the usual \gls{ram} model, where a single uniform memory is \alts{presumed,
  assumed, supposed, premised},}
  can be quite inaccurate in the presence of a memory hierarchy.},
}
\alts{
  To \alts{escape, avert, prevent} having to rely \alts{only, entirely, exclusively} on
  empirical results, To overcome this problem,
},
abstract machine models taking the non-uniform \alts{memories of, nature of memory in}
real-world computers into account \alts{can be used, are used, have been proposed}.
% By whom?
One of these is the \gls{emm}\alts{
  ,
  ~\cite[5]{afmh},
  {, used \alts{in~\cite{afmh}, by \textcite{afmh}}.},
}.

\subsection{External Memory Model}

% https://en.wikipedia.org/wiki/Random-access_machine

% TODO: go into more detail about the RAM model?
\alts[2]{
  {\alts{Establishing the basis of the, The} \gls{emm}, the widely used \gls{ram} model
  assumes \say{a \say{sufficiently} large uniform memory}~\cite[5]{afmh} with a constant
  access time.
  The},
  {The \gls{emm} is an extension of the \gls{ram} model.  While the latter assumes \say{a
  \say{sufficiently} large uniform memory}~\cite[5]{afmh} with a constant access time,
  the},
}
\gls{emm} \alts{divides, \alts{expands upon, extends} this by dividing} the memory into
\emph{internal} and \emph{external}.
\alts{The, Only the} internal memory is accessed directly, but its size is limited to
\(M\) \alts{items, words}.
The external memory is unbounded, but \alts{
  can only be accessed indirectly by loading data,
  data \x{is not accessed directly and} has to be loaded,
}
into internal memory \say{using \glslink{io}{I/Os} that move \(B\) contiguous
\alts{[items], words}}~\cite[5]{afmh}.

% > Although the word "I/O" suggests that external memory should be identified with disk
% > memory, we are free to choose any two levels of the memory hierarchy for internal and
% > external memory in the model.
%        -- Algorithms for Memory Hierarchies, page 6
The use of the term \emph{\acrshort{io}} here is somewhat non-standard.  While it suggests
\x{\alts{that, the}} external memory represents an \gls{hdd}, \gls{ssd} or similar, it is
not constrained which \alts{physical storages, members of an actual memory hierarchy} are
\alts[2]{represented by, associated \alts{with, to}} \x{the} internal and external memory.
% Which members of an actual memory hierarchy are \alts{represented by, associated to} the
% internal and external memory is \alts{not constrained, as desired}.
If we choose the set of all CPU caches and the main memory, \(B\) becomes the cache line
size and \(M\) may be in the order of a few \si{\mebi\byte}.

% > The relative miss penalty is much lower for caches or the TLB than for disks, so
% > constant factors are important, we find that asymptotic analysis is not enough to
% > determine the performance of algorithms.
%      -- Algorithms for Memory Hierarchies, page 181

% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
% > of computation, hence the design of the algorithm is usually motivated by the need to
% > minimise the number of I/Os.
%      -- Algorithms for Memory Hierarchies, page 188

% \subsubsection{Caveats}

The number of \acrshortpl{io} an algorithm requires in the \gls{emm} can augment the
information provided by standard asymptotic complexity analysis%
\x{, if \acrshort{io} is pivotal to an algorithm's runtime},
but \alts[2]{can't substitute, is no substitute for} measurements.
For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
input\x{, like in \cref{lst:ithare},}\ is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
linked list in \cref{lst:ithare} probably takes almost \(N\) \acrshortpl{io}, since
consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
the maximum predicted performance difference between \mintinline{cpp}{std::vector} and
\mintinline{cpp}{std::list} is \(B\), the number of items a cache line can hold, which is
\si{16} in this case.%
\footnote{%
  % TODO: elaborate on the size of `int`?
  A cache line has \alts{\si{64}, \SI{64}{\byte}} and an \mintinline{text}{int} \si{4}
  bytes on the machine used.
}
Recalling that the actual performance difference was \num[round-mode=places,
round-precision=0]{\speedup}, this is pretty inaccurate, but more informative than saying
both algorithms' time complexities are \x{in} \(\Theta(N)\).

% This analysis only makes sense when \acrshort{io} is pivotal to an algorithm's
% efficiency.

% The objective is to \alts{
%   {design algorithms requiring a \alts{minimum number, minimum, minimal number} of
%   \acrshortpl{io}.},
%   {minimize the number of \alts{\acrshortpl{io}, I/Os} \alts{an algorithm needs,
%   needed}.},
% }

\x{
% Applying this model to \cref{lst:ithare} suggests
The \gls{emm} suggest that we can at best sum containers like in \cref{lst:ithare} using
\(\nicefrac{N}{B}\) \acrshortpl{io}, where \(N\) is the number of items to sum and \(B\)
the amount of them that fit in one block.
% Since there's 5000 containers of size 5000 and my cache line
% size is \SI{64}{B} and fits 8 \mintinline{cpp}{int} values:
% \begin{equation*}
%   N/B = 5000^2/8 = 3125000
% \end{equation*}
We can assume that the version with the list uses almost \(N\) \acrshortpl{io} because of
the random memory access.  Since a cache line fits 16 \mintinline{text}{int} values (\(B =
16\)) it
predicts a performance difference of 16.  This is not very accurate.
Analysis of \acrshortpl{io} needed by an algorithms \x{or data structure} may help inform
the decision of whether to consider it for empirical comparisons, though.
% that can be used prior to empirical comparisons.

% Apparently, the model is not practical for comparing these two data structures.
}

% Now for something completely different.
\x{
The \gls{emm} also suggests another approach to improve \cref{lst:ithare}, if we don't
want to \alts{trade, forgo, give up} the \x{asymptotically} constant time complexity of
insertions and deletions anywhere in the container: an \emph{unrolled list}.
}

% Notes and potential problems:
% * The concept of I/Os that move B contiguous words directly translates to cache lines.
% * There's nothing equivalent to prefetching in the EMM.
% * > The main problem with hardware caches is that they use a fixed simplistic strategy
%   > for deciding which blocks are kept whereas the external memory model gives the
%   > programmer full control over the content of internal memory.
%        -- Algorithms for Memory Hierarchies, page 6
% * We lost associativity too.
% * And we don't model multiple levels of CPU caches, of course.

% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
% > of computation, hence the design of the algorithm is usually motivated by the need to
% > minimise the number if I/Os.  Howeever in the CMM and the IMM the relative miss
% > penalties are far smaller and we have to consider computation costs.
%      -- Algorithms for Memory Hierarchies, Section 8.6.4, page 188

% > In general it is best to hardcode cache line sizes at compile time by using the
% > `getconf` utility as in:
% >    gcc -DCLS=$(getconf LEVEL1_DCACHE_LINESIZE) ...
% > If the binaries are supposed to be generic, the largest cache line size should be
% > used.
%      -- Drepper, p. 50

% \subsubsection{Notes}
\subsubsection{Limitations and Other Models}

% Replacement policies.  Associativity.  More than two levels of memory.

% While \alts{
While the concept of \acrshortpl{io} directly models \alts{
  most\footnote{%
    TODO: not associativity though.
  },
  the
}
\alts{effects, existence} of cache lines,
% are directly modeled by the concept of \acrshortpl{io},
nothing in the \gls{emm} accounts for \x{those of} prefetching.

\input{tex/com.tex}

% [1]: https://en.wikipedia.org/wiki/Analysis_of_algorithms

% vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-13 tex/list-vs-vector.tex
Write a bit about the EMM; revise and tweak stuff

@@
 round-precision=0]{\speedup} times faster when using \mintinline{cpp}{std::vector}.  Some
-of this difference can certainly be attributed to space overhead of the linked list and
-the added indirection, but the more \emph{cache-friendly} memory access pattern of
+of this difference can \x{certainly} be attributed to space overhead of the linked list
+and the added indirection, but the more cache-friendly memory access pattern of
 \mintinline{cpp}{std::vector} is \alts{decisive, key, crucial, vital}~\cite[5\psq]{bigos}.

### 2017-06-13 tex/abstract-.tex
Tweak

--- 2017-06-13 tex/emm.tex
@@
 % By whom?
-One of these is the \gls{emm}\alts{
-  ,
+One of these is the \gls{emm}\x{\alts{
   ~\cite[5]{afmh},
   {, used \alts{in~\cite{afmh}, by \textcite{afmh}}.},
-}.
+}}.
 
-\subsection{External Memory Model}
-
-% https://en.wikipedia.org/wiki/Random-access_machine
-
-% TODO: go into more detail about the RAM model?
-\alts[2]{
-  {\alts{Establishing the basis of the, The} \gls{emm}, the widely used \gls{ram} model
-  assumes \say{a \say{sufficiently} large uniform memory}~\cite[5]{afmh} with a constant
-  access time.
-  The},
-  {The \gls{emm} is an extension of the \gls{ram} model.  While the latter assumes \say{a
-  \say{sufficiently} large uniform memory}~\cite[5]{afmh} with a constant access time,
-  the},
-}
-\gls{emm} \alts{divides, \alts{expands upon, extends} this by dividing} the memory into
-\emph{internal} and \emph{external}.
-\alts{The, Only the} internal memory is accessed directly, but its size is limited to
-\(M\) \alts{items, words}.
-The external memory is unbounded, but \alts{
-  can only be accessed indirectly by loading data,
-  data \x{is not accessed directly and} has to be loaded,
-}
-into internal memory \say{using \glslink{io}{I/Os} that move \(B\) contiguous
-\alts{[items], words}}~\cite[5]{afmh}.
-
-% > Although the word "I/O" suggests that external memory should be identified with disk
-% > memory, we are free to choose any two levels of the memory hierarchy for internal and
-% > external memory in the model.
-%        -- Algorithms for Memory Hierarchies, page 6
-The use of the term \emph{\acrshort{io}} here is somewhat non-standard.  While it suggests
-\x{\alts{that, the}} external memory represents an \gls{hdd}, \gls{ssd} or similar, it is
-not constrained which \alts{physical storages, members of an actual memory hierarchy} are
-\alts[2]{represented by, associated \alts{with, to}} \x{the} internal and external memory.
-% Which members of an actual memory hierarchy are \alts{represented by, associated to} the
-% internal and external memory is \alts{not constrained, as desired}.
-If we choose the set of all CPU caches and the main memory, \(B\) becomes the cache line
-size and \(M\) may be in the order of a few \si{\mebi\byte}.
-
-% > The relative miss penalty is much lower for caches or the TLB than for disks, so
-% > constant factors are important, we find that asymptotic analysis is not enough to
-% > determine the performance of algorithms.
-%      -- Algorithms for Memory Hierarchies, page 181
-
-% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
-% > of computation, hence the design of the algorithm is usually motivated by the need to
-% > minimise the number of I/Os.
-%      -- Algorithms for Memory Hierarchies, page 188
-
-% \subsubsection{Caveats}
-
-The number of \acrshortpl{io} an algorithm requires in the \gls{emm} can augment the
-information provided by standard asymptotic complexity analysis%
-\x{, if \acrshort{io} is pivotal to an algorithm's runtime},
-but \alts[2]{can't substitute, is no substitute for} measurements.
-For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
-input\x{, like in \cref{lst:ithare},}\ is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
-linked list in \cref{lst:ithare} probably takes almost \(N\) \acrshortpl{io}, since
-consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
-the maximum predicted performance difference between \mintinline{cpp}{std::vector} and
-\mintinline{cpp}{std::list} is \(B\), the number of items a cache line can hold, which is
-\si{16} in this case.%
-\footnote{%
-  % TODO: elaborate on the size of `int`?
-  A cache line has \alts{\si{64}, \SI{64}{\byte}} and an \mintinline{text}{int} \si{4}
-  bytes on the machine used.
-}
-Recalling that the actual performance difference was \num[round-mode=places,
-round-precision=0]{\speedup}, this is pretty inaccurate, but more informative than saying
-both algorithms' time complexities are \x{in} \(\Theta(N)\).
-
-% This analysis only makes sense when \acrshort{io} is pivotal to an algorithm's
-% efficiency.
-
-% The objective is to \alts{
-%   {design algorithms requiring a \alts{minimum number, minimum, minimal number} of
-%   \acrshortpl{io}.},
-%   {minimize the number of \alts{\acrshortpl{io}, I/Os} \alts{an algorithm needs,
-%   needed}.},
-% }
-
-\x{
-% Applying this model to \cref{lst:ithare} suggests
-The \gls{emm} suggest that we can at best sum containers like in \cref{lst:ithare} using
-\(\nicefrac{N}{B}\) \acrshortpl{io}, where \(N\) is the number of items to sum and \(B\)
-the amount of them that fit in one block.
-% Since there's 5000 containers of size 5000 and my cache line
-% size is \SI{64}{B} and fits 8 \mintinline{cpp}{int} values:
-% \begin{equation*}
-%   N/B = 5000^2/8 = 3125000
-% \end{equation*}
-We can assume that the version with the list uses almost \(N\) \acrshortpl{io} because of
-the random memory access.  Since a cache line fits 16 \mintinline{text}{int} values (\(B =
-16\)) it
-predicts a performance difference of 16.  This is not very accurate.
-Analysis of \acrshortpl{io} needed by an algorithms \x{or data structure} may help inform
-the decision of whether to consider it for empirical comparisons, though.
-% that can be used prior to empirical comparisons.
-
-% Apparently, the model is not practical for comparing these two data structures.
-}
-
-% Now for something completely different.
-\x{
-The \gls{emm} also suggests another approach to improve \cref{lst:ithare}, if we don't
-want to \alts{trade, forgo, give up} the \x{asymptotically} constant time complexity of
-insertions and deletions anywhere in the container: an \emph{unrolled list}.
-}
-
-% Notes and potential problems:
-% * The concept of I/Os that move B contiguous words directly translates to cache lines.
-% * There's nothing equivalent to prefetching in the EMM.
-% * > The main problem with hardware caches is that they use a fixed simplistic strategy
-%   > for deciding which blocks are kept whereas the external memory model gives the
-%   > programmer full control over the content of internal memory.
-%        -- Algorithms for Memory Hierarchies, page 6
-% * We lost associativity too.
-% * And we don't model multiple levels of CPU caches, of course.
-
-% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
-% > of computation, hence the design of the algorithm is usually motivated by the need to
-% > minimise the number if I/Os.  Howeever in the CMM and the IMM the relative miss
-% > penalties are far smaller and we have to consider computation costs.
-%      -- Algorithms for Memory Hierarchies, Section 8.6.4, page 188
-
-% > In general it is best to hardcode cache line sizes at compile time by using the
-% > `getconf` utility as in:
-% >    gcc -DCLS=$(getconf LEVEL1_DCACHE_LINESIZE) ...
-% > If the binaries are supposed to be generic, the largest cache line size should be
-% > used.
-%      -- Drepper, p. 50
-
-% \subsubsection{Notes}
-\subsubsection{Limitations and Other Models}
-
-% Replacement policies.  Associativity.  More than two levels of memory.
-
-% While \alts{
-While the concept of \acrshortpl{io} directly models \alts{
-  most\footnote{%
-    TODO: not associativity though.
-  },
-  the
-}
-\alts{effects, existence} of cache lines,
-% are directly modeled by the concept of \acrshortpl{io},
-nothing in the \gls{emm} accounts for \x{those of} prefetching.
-
-\input{tex/com.tex}
+\input{tex/emm}
+\input{tex/com}
 

### 2017-06-13 tex/emm.tex
Tweak

@@
-\section{Abstract?}
-% \section{Analysis of Algorithms}
-
-% > However the efficiencies of any two "reasonable" implementations of a given algorithm
-% > are related by a constant multiplicative factor called a *hidden constant*.
-%      -- https://en.wikipedia.org/wiki/Analysis_of_algorithms
-
-% Mathematical analysis of algorithms to obtain asymptotic upper bounds is nice.  Ignoring
-% the non-uniform nature of memory hierarchies can yield pretty bad results, though.  What
-% can be done about this?
-
-We have seen that
-\alts{
-  {the hidden constant \alts{separating, distinguishing} the time complexities of two
-  reasonable algorithms under asymptotic analysis can get quite big in the presence of a
-  memory hierarchy.},
-  {analyzing \alts{an algorithm's,} \alts{time, asymptotic, algorithmic} complexity,
-  % This is not true: it doesn't matter for asymptotic complexity if there are different
-  % types of memory.
-  \x{under the usual \gls{ram} model, where a single uniform memory is \alts{presumed,
-  assumed, supposed, premised},}
-  can be quite inaccurate in the presence of a memory hierarchy.},
-}
-\alts{
-  To \alts{escape, avert, prevent} having to rely \alts{only, entirely, exclusively} on
-  empirical results, To overcome this problem,
-},
-abstract machine models taking the non-uniform \alts{memories of, nature of memory in}
-real-world computers into account \alts{can be used, are used, have been proposed}.
-% By whom?
-One of these is the \gls{emm}\alts{
-  ,
-  ~\cite[5]{afmh},
-  {, used \alts{in~\cite{afmh}, by \textcite{afmh}}.},
-}.
-
 \subsection{External Memory Model}
@@
 The use of the term \emph{\acrshort{io}} here is somewhat non-standard.  While it suggests
-\x{\alts{that, the}} external memory represents an \gls{hdd}, \gls{ssd} or similar, it is
+\x{\alts{that, the}} external memory represents an \alts{
+  {\gls{hdd} or \gls{ssd}},
+  {\gls{hdd}, \gls{ssd} or similar},
+}, it is
 not constrained which \alts{physical storages, members of an actual memory hierarchy} are
@@
 For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
-input\x{, like in \cref{lst:ithare},}\ is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
-linked list in \cref{lst:ithare} probably takes almost \(N\) \acrshortpl{io}, since
+input, like in \cref{lst:ithare}, is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
+linked list in that example probably takes almost \(N\) \acrshortpl{io}, since
 consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
-the maximum predicted performance difference between \mintinline{cpp}{std::vector} and
-\mintinline{cpp}{std::list} is \(B\), the number of items a cache line can hold, which is
-\si{16} in this case.%
+the predicted performance difference between \mintinline{cpp}{std::vector} and
+\mintinline{cpp}{std::list} is at most \(B\), the number of items a cache line can hold,
+which is \si{16} in this case.%
 \footnote{%
@@
 
-\input{tex/com.tex}
-
-% [1]: https://en.wikipedia.org/wiki/Analysis_of_algorithms
-
 % vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-15 false-sharing/herb.cpp
WIP

@@ -1,0 +1,48 @@
+// vim: tw=72 sts=-1 sw=3 et
+
+#include <thread>
+#include <algorithm>
+#include <vector>
+#include <iostream>
+#include <cstdlib>  // std::rand
+#include <ctime>    // std::clock, std::time, std::clock_t
+#include <chrono>   // std::chrono::high_resolution_clock
+
+constexpr int DIM = 20000;
+
+int main() {
+   auto matrix_ptr = std::make_unique<std::array<int, DIM * DIM>>();
+   auto& matrix = *matrix_ptr;
+   std::srand(std::time(nullptr));
+   for (std::size_t i = 0; i < DIM * DIM; ++i) {
+      matrix[i] = std::rand();
+   }
+
+   auto t0 = std::chrono::high_resolution_clock::now();
+   int result[THREADS];
+   std::vector<std::thread> threads;
+   // Each of THREADS parallel workers processes 1/THREADS-th of the
+   // data; the p-th worker records it's partial count in result[p]
+   for (std::size_t p = 0; p < THREADS; ++p)
+      threads.push_back(std::thread([&, p] {
+         result[p] = 0;
+         int count = 0;  // XXX
+         int chunkSize = DIM / THREADS + 1;
+         int myStart = p * chunkSize;
+         int myEnd = std::min(myStart + chunkSize, DIM);
+         for (int i = myStart; i < myEnd; ++i)
+            for (int j = 0; j < DIM; ++j)
+               // if (matrix[i * DIM + j] % 2 != 0) ++result[p];
+               if (matrix[i * DIM + j] % 2 != 0) ++count;
+         result[p] = count;  // XXX
+      }));
+   // Wait for the parallel work to complete...
+   for (auto& thread : threads) thread.join();
+   // Finally, do the sequential "reduction" step to combine the results
+   int odds = 0;
+   for (int p = 0; p < THREADS; ++p) odds += result[p];
+   auto t1 = std::chrono::high_resolution_clock::now();
+
+   std::chrono::duration<double, std::milli> duration = t1 - t0;
+   std::cout << odds << '\n' << duration.count() << '\n';
+}

### 2017-06-19 tex/com.tex
Expand content, revise, tweak

@@
-\subsection{Cache-Oblivious Model} % Cache-Oblivious Algorithms
+\subsection{Cache-Oblivious Model}
+%           Model-Oblivious Algorithms
+%           Ideal-Cache Model
+%           \Gls{com}
 
-% The \gls{com} is like the \gls{emm}, but we pretend not to know the values of B and M.
-% This seems a bit silly.
+% This one's kind of nice (and Scott Meyers seems to think it has merit).  It also has a
+% Wikipedia page, unlike the EMM, CMM, or IMM (a very rough index of practicality).
+
+% This seems like FUD:
+%
+% > Typical cache-efficient algorithms require tuning to several cache parameters which
+% > are not always available from the manufacturer and often difficult to extract
+% > automatically.
+%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf
+
+% > In computing, a cache-oblivious algorithm (or cache-transcendent algorithm) is an
+% > algorithm designed to take advantage of a CPU cache without having the size of the
+% > cache (or the length of the cache lines, etc.) as an explicit parameter.
+%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
+%
+
+The \emph{\gls{com}} or \emph{ideal-cache model}, introduced by
+\textcite{coa-for-publication}, \alts{concedes, leaves, forfeits} most of the
+aforementioned problems to empirical evaluation and further increases the level of
+abstraction.
+%
+Algorithms for the \gls{com}, called \emph{cache-oblivious algorithms}, are designed
+without the \alts{
+  cache size \(M\) or block size \(B\),
+  values of \(M\) and \(B\),
+}
+as parameters.  This seems silly since these values typically can be queried easily
+\x{automatically} at both run and compile time~\cite[50]{drepper2007}
+% See <https://stackoverflow.com/a/7284876>.  Also, cache line sizes are super homogeneous
+% anyway.
+but, perhaps counterintuitively, models one aspect of memory hierarchies better than the
+\gls{emm}.
+% > This model was born out of the necessity to capture the hierarchical nature of memory
+% > organization. [...] Although there have been other attempts to capture this
+% > hierarchical information the cache oblivious model seems to be one of the most simple
+% > and elegant ones.
+%      -- Algorithms for Memory Hierarchies, page 194
+%
+% > The ideal cache oblivious model enables us to reason about a two level memory like the
+% > external memory model but prove results about a multi-level memory model.
+%      -- Algorithms for Memory Hierarchies, page 195
+%
+% > One consequence is that, if a cache-oblivious algorithm performs well between two
+% > levels of the memory hierarchy (nominally called cache and disk), then it must
+% > automatically work well between any two adjacent levels of the memory hierarchy.
+%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf (erikcom)
+%
+% > Good algorithms for this model give us good algorithms for all values of B and M.
+% > They are especially useful for multi-level caches.
+%      -- https://ocw.mit.edu/courses/electrical-engineering-and-computer-science/6-851-advanced-data-structures-spring-2012/calendar-and-notes/MIT6_851S12_L7.pdf
+%
+% > From a theoretical standpoint, the cache-oblivious model is appealing because it is
+% > very clean.  A cache-oblivious algorithm is simply a RAM algorithm; it is only the
+% > analysis that differs.  The cache-oblivious model also works well for multilevel
+% > memory hierarchies, unlike the external-memory model, which only captures a two-level
+% > hierarchy.
+%      -- https://pdfs.semanticscholar.org/2f5a/76ccdc71971b746fbe7ff54db86c65a73e91.pdf
+%
+% > Frigo et al. showed that for many problems, an optimal cache-oblivious algorithm will
+% > also be optimal for a machine with more than two memory hierarchy levels.
+%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
+%
+% > [W]e prove that an optimal cache-oblivious algorithm designed for two levels of memory
+% > is also optimal for multiple levels.  We also prove that any optimal cache-oblivious
+% > algorithm is also optimal in the previously studied HMM and SUMH models.
+%      -- https://pdfs.semanticscholar.org/19ed/9795adc7204d3c9745b3c9f71f8496d26bb6.pdf
+%
+% > We prove that an optimal cache-oblivious algorithm designed for two levels of memory
+% > is also optimal for multiple levels and that the assumption of optimal replacement in
+% > the ideal-cache model can be simulated efficiently by LRU replacement.
+%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1, Abstract
+%
+An algorithm that performs well in the \gls{com} performs well across the entire memory
+hierarchy~\cites[194\psq]{afmh}[4]{erikcom}; the same argument \alts[2]{\alts{for,
+showing} asymptotically optimal movement of data, for data movement being
+\x{asymptotically} \emph{optimal}} applies between any two levels of memory%
+~\cite[lemma 15, \pno~10]{coa-paper}.
+% TODO.  Is asymptotic analysis even useful, though?  The whole point of these models is
+% that the hidden constants of "standard" asymptotic analysis matter.
+
+% What does "optimal" mean?
+%
+% > An optimal cache-oblivious algorithm is a cache-oblivious algorithm that uses the
+% > cache optimally (in an asymptotic sense, ignoring constant factors).
+%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
+%
+% > We say that [an algorithm] is *cache optimal* if the number of cache misses meet the
+% > asymptotic lower bound for I/Os in the EMM for that problem.
+%      -- Algorithms for Memory Hierarchies, page 181
+%
+% > [T]hese algorithms use an optimal amount of work and move data optimally among
+% > multiple levels of cache.
+%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1, Abstract
+%
+% > For simplicity in this paper, we use the term "optimal" as a synonym for
+% > "asymptotically optimal", since all our analyses are asymptotic.
+%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=2, footnote
+Optimal means that the asymptotic~\cite[2]{coa-paper} number of cache misses incurred
+\x{by a cache-oblivious algorithm} matches the problem's lower bound in the \gls{com}.
+
+% > In contrast to the external-memory model, algorithms in the cache-oblivious model
+% > cannot explicitly manage the cache (issue block-read and block-write requests).  This
+% > loss of freedom is necessary because the block and cache sizes are unknown.
+%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf (erikcom), page 5
+Cache misses take the place of \acrshortpl{io} \x{used in the \gls{emm}} because
+cache-oblivious algorithms don't \alts{manage the cache, read or write cache lines}
+explicitly.  This wouldn't be possible since the algorithms know neither the cache nor
+the cache line size~\cite[5]{erikcom}.
+% > The ideal cache uses the optimal off-line strategy of replacing the cache line whose
+% > next access is furthest in the future
+%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1
+Instead, the \gls{com} uses the optimal replacement strategy of evicting the cache line
+that won't be accessed for the longest time in the future (Bélády's Algorithm).  This
+\alts{
+  is strangely \alts{at odds, out of touch, unrealistic} with real-world caches that don't
+  know the future,
+  {seems like out-of-touch, theoretical ivory-tower nonsense},
+}.
+% > The ideal-cache model makes the perhaps-questionable assumptions that there are only
+% > two levels in the memory hierarchy, that memory is managed automatically by an optimal
+% > cache-replacement strategy, and that the cache is fully associative.  We address these
+% > assumptions in Section 6, showing that to a certain extent, these assumptions entail
+% > no loss of generality.
+%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=2
+%
+% > LRU and FIFO replacement do just as well as optimal replacement up to a constant
+% > factor of memory transfers and up a constant factor wastage of the cache.
+%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
+%
+% > The same argument extends to a variety of other replacement strategies.
+%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46
+%
+% > We show that algorithms with regular complexity bounds (Equation (7.1)) (including all
+% > algorithms heretofore presented) can be ported to less-ideal caches incorporating
+% > least-recently-used (LRU) or first-in, first-out (FIFO) replacement policies [24, p.
+% > 378].
+%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=51
+%
+% > [I]n many cases [the COM] is provably within a constant factor of a more realistic
+% > cache's performance.
+%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
+%
+% > [A]s long as the number of memory transfers depends polynomially on the cache size M,
+% > then halving M will only affect the running time by a constant factor.
+%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
+%
+% > Intuitively, algorithms that slow down by a constant factor when memory (M) is reduced
+% > to half, are called regular.
+%      -- Algorithms for Memory Hierarchies, page 196
+However, \citeauthor{coa-thesis} proves that for many algorithms\footnote{%
+   Those satisfying equation (7.1) in~\cite[46]{coa-thesis}.  If the number of cache
+   misses incurred by the algorithm only \say{depends polynomially on the cache size
+   \(M\)}, the \alts{equation, condition} is satisfied~\cite[6]{erikcom}.
+   % Halving the cache size only increases the number of cache misses by a constant
+   % factor?
+}
+it only increases cache misses by a constant factor compared to various feasible
+replacement strategies%
+% ~\cite[lemma 12, \pno~10]{coa-paper}.%
+% ~\cite[corollary 13, \pno~10]{coa-paper}.%
+~\cite[corollary 19, \pno~46]{coa-thesis}.%
+% See <http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46>.
+\footnote{%
+   E.g. LRU, FIFO, and random replacement
+}
+
+% According to `coa-paper` (page 9), the "four major assumptions" are:
+% *  automatic replacement
+% *  optimal replacement
+% *  two levels of memory
+% *  full associativity
+%
+% All simplifications and assumptions (I think):
+% *  inherited/adopted from the EMM
+%    *  two levels of memory
+%    *  full associativity
+%    *  tall cache
+%       *  "It is also commonly assumed in external-memory algorithms."
+%             -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=7
+%       *  I don't grok this one.
+% *  new
+%    *  automatic, optimal replacement
+%    *  inclusion property
+%       *  "[T]he values stored in cache i are also stored in cache i+1"
+%             -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=10
+%
+% TODO: reiterate all simplifications and assumptions?
+
+% > We show that the assumptions of two hierarchical memory models in the literature, in
+% > which memory movement is programmed explicitly, are actually no weaker than ours.
+% > Specifically, we prove (with only minor assumptions) that optimal cache-oblivious
+% > algorithms in the ideal-cache model are also optimal in the hierarchical memory model
+% > (HMM) [1] and in the serial uniform memory hierarchy (SUMH) model [5, 42].
+%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=12
+%
+% Exactly the same text is in the paper.  But only in the earlier version [1].
+% [1]: https://pdfs.semanticscholar.org/19ed/9795adc7204d3c9745b3c9f71f8496d26bb6.pdf
+%
+% > An optimal cache-oblivious algorithm whose cache-complexity bound satisfies the
+% > regularity condition (14) can be implemented optimally in expectation in multilevel
+% > models with explicit memory management.
+%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=11, theorem 17
+%
+% > [W]e have shown that optimal cache-oblivious algorithms in the ideal-cache model are
+% > also optimal in the hierarchical memory model (HMM).
+%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=56
+%
+\citeauthor{coa-thesis} further justifies the model by proving, \say{with only minor
+assumptions}~\cite[12]{coa-thesis}, that cache-oblivious algorithms that are optimal in
+the \gls{com} \alts{can by executed with an optimal amount of \acrshortpl{io}, are also
+optimal} in the \gls{emm}~\cite[theorem 32, \pno~56]{coa-thesis}.  In other words, most
+cache-oblivious algorithms can be systematically transformed into \alts{cache-aware,
+\gls{emm}} algorithms that asymptotically require the same amount of memory transfers in
+the \gls{emm} as the cache-oblivious variant in the \gls{com}.
+
+% TODO: given our motivation that "constant factors matter", what have we been doing here?
+
+% TODO.  A regular algorithm that isn't explicitly for a hierarchical memory model
+% (doesn't explicitly move data between different levels of memory) is already
+% cache-oblivious, right?
+
+% Time to give an example algorithm.  Options:
+% *  matrix transposition
+% *  matrix multiplication
+% *  searching a tree (Van Emde Boas layout vs. standard layout)
+% *  sorting
+%    *  not easy
+%    *  good performance seems problematic
+%    *  looks like the autors of [1] dit it, though (lazy funnelsort)
+% *  something about linked lists?
+%
+% Maybe matrix transposition...
+%
+% [1] https://app.cs.amherst.edu/~ccmcgeoch/cs34/papers/a2_2-brodal.pdf
+\subsubsection{Cache-Oblivious Matrix Transposition}
+
+% Work by Erik Demaine:
+%
+% [1]: http://erikdemaine.org/papers/BRICS2002/paper.pdf
+%      "Cache-Oblivious Algorithms and Data Structures - 2002"
+% [2]: https://web.archive.org/web/20160320051801/http://courses.csail.mit.edu/6.897/spring03/scribe_notes/L15/lecture15.pdf
+%      "6.897: Advanced Data Structures - Spring 2003"
+% [3]: https://pdfs.semanticscholar.org/2f5a/76ccdc71971b746fbe7ff54db86c65a73e91.pdf
+%      "6.897: Advanced Data Structures - Spring 2005"
+% [4]: https://ocw.mit.edu/courses/electrical-engineering-and-computer-science/6-851-advanced-data-structures-spring-2012/calendar-and-notes/MIT6_851S12_L7.pdf
+%      "6.851: Advanced Data Structures - Spring 2012"
 

### 2017-06-19 tex/emm.tex
Expand content, revise, tweak

@@
 
-% > The relative miss penalty is much lower for caches or the TLB than for disks, so
-% > constant factors are important, we find that asymptotic analysis is not enough to
-% > determine the performance of algorithms.
-%      -- Algorithms for Memory Hierarchies, page 181
-
-% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
-% > of computation, hence the design of the algorithm is usually motivated by the need to
-% > minimise the number of I/Os.
-%      -- Algorithms for Memory Hierarchies, page 188
-
-% \subsubsection{Caveats}
-
 The number of \acrshortpl{io} an algorithm requires in the \gls{emm} can augment the
@@
 \x{, if \acrshort{io} is pivotal to an algorithm's runtime},
-but \alts[2]{can't substitute, is no substitute for} measurements.
+but \alts[2]{can't substitute, is no substitute for} measurements\x{~\cite[181]{afmh}}.
 For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
-input, like in \cref{lst:ithare}, is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
-linked list in that example probably takes almost \(N\) \acrshortpl{io}, since
-consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
-the predicted performance difference between \mintinline{cpp}{std::vector} and
-\mintinline{cpp}{std::list} is at most \(B\), the number of items a cache line can hold,
-which is \si{16} in this case.%
+input of size \(N\), like in \cref{lst:ithare}, is
+% See <http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=7>.
+\alts[3]{
+  \(\lceil\nicefrac{N}{B}\rceil\),
+  \(\left(\lfloor\nicefrac{N}{B}\rfloor + 2\right)\),
+  \(\left(\lceil\nicefrac{N}{B}\rceil + 1\right)\),
+}
+in the \gls{emm}.\footnote{%
+  TODO.
+}
+The linked list in that example probably takes almost \(N\) \acrshortpl{io},
+though, since consecutive \alts{nodes, elements} are unlikely to fall into the same cache
+line.  Thus, the predicted performance difference between \mintinline{cpp}{std::vector}
+and \mintinline{cpp}{std::list} is at most \(B\), the number of items a cache line can
+hold, which is \si{16} in this case.%
 \footnote{%
   % TODO: elaborate on the size of `int`?
-  A cache line has \alts{\si{64}, \SI{64}{\byte}} and an \mintinline{text}{int} \si{4}
-  bytes on the machine used.
+  A cache line is \alts{\si{64} bytes, \SI{64}{\byte}} on my laptop's CPU and an
+  \mintinline{text}{int} \si{4} bytes with my compiler.
 }
-Recalling that the actual performance difference was \num[round-mode=places,
-round-precision=0]{\speedup}, this is pretty inaccurate, but more informative than saying
-both algorithms' time complexities are \x{in} \(\Theta(N)\).
+Recalling that the \alts{measured, actual} performance difference was
+\num[round-mode=places, round-precision=0]{\speedup}, this is pretty inaccurate, but more
+informative than saying both \x{algorithms'} data structure's time complexities for
+traversal are \x{in} \(\Theta(N)\).
+
+% An extensive repository of the theoretical lower bounds of \acrshortpl{io} needed in the
+% \gls{emm} for many problems is available in the literature (for example \cite{afmh}).
+
+Even with the simplifications made by the \gls{emm}, algorithmic analysis is usually only
+done asymptotically: the number of \acrshortpl{io} is expressed \alts{in terms of, as}
+\(\mathcal{O}\left(f\left(N, M, B\right)\right)\) or one of the related symbols.
+Theoretical lower bounds of the \acrshortpl{io} needed in the \gls{emm} are available in
+the literature (for example~\cite{afmh}) for many problems.
 
@@
 
+% > In general it is best to hardcode cache line sizes at compile time by using the
+% > `getconf` utility as in:
+% >    gcc -DCLS=$(getconf LEVEL1_DCACHE_LINESIZE) ...
+% > If the binaries are supposed to be generic, the largest cache line size should be
+% > used.
+%      -- Drepper, p. 50
+
+% \subsubsection{Notes}
+% \subsubsection{Caveats}
+% \subsubsection{Limitations and Other Models}
+\subsubsection{Limitations}
+
 % Notes and potential problems:
@@
 
+% Replacement policies.  Associativity.  More than two levels of memory.
+
+% While the concept of \acrshortpl{io} directly models \alts{
+%   most\footnote{%
+%     TODO: not associativity though.
+%   },
+%   the
+% }
+% \alts{effects, existence} of cache lines,
+% % are directly modeled by the concept of \acrshortpl{io},
+% nothing in the \gls{emm} accounts for \x{those of} prefetching.
+
+While the concept of \acrshortpl{io} directly models cache lines, most other
+characteristics of memory hierarchies are ignored by the \gls{emm}.  For example:
+\begin{itemize} % TODO: probably don't use a list structure.
+  \item prefetching, or more generally the advantages of sequential access patterns,
+  \item multi-level caches,
+  \item the lack of direct control over the contents of caches, % (\alts{CPUs,
+    % \glspl{cpu}} employ fixed, hardware-controlled replacement strategies)
+  \item associativity,\footnote{%
+      Associativity is not discussed in this paper; see \textcite{drepper2007} instead.
+    }
+  \item \gls{tlb}.
+  % critical word stuff...
+\end{itemize}
+%
 % > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
 % > of computation, hence the design of the algorithm is usually motivated by the need to
-% > minimise the number if I/Os.  Howeever in the CMM and the IMM the relative miss
+% > minimise the number of I/Os.  However in the CMM and the IMM the relative miss
 % > penalties are far smaller and we have to consider computation costs.
 %      -- Algorithms for Memory Hierarchies, Section 8.6.4, page 188
+%
+% > The relative miss penalty is much lower for caches or the TLB than for disks, so
+% > constant factors are important, we find that asymptotic analysis is not enough to
+% > determine the performance of algorithms.
+%      -- Algorithms for Memory Hierarchies, page 181
+%
+\alts{More fundamentally, Additionally}, the model's premise is that \acrshortpl{io} are
+much more expensive than computation~\cite[188]{afmh}.
+% This doesn't apply to data transfer between main memory and caches to the extent it does
+% when accessing \glspl{hdd}.
+While this is \alts{plausible, sensible, reasonable} when accessing \glspl{hdd}, it
+doesn't apply to data transfer between main memory and caches to nearly the same extent.
+%
+Some of these shortcomings are addressed by refined machine models, which include more
+details of real caches~\cite[178]{afmh}.
+% I.e., CMM (cache memory model) and IMM (internal memory model).  There doesn't seem to
+% be much literature about them [1].  Algorithms for Memory Hierarchies only cites a
+% single paper that uses the IMM.  Don't discuss them.
+%
+% [1]: https://www.google.com/search?q=cache+memory+model+cmm
+This complicates mathematical analysis~\cite[181]{afmh} and heuristics may be
+used~\cite[191]{afmh}, which in turn \alts{exacerbates, amplifies} the need for
+accompanying measurements~\cite[181]{afmh}.
 
-% > In general it is best to hardcode cache line sizes at compile time by using the
-% > `getconf` utility as in:
-% >    gcc -DCLS=$(getconf LEVEL1_DCACHE_LINESIZE) ...
-% > If the binaries are supposed to be generic, the largest cache line size should be
-% > used.
-%      -- Drepper, p. 50
+\x{
+\alts{
+  {\alts[2]{On the contrary, Contrarily}, the \emph{\gls{com}} further increases the level
+  of abstraction},
+  {The opposite approach -- towards more abstraction -- is taken by the \gls{com}},
+}.
+}
 
-% \subsubsection{Notes}
-\subsubsection{Limitations and Other Models}
-
-% Replacement policies.  Associativity.  More than two levels of memory.
-
-% While \alts{
-While the concept of \acrshortpl{io} directly models \alts{
-  most\footnote{%
-    TODO: not associativity though.
-  },
-  the
-}
-\alts{effects, existence} of cache lines,
-% are directly modeled by the concept of \acrshortpl{io},
-nothing in the \gls{emm} accounts for \x{those of} prefetching.
+% CMM and IMM add more detail ... bla bla bla ... COM actually abstracts further ...  bla
+% bla ... counterintuitively this makes it more fitting in one aspect ... it works across
+% the whole memory hierarchy ...
 

### 2017-06-19 tex/list-vs-vector.tex
Expand content, revise, tweak

@@
 \section{Example: \texttt{std::vector} vs. \texttt{std::list}}
+\label{sec:lvv}
 
@@
 
+\subsection{``True OO Style''}
+
+% TODO.  Having a vector of pointers to dynamically allocated object derived from some
+% base class in true OO style (tm) probably isn't going to make the cache happy.  Bjarne
+% Stroustrup talked about this [1].  Strongly related: data-oriented design [3].
+%
+% [1]: https://youtu.be/0iWb_qi2-uI?t=51m22s
+%      "GoingNative 2012: Day 1 Keynote - Bjarne Stroustrup: C++11 Style"
+% [2]: https://channel9.msdn.com/Events/GoingNative/GoingNative-2012/Keynote-Bjarne-Stroustrup-Cpp11-Style
+%      "GoingNative 2012: Day 1 Keynote - Bjarne Stroustrup: C++11 Style"
+% [3]: https://youtu.be/rX0ItVEVjHc
+%      "CppCon 2014: Mike Acton 'Data-Oriented Design and C++'"
+
+% A nice compact, flat container isn't going to cut it if it contains pointers.
+
+In object-oriented systems, objects are typically referred to by pointers to a common base
+class.  A polymorphic container of such pointers allows for dynamic dispatch of virtual
+functions.  This carries the risk of degrading the \x{superior} performance of a
+sequential data structure to that of a list.
+
+TODO: create an illustration with TikZ.
+\begin{tikzpicture}
+\node [draw] (object1) {};
+\end{tikzpicture}
+
+TODO~\cite[42:00]{stroustrup-talk}.
+
+\subsection{Unrolled List}
+
+% TODO.  Talk about unrolled lists and software prefetching.
+
 % vim: tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-19 tex/list-vs-vector.tex
Add TikZ graphic representing an array of pointers

@@
 
-\subsection{``True OO Style''}
+\subsection{``True'' OO Style}
 
@@
 class.  A polymorphic container of such pointers allows for dynamic dispatch of virtual
-functions.  This carries the risk of degrading the \x{superior} performance of a
-sequential data structure to that of a list.
+functions.  However, this carries the risk of degrading the \x{superior} performance of a
+sequential data structure to that of a list%
+% ~\cite[\href{https://youtu.be/0iWb_qi2-uI?t=51m22s}{51:22}]{stroustrup-talk}.
+~\cite[51:22]{stroustrup-talk}.
+See \cref{fig:true-oo}.
 
-TODO: create an illustration with TikZ.
-\begin{tikzpicture}
-\node [draw] (object1) {};
-\end{tikzpicture}
-
-TODO~\cite[42:00]{stroustrup-talk}.
+\begin{figure}
+   \centering
+   \input{tex/oo-picture}
+   \captionsetup{width=.9\linewidth} % FIXME: hacks!
+   \caption[Array of Pointers]{Array of Pointers; while the array is compact, the actual
+   objects may not be laid out sequentially in memory.}
+   \label{fig:true-oo}
+\end{figure}
 

### 2017-06-22 tex/com.tex
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/

@@
 \footnote{%
-   E.g. LRU, FIFO, and random replacement
+   e.g. LRU, FIFO, and random replacement
 }
@@
 
+The straightforward way to transpose \alts{
+   an \(m\times n\) matrix \(D\),
+   a matrix \(D\in\mathbb{R}^{m\times{}n}\),
+}
+out-of-place is to use two loops like so:
+\begin{minted}[autogobble]{c}
+   for (int i = 0; i < m; ++i)
+      for (int j = 0; j < n; ++j)
+         E[j][i] = D[i][j];
+\end{minted}
+Assuming \(D\) and \(E\) are stored in row-major layout (as they would be in the C and
+\cpp{} languages), the reads from \(D\) are sequential memory accesses but the writes to
+\(E\) are not.
+
+\alts{When \(D^\mathsf{T}\) has, For} sufficiently long rows (\(m > B\))\footnote{%
+   \(D^\mathsf{T}\) will be an \(n\times m\) matrix, so \(m\) is its number of columns.
+},
+every consecutive access will be to a different cache line.  If it has sufficiently many
+rows, no access will be to a cache line that is still loaded; once the inner loop
+completes, the cache line holding \alts[2]{\mintinline{c}{E[0][i]}, \(B_{0,i}\)} will have
+been evicted.%
+\footnote{%
+   assuming LRU or FIFO replacement % FIXME: but we use optimal replacement!
+}
+Therefore, this algorithm incurs \(\Theta(mn)\) cache misses.
+
+The algorithm is \alts{by definition, technically} already cache-oblivious since it
+doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
+with a divide-and-conquer approach.
+% The code may only be 72 characters wide (75 when counting the indentation that will be
+% stripped).  Note that the array arguments will decay to pointers.
+\begin{minted}[autogobble, mathescape]{c}
+   // Transpose the submatrix $(d_{ij})_{i\in I,\:j\in J}$.
+   void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
+      int num_rows = 1 + I[1] - I[0];
+      int num_cols = 1 + J[1] - J[0];
+      if (num_cols == 1 && num_rows == 1) {
+         E[J[0]][I[0]] = D[I[0]][J[0]];
+      } else if (num_cols <= num_rows) {
+         // Horizontally slice D into two submatrices and recurse.
+         transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
+         transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
+      } else { /* Vertically slice D analogously... */ }
+   }
+\end{minted}
+% // Transpose the submatrix $(d_{ij}),\: i\in I, j\in J$.
+% // Transpose the submatrix $(d_{ij})_{I_0<i<I_1, J_0<j<J_1}$
+% // Transpose the submatrix $D\left[...\right]$
+
+\begin{figure}
+   \centering
+   \begin{tikzpicture}
+      \datavisualization
+      [scientific axes=clean,
+       x axis={logarithmic,
+               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
+                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256},
+               },
+              },
+       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
+         % ticks={major also at={1}}
+       },
+       visualize as scatter,
+       scatter={style={mark=*, mark options={scale=.65}}}]
+         data [read from file=xpose/speedup.csv, separator=\space];
+   \end{tikzpicture}
+   \caption{TODO}
+   \label{fig:TODO}
+\end{figure}
+
+% > A cache-oblivious algorithm is simply a RAM algorithm; it is only the analysis that
+% > differs.
+%      -- https://pdfs.semanticscholar.org/2f5a/76ccdc71971b746fbe7ff54db86c65a73e91.pdf
+%
+% Algorithms are cache-oblivious by default.
+
 % Work by Erik Demaine:

### 2017-06-22 tex/list-vs-vector.tex
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/

@@
 \alts{%
-   The C++ program shown in \cref{lst:ithare},
-   The following C++ program,
+   The \cpp{} program shown in \cref{lst:ithare},
+   The following \cpp{} program,
 }%
@@
 
-\subsection{``True'' OO Style}
+\subsection{Notes}
+
+\subsubsection{``True'' OO Style}
 
@@
 
-In object-oriented systems, objects are typically referred to by pointers to a common base
+In \gls{oo} systems, variables are typically referred to by pointers to a common base
 class.  A polymorphic container of such pointers allows for dynamic dispatch of virtual
@@
    \captionsetup{width=.9\linewidth} % FIXME: hacks!
-   \caption[Array of Pointers]{Array of Pointers; while the array is compact, the actual
-   objects may not be laid out sequentially in memory.}
+   \caption[Array of Pointers]{Array of Pointers; the array is compact but the actual
+   objects may \alts{be scattered across memory pretty randomly, not be laid out
+   sequentially in memory}.}
    \label{fig:true-oo}
@@
 
-\subsection{Unrolled List}
+\subsubsection{Unrolled List}
 

### 2017-06-22 line-size/line-size.c

@@ -1,9 +1,9 @@
 // vim: tw=72 sts=-1 sw=3 et cms=//%s
 
-#include "stdint.h"  // int64_t, size_t
-#include "stdio.h"   // printf
-#include "stdlib.h"  // calloc
-#include "time.h"    // clock
+#include <stdint.h>  // int64_t, size_t
+#include <stdio.h>   // printf
+#include <stdlib.h>  // calloc
+#include <time.h>    // clock
 
 // Adapted from <igoro.com/archive/gallery-of-processor-cache-effects/>.
 //

### 2017-06-23 tex/basic-concepts/lines.tex
Add more slides, tweak stuff

--- 2017-06-05 tex/basic-concepts/lines.tex
@@
 
-% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
-% interrupting the source code listing.
-\begin{figure}%[hbp]
+\begin{figure}
    \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={logarithmic,
-               ticks={major at={1, 2, 4, 8 as \textbf{8}, 16, 64, 256, 1024},
-                      minor at={32, 128, 512}},
-               grid={at=8},
-               label={Step Size}, length=0.8\textwidth},
-       y axis={logarithmic,
-               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
-               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
-               ticks={major at={10, 20, 40, 80, 160, 320}},
-               grid=at ticks,
-               label={Processor Time (ms)}, length=6cm},
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         % See <https://tex.stackexchange.com/q/198323>.
-         data [read from file=line-size/line-size.csv, separator=\space];
-   \end{tikzpicture}
+   \input{tex/graphics/line-size-plot}
    \caption{Processor Times for Running \Cref{lst:line-size}}
@@
 
-% vim: tw=90 sts=-1 sw=3 et fdm=marker
+% vim: ft=tex tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-25 tex/com.tex
Minor tweaks and additions

@@
 every consecutive access will be to a different cache line.  If it has sufficiently many
-rows, no access will be to a cache line that is still loaded; once the inner loop
-completes, the cache line holding \alts[2]{\mintinline{c}{E[0][i]}, \(B_{0,i}\)} will have
-been evicted.%
+rows,
+\alts{{%
+      the constant amount of lines that are still kept in cache by the optimal replacement
+      strategy once the inner loop completes and we need them again becomes
+      negligible.\x{\footnote{%
+         With LRU or FIFO replacement, ...
+   }}}, {
+      no access will be to a cache line that is still loaded; once the inner loop
+      completes, the cache line holding \alts[2]{\mintinline{c}{E[0][i]}, \(B_{0,i}\)}
+      will have been evicted.%
       \footnote{%
-   assuming LRU or FIFO replacement % FIXME: but we use optimal replacement!
+         assuming LRU or FIFO replacement % XXX: but we use optimal replacement!
+      }%
+   },
 }
@@
 doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
-with a divide-and-conquer approach.
+with a divide-and-conquer approach.  The idea is to recursively divide the input matrix
+\(D\) into two equal-sized submatrices along the greater dimension.  If \(m \geq n\) (more
+rows than columns), let
+\begin{equation*}
+   D = \begin{bmatrix}
+      D_1\\
+      D_2
+   \end{bmatrix}
+\end{equation*}
+and use \(D^\mathsf{T} = \begin{bmatrix} D_1^\mathsf{T} & D_2^\mathsf{T} \end{bmatrix}\)
+to compute the transpose.  If \(m < n\), analogously split \(D\) vertically.  Eventually,
+the submatrices will become small enough so that pairs of input and output submatrices fit
+into cache at the same time, at which point it doesn't matter in what order we access the
+elements.\footnote{%
+   No cache line will have to be evicted once loaded.
+}
+The recursion continues all the way down to \(1 \times 1\) submatrices, but this doesn't
+change the theoretical analysis.  According to
+\textcite[theorem 2 and 3, \ppno~19--21]{coa-thesis}
+% \textcite[theorem 2 and 3, \ppno~19\psqq]{coa-thesis}
+this algorithm incurs an optimal amount of \(\Theta(1 + mn / B)\) cache misses.
+% This reduces the amount of cache misses to \(\Theta(1 + \nicefrac{mn}{B})\) according to
+% \textcite[theorem 2, \ppno~19\psq]{coa-thesis}.
+
+In practice, the new algorithm performs as badly as one might expect from this excessive
+use of recursion.  \Cref{fig:xpose-speedup} gives the speedup compared to the
+straightforward implementation, which is usually faster.
+
 % The code may only be 72 characters wide (75 when counting the indentation that will be
@@
    \caption{TODO}
-   \label{fig:TODO}
+   \label{fig:xpose-speedup}
 \end{figure}

### 2017-06-25 ithare/sw-prefetching/list-vs-vector.cpp
Run and rerun some measurements, add `xpose.c`
I think I didn't run `list-vs-vector.cpp` under the setup used to reduce
noise (few processes running, CPU frequency scaling disabled) before.

@@ -1,0 +1,41 @@
+// vim: tw=72 sts=-1 sw=3 et
+
+#include <cstdlib>  // std::rand, std::srand
+#include <ctime>    // std::clock, std::time, std::clock_t
+#include <iostream>
+#include <list>
+
+#include <xmmintrin.h>
+
+// No prefetching can save this program.
+
+using Container = std::list<int>;
+
+constexpr int N = 5000;
+
+int main() {
+   Container containers[N];
+   std::srand(std::time(nullptr));
+   // Append an average of 5000 random values to each container.
+   for (int i = 0; i < N * 5000; ++i) {
+      containers[std::rand() % N].push_back(std::rand());
+   }
+
+   int sum = 0;
+   std::clock_t t0 = std::clock();
+   for (int m = 0; m < N; ++m) {
+      auto it = containers[m].begin();
+      auto next = std::next(it, 34);
+      auto end = containers[m].end();
+      while (it != end) {
+         ++next;
+         _mm_prefetch(next._M_node, _MM_HINT_T0);
+         sum += *it;
+         ++it;
+      }
+   }
+   std::clock_t t1 = std::clock();
+
+   // Also print the sum so the loop doesn't get optimized out.
+   std::cout << sum << '\n' << (t1 - t0) << '\n';
+}

### 2017-06-25 xpose/xpose.c

@@ -1,0 +1,65 @@
+#include <stdint.h>  // size_t
+#include <stdio.h>   // printf
+#include <stdlib.h>  // calloc, `int rand()`, srand
+#include <time.h>    // clock
+
+// #define m 5000
+// #define n 5000
+
+#ifdef SIMPLE
+void transpose(int D[m][n], int E[n][m]) {
+   for (int i = 0; i < m; ++i)
+      for (int j = 0; j < n; ++j) E[j][i] = D[i][j];
+}
+#else
+void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
+   int num_rows = 1 + I[1] - I[0];
+   int num_cols = 1 + J[1] - J[0];
+   if (num_cols == 1 && num_rows == 1) {
+      E[J[0]][I[0]] = D[I[0]][J[0]];
+   } else if (num_cols <= num_rows) {
+      // Horizontally slice D into two submatrices and recurse.
+      transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
+      transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
+   } else {
+      // Vertically slice D into two submatrices and recurse.
+      transpose(I, (int[2]){J[0], J[0] + num_cols / 2 - 1}, D, E);
+      transpose(I, (int[2]){J[0] + num_cols / 2, J[1]}, D, E);
+   }
+}
+#endif
+
+void print_matrix(int rows, int cols, int D[rows][cols]) {
+   if (rows > 42 || cols > 42) return;
+   for (int i = 0; i < rows; ++i) {
+      for (int j = 0; j < cols; ++j) {
+         printf("%d ", D[i][j]);
+      }
+      printf("\n");
+   }
+}
+
+int main() {
+   int(*D)[n] = (int(*)[n])calloc(n * m, sizeof(int));
+   srand(time(0));
+   for (size_t i = 0; i < n * m; ++i) {
+      ((int*)D)[i] = rand();
+   }
+   int(*E)[m] = (int(*)[m])calloc(n * m, sizeof(int));
+   // int D[m][n] = {{1, 2}, {3, 4}, {5, 6}};
+   // int E[n][m];
+   // print_matrix(m, n, D);
+   // printf("\n");
+   clock_t t0 = clock();
+#ifdef SIMPLE
+   transpose(D, E);
+#else
[…]

### 2017-06-25 tex/com.tex
Expand, tweak, add some code but don't use it
I expected this code to perform better...

@@
 and use \(D^\mathsf{T} = \begin{bmatrix} D_1^\mathsf{T} & D_2^\mathsf{T} \end{bmatrix}\)
-to compute the transpose.  If \(m < n\), analogously split \(D\) vertically.  Eventually,
-the submatrices will become small enough so that pairs of input and output submatrices fit
-into cache at the same time, at which point it doesn't matter in what order we access the
-elements.\footnote{%
+to compute the transpose.  If \(m < n\), analogously \alts{slice, split} \(D\) vertically.
+Eventually, the submatrices will become small enough so that pairs of input and output
+submatrices fit into cache at the same time, at which point it doesn't matter in what
+order we access the elements.\footnote{%
    No cache line will have to be evicted once loaded.
@@
 % \textcite[theorem 2, \ppno~19\psq]{coa-thesis}.
+\Cref{lst:xpose} shows my implementation in C.
 
-In practice, the new algorithm performs as badly as one might expect from this excessive
-use of recursion.  \Cref{fig:xpose-speedup} gives the speedup compared to the
-straightforward implementation, which is usually faster.
-
-% The code may only be 72 characters wide (75 when counting the indentation that will be
-% stripped).  Note that the array arguments will decay to pointers.
+% FIXME: get rid of this `center` hack and get the vertical spacing around listings with
+% captions to work in a better way.
+\begin{center}
+   % The code may only be 72 characters wide (78 when counting the indentation that will
+   % be stripped).  Note that the array arguments will decay to pointers.
    \begin{minted}[autogobble, mathescape]{c}
@@
    % // Transpose the submatrix $D\left[...\right]$
+   \captionof{listing}{\acrshort{com}-Optimal Matrix Transposition}
+   \label{lst:xpose}
+\end{center}
 
-\begin{figure}
+% In practice, the new algorithm performs as badly as one might expect from this excessive
+% use of recursion.
+In practice, the new algorithm performs worse than the straightforward one for a lot of
+matrix sizes before it eventually pulls ahead.
+\alts{
+   {\Cref{fig:xpose-speedup} gives the speedups for square matrices of various sizes},
+   {\Cref{fig:xpose-speedup} gives the speedup compared to the straightforward
+   implementation\x{, which is usually faster}},
+}.
+The problem may be the excessive use of recursion.
+
+Kumar works around this by \say{[stopping] the recursion when the problem size becomes
+less than a certain block size and then [using] the simple for loop implementation inside
+the block}~\cite[199--201]{afmh}.  This seems to \alts{compromise, jeopardize} the
+\alts{cache-obliviousness, cache-oblivious nature} of the algorithm.
+
+Perhaps cache-oblivious algorithms should be seen as a starting point for further
+optimizations that \alts{explicitly use the available information about the memory
+hierarchy of the machine a program runs on, take the properties of the target machine into
+account}.
+
+\x{
+Another approach may be to
+\alts{
+   {slice the matrix so that the split dimension becomes an even power of \(2\) for one of
+   the submatrices},
+   {\alts{split, divide} the matrix so that one of the resulting submatrices has a size
+   that is an even power of \(2\) along the dimension we split}
+}.
+\alts{
+   {Once a square submatrix whose dimensions are an even power of \(2\) is obtained, the
+   identical order of traversal that would result from the recursive algorithm can be
+   reproduced with a single loop.},
+   {Once a submatrix whose dimensions both are an even power of \(2\) is obtained, the
+   identical order of traversal that would result from the recursive algorithm can be
+   reproduced with loops relatively easily.},
+}
+}
+
+\begin{figure}[hbp]
    \centering
@@
                label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
-                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256},
+                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024},
                },
@@
    \end{tikzpicture}
-   \caption{TODO}
+   \captionsetup{width=.9\linewidth} % FIXME: hacks!
+   \caption[Speedup Achieved by \acrshort{com}-Optimal Matrix Transposition]{Speedup
+   Achieved by \acrshort{com}-Optimal Matrix Transposition; the CPU time used by the
+   straightforward matrix transposition algorithm divided by that used by the optimal one}
    \label{fig:xpose-speedup}

### 2017-06-25 tex/list-vs-vector.tex
Create more slides
Minor tweaks and additions
Improve spacing between listings and captions
Expand, tweak, add some code but don't use it
I expected this code to perform better...

@@
 \begin{center}
-   \inputminted[firstline=29]{cpp}{ithare/list-vs-vector.cpp}
-   \captionof{listing}{Compute the Sum of a Container, adapted from \textcite{bigos}}
+   \inputminted[firstline=29]{cpp}{ithare/list-vs-vector.cpp}%
+   \captionof{listing}[Compute the Sum of a Container]{Compute the Sum of a Container;
+   adapted from \textcite{bigos}}
    \label{lst:ithare}
@@
    \centering
-   \input{tex/oo-picture}
+   \input{tex/graphics/oo-picture}
    \captionsetup{width=.9\linewidth} % FIXME: hacks!
@@
 
-\subsubsection{Unrolled List}
+% \subsubsection{Unrolled List}
 

### 2017-06-25 xpose/xpose.c

@@ -11,6 +11,86 @@ void transpose(int D[m][n], int E[n][m]) {
    for (int i = 0; i < m; ++i)
       for (int j = 0; j < n; ++j) E[j][i] = D[i][j];
 }
+#elif defined COMPLICATED
+int isolate_rightmost_bit(int i) { return i & (-i); }
+
+int is_power_of_two(int i) {
+   // See <https://stackoverflow.com/q/108318>.
+   return (i == 1) | ((i & (i - 1)) == 0);
+}
+
+// From <http://asgerhoedt.dk/?p=276>.
+unsigned int compact(int x) {
+    x &= 0x55555555;                 // x = -f-e -d-c -b-a -9-8 -7-6 -5-4 -3-2 -1-0
+    x = (x ^ (x >> 1)) & 0x33333333; // x = --fe --dc --ba --98 --76 --54 --32 --10
+    x = (x ^ (x >> 2)) & 0x0f0f0f0f; // x = ---- fedc ---- ba98 ---- 7654 ---- 3210
+    x = (x ^ (x >> 4)) & 0x00ff00ff; // x = ---- ---- fedc ba98 ---- ---- 7654 3210
+    x = (x ^ (x >> 8)) & 0x0000ffff; // x = ---- ---- ---- ---- fedc ba98 7654 3210
+    return x;
+}
+
+// From <http://asgerhoedt.dk/?p=276>.
+void morton_decode(int c, int* x, int* y) {
+   *x = compact(c);
+   *y = compact(c >> 1);
+}
+
+// Transpose a square matrix with dimensions that are even powers of two
+// accessing the elements in Z-order.
+void z_transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
+   // printf("%i %i\n", 1 + I[1] - I[0], 1 + J[1] - J[0]);
+   /*
+   for (int i = I[0]; i <= I[1]; ++i)
+      for (int j = J[0]; j <= J[1]; ++j) E[j][i] = D[i][j];
+   */
+   int num_elements = (1 + I[1] - I[0]) * (1 + J[1] - J[0]);
+   for (int z = 0; z < num_elements; ++z) {
+      int i, j;
+      morton_decode(z, &i, &j);
+      i += I[0];
+      j += J[0];
+      // printf("z: %i, (i,j): (%i, %i)\n", z, i, j);
+      E[i][j] = D[j][i];
+   }
+}
+
[…]

### 2017-06-25 paper.tex
Tweak
Expand content, revise, tweak
Add TikZ graphic representing an array of pointers
Tweak float placement
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/
Minor tweaks and additions
Tweak stuff and fill in some blanks

@@
 
+% Required for `\mathbb{}`.  TODO: `amssymb` loads `amsfonts`; which one should I use?
+\usepackage{amsfonts}
+
+% Used to put TikZ graphics into their own dedicated files and compile them separately.
+% From the manual: "Load the standalonepackage very early in the main document."
+\usepackage{standalone}
+
 % `$\nicefrac{a}{b}$` looks better than `$a/b$` in my opinion.  I also think it looks
@@
 
-% Prevent floats from moving into another section (https://ctan.org/pkg/placeins).
-\usepackage[above, below, section]{placeins}
+% XXX: I used to specify the `section` option here to prevent floats from moving into
+% another section.  It redefines the `\section` command to add a `\FloatBarrier`.  Turns
+% out this doesn't work the way I expected.  Apparently, `\FloatBarrier` just issues
+% `\clearpage` if there are floats queued [1][2].  The result can be mostly empty pages.
+%
+% [1]: https://tex.stackexchange.com/q/223917#comment525956_223917
+% [2]: https://tex.stackexchange.com/q/272858#comment655658_272858
+% [3]: https://tex.stackexchange.com/q/88657
+% [4]: https://ctan.org/pkg/placeins)
+\usepackage[above, below]{placeins}
+
+% TODO: `\usepackage{afterpage}`?
+%
+% [1]: https://tex.stackexchange.com/q/88657
+% [2]: https://www.ctan.org/pkg/afterpage
 
@@
 
-% Stack for generating tables from CSV files.  Loading is done by [pgfplotstable].  It can
-% also round, format and post-process data.  See [1].
-% [siunitx] adds the `S` column type for advanced aligning in tables like around the
-% decimal marker.  Getting this to work is a bit tricky; see [2].  There is some support
-% for aligning around the decimal marker in [pgfplotstable] (see [3]), but I couldn't get
-% it to work.  One can also right-align numbers relatively to each other but still center
-% the column as a whole with [siunitx] (see [4]).
-% [pgfplotstable]: https://ctan.org/pkg/pgfplotstable
-% [siunitx]: https://ctan.org/pkg/siunitx
-% [1]: https://tex.stackexchange.com/q/146716
-% [2]: https://tex.stackexchange.com/q/131081
-% [3]: https://tex.stackexchange.com/q/276485
-% [4]: https://tex.stackexchange.com/q/9203
-\usepackage{pgfplotstable}
-\usepackage[binary-units]{siunitx} % Load `\kibi` and `\mebi` etc. (`binary-units`).
-% These packages are recommended by [pgfplotstable].  The booktabs package adds the
-% `\toprule`, `\midrule`, and `\bottomrule` commands for better looking tables.  See [5].
-\usepackage{booktabs, array, colortbl}
-% [booktabs]: https://ctan.org/pkg/booktabs
-% [5]: https://en.wikibooks.org/wiki/LaTeX/Tables#Professional_tables
-
-% This gets loaded by the `pgfplotstable` sub-package as far as I know.
-%
-% \usepackage{pgfplots}
-%
-% It then suggests:
-%
-%    Package pgfplots Warning: running in backwards compatibility mode (unsuitable t
-%    ick labels; missing features). Consider writing \pgfplotsset{compat=1.14} into
-%    your preamble
-%
-% So I guess I'll go that.
-\pgfplotsset{compat=1.14}
+\input{tex/pgfplotstablesetup}
 
@@
 \usepackage{tikz}
-\input{tex/datavis.tex}
+\usetikzlibrary{math}
+\input{tex/datavis}
 
@@
 
+% Macro for typesetting "C++".  Just writing it verbatim looks very ugly.  See
+% <https://stackoverflow.com/q/2724760>.
+\usepackage{relsize} % For `\smaller`.
+\def\ifmonospace{\ifdim\fontdimen3\font=0pt}
+\def\cpp{%
+\ifmonospace%
+    C++%
+\else%
+    C\kern-.0667em\raise.30ex\hbox{\smaller{++}}%
+\fi%
+\spacefactor1000}
+% Alternative from <https://tex.stackexchange.com/q/4302>:
+% \def\cpp{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bfseries++}}}
+
 % Top matter {{{1
-\title{Basics of Hardware Cache Optimization}
+\title{Hardware Caches and Optimization}
+% \title{Optimizing for Hardware Caches}
+% \title{Basics of Hardware Cache Optimization}
 % \title{Algorithms for Hardware Caches}
@@
 % \title{Introduction to Optimizations for Hardware Caches}
-% \title{Optimizing for Hardware Caches}
 % \subtitle{The Basics}
@@
 
-   The abstract \gls{emm} for memory hierarchies is explained and a small selection of
-   algorithms \alts{developed for it, analyzed under it} are explored mathematically and
-   empirically.
+   The abstract \gls{emm} for memory hierarchies and the \gls{com} derived from it are
+   presented briefly.
 \end{abstract}
@@
 \input{tex/motivation}
-% \section{Hardware Caches} % Cache, Structure of Caches
+% TODO: "Hardware Caches", "Cache", or "Structure of Caches" section?
 \input{tex/operation}
@@
 \input{tex/key-concepts}
-\input{tex/list-vs-vector.tex}
-\label{sec:lvv}
-% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
-% https://en.wikipedia.org/wiki/Cache_replacement_policies
-\input{tex/emm.tex}
-\section{Parallel Computing}
-\subsection{Cache Coherence} % Coherence or coherency?
-% https://en.wikipedia.org/wiki/Cache_coherence
-% Dirty.  Invalid.
-\subsection{False Sharing}
+\input{tex/list-vs-vector}
+% TODO: "Cache Replacement Policies", "Eviction (Policies|Strategies|Algorithms)" section?
+% See <https://en.wikipedia.org/wiki/Cache_replacement_policies>.
+\input{tex/abstract-}
+% \input{tex/parallel-computing}
+% \input{tex/false-sharing}
+% TODO: "Roofline" section?  See <https://en.wikipedia.org/wiki/Roofline_model>.
 

### 2017-06-25 tex/types.tex
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/
Add more slides, tweak stuff
Improve spacing between listings and captions
Tweak stuff and fill in some blanks

@@
    \item Often a unified \gls{l3} cache of 2 to 16 MiB shared between all cores.
-   \item TODO: Some \glspl{tlb} I guess.
+   \item One or more \glspl{tlb} per core.  \alts{They cache virtual-to-physical address
+      associations of memory pages, They cache associations of virtual to physical
+      addresses belonging to memory pages}.
 \end{itemize}
@@
    \begin{figure}
-      \inputminted[firstline=9]{c}{access-times/access-times.c}
+      \inputminted[firstline=9]{c}{access-times/access-times.c}%
       \captionof{listing}{TODO}
@@
 % \begin{code}
-%    \inputminted[firstline=9]{c}{access-times/access-times.c}
+%    \inputminted[firstline=9]{c}{access-times/access-times.c}%
 %    \caption{TODO}
@@
 \begin{center} % XXX: hack to get normal spacing after the caption and before the listing.
-   \inputminted[firstline=12]{c}{access-times/access-times.c}
+   \inputminted[firstline=12]{c}{access-times/access-times.c}%
    % \captionof{listing}{}
@@
    \centering
-   \tikz \datavisualization[array size vs cycles plot]
-      data [read from file=access-times/access-times.csv, separator=\space];
-   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
+   \input{tex/graphics/access-time-plot}
    \caption{Access Times for Random Reads}

### 2017-06-27 tex/com.tex
Tweak stuff and fill in some blanks
Expand, revise, tweak, and run some measurements

@@
 }
+\alts{
+   cache misses only increase by a constant factor when switching to a feasible
+   replacement strategy,
    it only increases cache misses by a constant factor compared to various feasible
-replacement strategies%
+   replacement strategies,
+}%
 % ~\cite[lemma 12, \pno~10]{coa-paper}.%
@@
 \alts{cache-obliviousness, cache-oblivious nature} of the algorithm.
-
 Perhaps cache-oblivious algorithms should be seen as a starting point for further
@@
    \centering
-   \begin{tikzpicture}
-      \datavisualization
-      [scientific axes=clean,
-       x axis={logarithmic,
-               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
-                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024},
-               },
-              },
-       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
-         % ticks={major also at={1}}
-       },
-       visualize as scatter,
-       scatter={style={mark=*, mark options={scale=.65}}}]
-         data [read from file=xpose/speedup.csv, separator=\space];
-   \end{tikzpicture}
+   \input{tex/graphics/xpose-speedup-plot}
    \captionsetup{width=.9\linewidth} % FIXME: hacks!
@@
 \end{figure}
+
+% \begin{figure}[hbp]
+%    \centering
+%    \input{tex/graphics/xpose-speedup-complicated-plot}
+%    \captionsetup{width=.9\linewidth} % FIXME: hacks!
+%    \caption{TODO}
+%    \label{fig:xpose-speedup-complicated}
+% \end{figure}
 

### 2017-06-27 tex/graphics/xpose-speedup-complicated-plot.tex
Expand, revise, tweak, and run some measurements

\documentclass[tikz, border=1pt]{standalone}

\input{tex/datavis}

% FIXME: DRY.
\usepackage[binary-units]{siunitx} % Load `\kibi` and `\mebi` etc. (`binary-units`).

% FIXME: DRY.
\pgfkeys{%
   /pgf/number format/.cd,
   1000 sep={\,},
   min exponent for 1000 sep=4,
}

\begin{document}
% FIXME: this is the same setup used by `xpose-speedup-plot.tex`... DRY
\tikz \datavisualization[scientific axes=clean, x axis={
   logarithmic, label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
         major at={1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024},
      },
   }, y axis={
      include value=0, grid=at ticks, label={Speedup}, length=6cm,
   % ticks={major also at={1}}
   },
   visualize as scatter,
   scatter={style={mark=*, mark options={scale=.65}}}]
   data [read from file=xpose/speedup-complicated.csv, separator=\space];
\end{document}

% vim: ft=tex tw=90 sts=-1 sw=3 et fdm=marker

### 2017-06-27 tex/methodology.tex
Expand, revise, tweak, and run some measurements

@@
 
-All programs were compiled and run on Linux using the \gls{gcc} with
-\mintinline{text}{-O2} (unless noted otherwise) and \mintinline{text}{-march=native}.
+All programs were compiled and run on Linux using the \gls{gcc}\footnote{%
+   in version 6.3.1
+}
+with \mintinline{text}{-O2} or \mintinline{text}{-O3} and
+\mintinline{text}{-march=native}.
 % Additionally,

### 2017-06-27 unrolled-list/bits/stl_list.h
Add code that didn't end up being used

@@ -1,0 +1,1964 @@
+// List implementation -*- C++ -*-
+
+// Copyright (C) 2001-2016 Free Software Foundation, Inc.
+//
+// This file is part of the GNU ISO C++ Library.  This library is free
+// software; you can redistribute it and/or modify it under the
+// terms of the GNU General Public License as published by the
+// Free Software Foundation; either version 3, or (at your option)
+// any later version.
+
+// This library is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+
+// Under Section 7 of GPL version 3, you are granted additional
+// permissions described in the GCC Runtime Library Exception, version
+// 3.1, as published by the Free Software Foundation.
+
+// You should have received a copy of the GNU General Public License and
+// a copy of the GCC Runtime Library Exception along with this program;
+// see the files COPYING3 and COPYING.RUNTIME respectively.  If not, see
+// <http://www.gnu.org/licenses/>.
+
+/*
+ *
+ * Copyright (c) 1994
+ * Hewlett-Packard Company
+ *
+ * Permission to use, copy, modify, distribute and sell this software
+ * and its documentation for any purpose is hereby granted without fee,
+ * provided that the above copyright notice appear in all copies and
+ * that both that copyright notice and this permission notice appear
+ * in supporting documentation.  Hewlett-Packard Company makes no
+ * representations about the suitability of this software for any
+ * purpose.  It is provided "as is" without express or implied warranty.
+ *
+ *
+ * Copyright (c) 1996,1997
+ * Silicon Graphics Computer Systems, Inc.
+ *
+ * Permission to use, copy, modify, distribute and sell this software
+ * and its documentation for any purpose is hereby granted without fee,
+ * provided that the above copyright notice appear in all copies and
+ * that both that copyright notice and this permission notice appear
[…]

### 2017-06-27 unrolled-list/unrolled_list.cpp

@@ -1,0 +1,112 @@
+// vim: tw=72 sts=-1 sw=3 et
+
+#include <cassert>
+#include <cstdlib>  // std::rand, std::srand
+#include <ctime>    // std::clock, std::time, std::clock_t
+#include <iostream>
+
+#include <xmmintrin.h>
+
+#ifdef UNROLL
+#include <array>
+#include "list"  // Include the copy in this directory.
+#else
+// #include <list>
+#include <vector>
+#endif
+
+constexpr int NODE_SIZE = 12;
+// constexpr int NODE_SIZE = 28;
+// constexpr int NODE_SIZE = 44;
+// constexpr int NODE_SIZE = 60;
+
+#ifdef UNROLL
+// Extremely ad-hoc.
+class unrolled_list : public std::list<std::array<int, NODE_SIZE>> {
+  public:
+   void push_back(const int& x) {
+      // _Node is a typedef for the actual node type GCC uses.
+      // assert(sizeof(_Node) == 64);
+      // std::cout << alignof(_Node) << '\n';
+      if (!this->empty()) {
+         // std::cout << long(this->end()._M_node) % 64 << '\n';
+         auto& last = this->back();
+         int& num_elements = last[0];
+         assert(num_elements >= 1);
+         // std::cout << last[0] << std::endl;
+         if (num_elements < NODE_SIZE - 1) {
+            // if (num_elements < 2) {
+            ++num_elements;
+            last[num_elements] = x;
+            return;
+         }
+      }
+      value_type new_node;
+      new_node[0] = 1;  // One element.
+      new_node[1] = x;
+      std::list<value_type>::push_back(new_node);
+   }
+};
+using Container = unrolled_list;
+#else
+// using Container = std::list<int>;
+using Container = std::vector<int>;
+#endif
+
+constexpr int N = 5000;
+
+int main() {
+   Container containers[N];
+   std::srand(std::time(nullptr));
+   // Append an average of 5000 random values to each container.
+   for (int i = 0; i < N * 5000; ++i) {
+      containers[std::rand() % N].push_back(std::rand());
+   }
+   // for (int i = 0; i < N; ++i)
+   //    for (int j = 0; j < 5000; ++j)
+   //       containers[i].push_back(std::rand());
+
+   int sum = 0;
+   std::clock_t t0 = std::clock();
+   for (int m = 0; m < N; ++m) {
+#ifdef UNROLL
[…]
